---
title: "Standard Model Diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{standard-model-diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will demonstrate `pmxhelpr` functions for generating standard model diagnostics. 

First, we will load the required packages.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(Hmisc, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
```

For this vignette, we will generate model diagnostics for a PK model (`model`) fit to `data_sad` dataset using the `data_sad_pkfit` dataset internal to `pmxhelpr`. We can take a quick look at the dataset using `glimpse()` from the dplyr package. Dataset definitions can also be viewed by calling `?data_sad_pkfit`, as one would to view the documentation for a package function.

```{r data-glimpse}
glimpse(data_sad_pkfit)
```

This dataset contains two additional variables relative to the observed dataset used in model development (`data_sad`):

+ `PRED`: population model predicted values accounting only for fixed effects (THETAs)
+ `IPRED`: individual model predicted values accounting for fixed effects (THETAs) and level 1 (inter-individual) random effects (ETAs)

Let's add some additional derived character variables, which may be useful for plotting. 

```{r data-plotdata}
plot_data <- data_sad_pkfit %>% 
  mutate(FoodStatus = ifelse(FOOD == 0, "Fasted", "Fed"), 
         DoseGroup = paste0(DOSE, " mg ", FoodStatus), 
         DoseGroup = factor(DoseGroup, levels = c("10 mg Fasted", "50 mg Fasted", "100 mg Fasted", "100 mg Fed", 
                                                  "200 mg Fasted", "400 mg Fasted")))
```


# Population Overlay Goodness of Fit Plots

## Overview of `plot_popgof`

 `pmxhelpr` includes a function for generating population overlay goodness-of-fit (GOF) plots for model evaluation: `plot_popgof`.
 
 *NOTE* `plot_popgof` inherits dataset variable handling functionality and arguments from `df_mrgsim_replicate`.
  
Variable processing requires specification of two named lists of time and output variables: 

-  `time_vars`: This argument provides a named vector of actual and nominal time variables. `NTIME` must be an exact binning variable.
    + Default is: c(`TIME`=`"TIME"`, `NTIME`=`"NTIME"`).
    
- `output_vars` This argument provides a named vector of input (observations) and output (model prediction) variables. 
    + Default is: c(`PRED`= `"PRED"`, `IPRED` = `"IPRED"`, `DV`= `"DV"`).
  

The example dataset `data_sad_pkfit` only differs from these defaults in the variable nname for the dependenrt variable, `"ODV"`. Thus, the most basic population GOF plot can be obtained with: 

```{r plot-popgof, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_popgof(data = plot_data, output_vars = c(DV ="ODV")) +
  facet_wrap(~PART)
```

The pooled analysis population includes multiple dose levels in a single plot, which is probably not optimal as a population PK model diagnostic. A better minimal plot representation of these data can be obtained by dose-normalizing and stratifying by study part to separate out the fast and food effect portions:

```{r plot-popgof-dn, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_popgof(data = plot_data, output_vars = c(DV ="ODV"), dosenorm = TRUE, 
            ylab = "Dose-normalized Conc. (ng/mL)") +
  facet_wrap(~PART)
```

Although, generally population goodness-of-fit plots are stratified by study design (extrinsic) factors (e.g., Dose and Food Status) in order to assess the adequacy of model fit in each unique study condition.

```{r plot-popgof-dosegrp, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_popgof(data = plot_data, output_vars = c(DV ="ODV"), 
            ylab = "Concentration (ng/mL)") +
  facet_wrap(~DoseGroup, scales = "free")
```

*NOTE* `plot_popgof` plot aesthetic functionality from `plot_dvtime`.

This vignette will assume familiarity with the Exploratory Data Analysis Vignette and `plot_dvtime`. The arguments and functionality inherited from `plot_dvtime` will not be reviewed in detail here, except where critical to interpretation of model diagnostics.

## Plot Elements and Colors

There is only one new argument to `plot_popgof` that is not present in `df_mrgsim_replicate` or `plot_dvtime`. This argument is `output_colors`, a named vector of pre-defined plot element and color pairs:

  - `PRED= "red"`
  - `IPRED`=`"green"`,
  - `DV`=`"blue"`.
  - `OBS`=`"darkgrey"`

The default plot elements are:

 - `PRED` - summary statistics for central tendency in the population model predictions (`output_vars["PRED"]`)
 - `IPRED` = summary statistics for central tendency in the individual model predictions (`output_vars["IPRED"]`)
 - `OBS` = observed data points for the dependent variable (`output_vars["DV"]`)
 - `DV` = summary statistics for central tendency (+/- variability) in the dependent variable (`output_vars["DV"]`)
 
`OBS` and `DV` represent the same data and both map back to the variable in `data` specified in `output_vars["DV"]`. They are differentiated only to allow visual distinction between individual observation concentration data and summary statistics of observed and model-predicted concentration data.

If one wanted to clarify that all the data mapping back to the dataset variable specified in `output_vars["DV]` with color, then the default for `output_colors["OBS"]` can be updated to to blue to align with `DV `.

```{r plot-popgof-dosegrp-blue, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_popgof(data = plot_data, output_vars = c(DV ="ODV"), 
            ylab = "Concentration (ng/mL)", output_colors = c(OBS = "blue")) +
  facet_wrap(~DoseGroup, scales = "free")
```

Alternatively, the observed data points could be removed with only the summary metrics plotted. Notice how different this plot appears when only visualizing central tendency!

```{r plot-popgof-dosegrp-noobs, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_popgof(data = plot_data, output_vars = c(DV ="ODV"), 
            ylab = "Concentration (ng/mL)", obs_dv = FALSE) +
  facet_wrap(~DoseGroup, scales = "free")
```

A better plot that still emphasizes central tendency, but includes some visualization of variability in the observed data, could also be obtained by specifying one of "".

```{r plot-popgof-dosegrp-noobs-mediqr, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_popgof(data = plot_data, output_vars = c(DV ="ODV"), 
            ylab = "Concentration (ng/mL)", obs_dv = FALSE, cent = "median_iqr") +
  facet_wrap(~DoseGroup, scales = "free")
```

## Specifying the Central Tendency

`plot_popgof` (and `plot_dvtime`) uses the `stat_summary` function from `ggplot2` to calculate and plot the central tendency measures and error bars. The summary statistics calculated are specified by the `cent` argument. Options include:

  + `"mean"` = arithmetic mean (linear y-axis), geometric mean (log y-axis)
  + `"median"` = median (50th percentile)
  + `"mean_sdl"` = arithmetic mean +/- arithmetic SD (linear y-axis), geometric mean +/- geometric SC (log y-axis)
  + `"median_iqr"` = median +/- interquartile range (25th to 75th percentiles)
  
Variability statistics (SD, IQR) are plotted for the observed data when requested and only the central tendency measures are calculated and returned for model predictions (PRED, IPRED).

An often overlooked feature of `stat_summary`, is that it calculates the summary statistics *after* any transformations to the data performed by changing the scales. This means that when `scale_y_log10()` is applied to the plot, the data are log-transformed for plotting and the central tendency measure returned with `"mean"` is the *geometric mean*. 

If the `log_y` argument is used to generate semi-log plots along with `show_captions = TRUE`, then the caption will delineate where arithmetic and geometric means are being returned. 

On the linear scale, the `"mean_sdl"` option produces unrealistic error bars that cross zero given the small sample size, variability, and log-normal distribution of concentration-time data. 

```{r plot-popgof-dosegrp-meansd, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_popgof(data = plot_data, output_vars = c(DV ="ODV"), cent = "mean_sdl" ,
            ylab = "Concentration (ng/mL)", log_y = FALSE) +
  facet_wrap(~DoseGroup, scales = "free")
```

However, when transformed to a semi-log scale with `log_y = TRUE`, the error bars now reflect the geometric standard deviation and are displayed correctly for lognormally distributed data and the caption updates to reflect this.

```{r plot-popgof-dosegrp-meansd-logy, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_popgof(data = plot_data, output_vars = c(DV ="ODV"), cent = "mean_sdl", 
            ylab = "Concentration (ng/mL)", log_y = TRUE) +
  facet_wrap(~DoseGroup, scales = "free")
```

The caption will *NOT* update if a new axis is added to the plot object outside of `plot_popgof` with `scale_y_log10`.

```{r plot-popgof-dosegrp-logscale, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_popgof(data = plot_data, output_vars = c(DV ="ODV"), cent = "mean_sdl",
            ylab = "Concentration (ng/mL)") +
  facet_wrap(~DoseGroup, scales = "free") + 
  scale_y_log10()
```

Transformation of the plot to a semi-log scale (log-scale y-axis only) is recommended to be performed using the `log_y` argument for the following benefits:

  + Includes log tick marks on the y-axis 
  + Updates the caption with the correct central tendency measure if `show_captions = TRUE`. 
  

## Defining imputations for BLQ data

There are some potential missfit in the model in the late terminal phase, with the observed geometric mean values skewing higher than the corresponding model predictions, particularly at lower doses. Is this real? Let's use imputation to assess the potential impact of the data missing due to assay sensitivity on assessment of model misspecification. The `loq_method` argument from `plot_dvtime` was extended in `plot_popgof` to also include model predictions. 

Options are:

  + `0` : No handling. Plot input dataset `DV` vs `TIME` as is. (default)
  + `1` : Impute all BLQ data at `TIME` <= 0 to 0 and all BLQ data at `TIME` > 0 to 1/2 x `loq`.
      Useful for plotting concentration-time data with some data BLQ on the linear scale
  + `2` : Impute all BLQ data at `TIME` <= 0 to 1/2 x `loq` and all BLQ data at `TIME` > 0 to 1/2 x `loq`. 
  
The `loq` argument species the value of the LLOQ. The `loq` argument must be specified when `loq_method` is `1` or `2`, but can be `NULL` *if* the variable `LLOQ` is present in the dataset. 

Let's take a look at our data stratified by dosing regimen. In general, it is good practice to use `loq_method = 1` for linear scale plots and `loq_method = 2` for log-scale plots, in which zero is not a valid value. 

```{r plot-popgof-dosegrp-logy-loq2, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_popgof(data = plot_data, output_vars = c(DV ="ODV"), cent = "mean", grp_dv = TRUE,
            ylab = "Concentration (ng/mL)", log_y = TRUE, loq_method = 2) +
  facet_wrap(~DoseGroup, scales = "free")
```

Interesting! It looks like the potential misspecification in the late terminal phase was in fact an artifact in the observed profile due to censoring of data below the LLOQ. This is a common finding in SAD studies and is usually dose dependent and more apparent at lower doses where a larger proportion of samples will be BLQ.


# Individual Concentration-time plots

The previous section provides an overview of how to generate population overlay GOF plots using `plot_popgof`; however, we can also use `plot_popgof` to generate subject-level visualizations with a little pre-processing of the input dataset. 

We can plot an individual subject by filtering the input dataset. This could be extended generate plots for all individuals using `for` loops, `lapply`,  `purrr::map()` functions, or other methods.

```{r plot-indgofplots, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}

ids <- sort(unique(plot_data$ID)) #vector of unique subject ids
n_ids <- length(ids) #count of unique subject ids
plots_per_pg <- 4
n_pgs <- ceiling(n_ids/plots_per_pg) #Total number of pages needed

plist<- list()
for(i in 1:n_ids){
  plist[[i]] <- plot_popgof(filter(plot_data, ID == ids[i]), 
                            output_vars = c(DV = "ODV"), 
                            ylab = "Concentration (ng/mL)", 
                            log_y = TRUE,
                            grp_dv = TRUE,
                            show_caption = FALSE) +
  facet_wrap(~DoseGroup)+
  labs(title = paste0("ID = ", ids[i]))+
  theme(legend.position="none")
}

lapply(1:n_pgs, function(n_pg) {
      i <-  (n_pg-1)*plots_per_pg+1
      j <- n_pg*plots_per_pg
      wrap_plots(plist[i:j])
})
```

