---
title: "Exploratory Data Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eda-pmxhelpr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will demonstrate `pmxhelpr` functions for exploratory data analysis. 

First oad the required packages.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(forcats, warn.conflicts = FALSE)
library(Hmisc, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
library(PKNCA, warn.conflicts = FALSE)
```

# Data

The datasets used in this vignette are based on a simple ascending dose (SAD) study of an orally drug product with a parallel group food effect (FE) cohort.

## data_sad

Dataset definitions can be viewed by calling `?data_sad`. We can take a quick look at the primary dataset for this analysis (`data_ssad`)  using `glimpse()` from the `dplyr` package.

```{r data-sad-glimpse}
glimpse(data_sad)
```

The study design consisted of two parts. Part 1 was a SAD study over a 10 to 400 mg dose range and Part 2 was a parallel 100 mg single dose food effect (FE) study. 

```{r data-sad-part}
distinct(data_sad, DOSE, PART, FOOD)
```

The dataset is formatted for population pharmacokinetic (PopPK) modeling in NONMEM with oral dose events (`EVID=1`) input into `CMT=1` and drug concentration observations (`EVID=0`) in`CMT=2`. Dose events are input in `AMT` with the nominal dose associated with each observation captured in `DOSE`.

Plasma drug concentration observations are expressed in multiple units:  

  - `ODV`: original units of the dependent variable [ng/mL[]
  - `LDV`: natural logarithm-transformed units of drug concentration [log(ng/mL)]

The dataset also contains multiple variables for time:

  - `TIME`: Actual time since first dose administration [hours]
  - `NTIME`: Nominal time dose event or sample collection per protocol [hours]
  - `NDAY`: Nominal day on study [day]

The actual time since first dose administration is assigned to the NONMEM reserved variable for time (`TIME`), as this represents the most accurate description of the time order of the repeated-measures data. The nominal time varibles are *exact binning variables*, which are useful for grouping data for exploratory data analysis or model evaluation....but more on that later.

```{r data-sad-ntime}
##Unique values of time variables
times <- unique(data_sad$TIME)
ntimes <- unique(data_sad$NTIME)
ntimes

##Comparison of number of unique values of NTIME and TIME
length(ntimes)
length(times)
```

## data_sad_pd

The companion dataset, `data_sad_pd`, is `data_sad` with an additional pharmacodynamic (PD) response biomarker observation type (`EVID=0`) in an additional compartment`CMT=3`. These response data are expressed as percentage of baseline activity (%) in both `ODV` and `LDV` and as percentage change from baseline activity in `CFB`.

```{r data-sad-pd-glimpse}
glimpse(data_sad_pd)
```

## df_addn

One should make sure to only pass the data of interest for visualization into any plotting function. Let's filter to only observations, and define some grouping variables with summative information for subsequent exploratory analysis.

The helper function `df_addn` takes an input dataset `data` and returns a dataset with the variable specified in `grp_var` transformed to a factor including the count of unique values of the identifier variable specified in `id_var` (default `"ID"`). A string constant separator can be added between the values in `grp_var` and the count of `id_var` using `sep`. A common use is to specify the dose units when linking dose and count of unique individuals receiving that dose.

```{r plot-data, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_data_pk <- data_sad %>% 
  filter(EVID == 0) %>% 
  mutate(`Food Status` = ifelse(FOOD == 0, "Fasted", "Fed"), 
         `Dose and Food` = paste(DOSE, "mg", `Food Status`),
         Dose = DOSE) %>% 
  df_addn(grp_var = "Dose", sep = "mg") %>% 
  df_addn(grp_var = "Dose and Food")

plot_data_pd <- data_sad_pd %>% 
  filter(EVID == 0) %>% 
  mutate(`Food Status` = ifelse(FOOD == 0, "Fasted", "Fed"), 
         `Dose and Food` = paste(DOSE, "mg", `Food Status`),
         Dose = DOSE) %>% 
  df_addn(grp_var = "Dose", sep = "mg") %>% 
  df_addn(grp_var = "Dose and Food")

unique(plot_data_pk$Dose)
unique(plot_data_pk$`Dose and Food`)
```
Unfortunately, often our factors will not be ordered in ascending numerical order due ordering of strings. In this case, the 50 mg dose group is sorted at the end. We can use the `forcats` package in the `tidyverse` to quickly reorder our new variables

```{r plot-data-reorder, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_data_pk <- plot_data_pk %>% 
  mutate(Dose = fct_relevel(Dose, "50 mg (n=6)", after = 1), 
         `Dose and Food` = fct_relevel(`Dose and Food`, "50 mg Fasted (n=6)", after = 1))

plot_data_pd <- plot_data_pd %>% 
  mutate(Dose = fct_relevel(Dose, "50 mg (n=6)", after = 1), 
         `Dose and Food` = fct_relevel(`Dose and Food`, "50 mg Fasted (n=6)", after = 1))
  
unique(plot_data_pk$Dose)
unique(plot_data_pk$`Dose and Food`)
```

# Population Concentration-time plots

## Overview of `plot_dvtime`

Now let's visualize the concentration-time data! `pmxhelpr` includes a function for common visualizations of observed concentration-time data in exploratory data analysis: `plot_dvtime`

Let's run it!

```{r plot-dvtime-fail, fig.height=4, fig.width=6, error=TRUE, message=TRUE}
plot_dvtime(plot_data_pk)
```
Hmm...that didn't work. Hmm...well checking `?plot_dvtime()` it seems that the argument `dv_var` specifies the dependent variable in the dataset and the default is `"DV"`. Since `data_sad` has the original units of the dependent variable as `ODV`, this name must be passed to the function.

```{r plot-dvtime-pass, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(plot_data_pk, dv_var = "ODV")
```
Hey a plot! Conveniently, time variables in `data_sad` had names aligned with the default argument; however, this may not always be the case. A caption prints by default describing the plot elements. The caption can be removed by specifying `show_caption = FALSE`.

```{r plot-dvtime-nocap, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = "ODV", show_caption = FALSE) 
```

These data are probably best visualized on a log-scale y-axis upweight the terminal phase profile. `plot_dvtime` includes an argument `log_y` which performs this operation with some additional formatting benefits over manually adding the layer to the returned object with `scale_y_log10`.

  + Includes log tick marks on the y-axis 
  + Updates the caption with the correct central tendency measure if `show_captions = TRUE`. 

`plot_dvtime` uses the `stat_summary` function from `ggplot2` to calculate and plot the central tendency measures and error bars. An often overlooked feature of `stat_summary`, is that it calculates the summary statistics *after* any transformations to the data performed by changing the scales. This means that when `scale_y_log10()` is applied to the plot, the data are log-transformed for plotting and the central tendency measure returned when requesting `"mean"` from `stat_summary` is the *geometric mean*. If the `log_y` argument is used to generate semi-log plots along with `show_captions = TRUE`, then the caption will  delineate where arithmetic and gemoetric means are being returned. 

```{r plot-dvtime-logy, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = "ODV", log_y) 
```


## Adjusting Time Variables and Breaks

The names of the actual and nominal time variables can be passed as a named character vector with default `time_vars=c(TIME=TIME", NTIME="NTIME")`. These elements of the argument may be specified one at a time, as in the example below showing all nominal values, or both together.

```{r plot-dvtime-nom, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(plot_data_pk, dv_var = "ODV", time_vars = c(TIME = "NTIME"))
```
 
`plot_dvtime` includes uses a helper function (`breaks_time`) to automatically determine x-axis breaks based on the units of the time variable! Two arguments in `plot_dvtime` are passed to `breaks_time`:

  + `timeu` character string specifying time units. Options include:
  
     * "hours" (default), "hrs", "hr", "h"
     * "days", "dys", "dy", "d"
     * "weeks", "wks", "wk", "w"
     * "months", "mons", "mos", "mo", "m"
      
  + `n_breaks` number breaks requested from the algorithm. Default = 8.
  
Let's pass the vector of nominal times we defined earlier into the `breaks_time` function and see what we get with different requested numbers of breaks!

```{r}
breaks_time(ntimes, unit = "hours")
breaks_time(ntimes, unit = "hours", n = 5)
breaks_time(ntimes, unit = "hours", n = length(ntimes))
```

We can see that the default (n = 8) gives an optimal number of breaks in this case whereas reducing the number of breaks (n=5) gives a less optimal distribution of values. Requesting breaks equal to the length of the vector of unique `NTIMES` will generally produce too many breaks. The default axes breaks behavior can always be overwritten by specifying the axis breaks manually using `scale_x_continuous()`.

The default `n_breaks = 8` is a good value for `data_sad`, and `data_sad` uses the default time units (`"hours"`); therefore, explicit specification of the `n_breaks` and `timeu` arguments is not required.

```{r plot-dvtime-breaks, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = "ODV") 
```
However, perhaps someone on the team would prefer the x-axis breaks in units of `days`. The x-axis breaks will transform to the new units automatically as long as we specify the new time unit with `timeu = "days"`.

```{r plot-dvtime-breaks-days, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_data_pk_days <- plot_data_pk %>% 
  mutate(TIME = TIME/24, 
         NTIME = NTIME/24)

plot_dvtime(data = plot_data_pk_days, dv_var = "ODV", timeu = "days") 
```

Nice! However, someone else on the team would prefer to see the first 24 hours of treatment in greater detail to visualize the absorption phase. We can either truncate the x-axis range using `scale_x_continuous()`, or filter the input data and allow the x-axis breaks to adjust automatically with the new time range in the input data!

```{r plot-dvtime-breaks-24h, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_data_24 <- plot_data_pk %>% 
  filter(NTIME <= 24)

plot_dvtime(data = plot_data_24, dv_var = "ODV") 
```


## Color Aesthetic and Central Trendency

The color aesthetic can be mapped to a dataset variable using the the `col_var` argument. Let's use a variable we defined earlier using `df_addn` to list unique study conditions.

```{r plot-dvtime-col, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = "ODV", log_y = TRUE, col_var = "Dose and Food", )
```
The argument `cent` specifies the method of calculating the central tendency (+/- variability) within levels of the variable passed to the `col_var`. The default is `cent = "mean"`; however, note that the calculation performed are *after* any transformations to the data and this option will return the geometric mean when `log_y=TRUE`. If the `log_y` argument is used to generate semi-log plots along with `show_captions = TRUE`, then the caption will automatically delineate where arithmetic and geometric means are being returned. 

```{r plot-dvtime-part-log, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = "ODV", col_var = "Dose and Food", cent = "mean", 
            ylab = "Concentration (ng/mL)") 
```

It looks like coadministration with food may impact the absorption profile. Luckily, `plot_dvtime` returns a `ggplot` object which we can modify like any other `ggplot`! Therefore, we can facet by PART by simply adding in another layer to our `ggplot` object. 

```{r plot-dvtime-part, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = "ODV", col_var = "Dose and Food", cent = "mean", 
            ylab = "Concentration (ng/mL)", log_y = TRUE) +
  facet_wrap(~PART)
```

The clinical team would like a simpler plot that clearly displays the central tendency. We can use the argument `cent = "mean_sdl"` to plot the mean with error bars and remove the observed points by specifying `obs_dv = FALSE`.

```{r plot-dvtime_part-meansdl, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = "ODV", col_var = "Dose and Food", cent = "mean_sdl", 
            ylab = "Concentration (ng/mL)", log_y = TRUE,
            obs_dv = FALSE) +
  facet_wrap(~PART)
```

We may want to only show the upper error bar, especially when computing the arithmetic mean +/- arithmetic SD on the linear scale. This can be accomplished by changing the `cent` argument to `mean_sdl_upper`.

```{r plot-dvtime_part-meansdl-upper, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = "ODV", col_var = "Dose and Food", cent = "mean_sdl_upper", 
            ylab = "Concentration (ng/mL)", obs_dv = FALSE) +
  facet_wrap(~PART)
```

We could also plot these data as median + interquartile range (IQR), if we do not feel the sample size is sufficient for parametric summary statistics. This can be accomplished by changing the `cent` argument to `median_iqr`.

```{r plot-dvtime_part-medianiqr, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = "ODV", col_var = "Dose and Food", cent = "median_iqr", 
            ylab = "Concentration (ng/mL)", log_y = TRUE,
            obs_dv = FALSE) +
  facet_wrap(~PART)
```

Hmm...there is some noise at the late terminal phase. This is likely artifact introduced by censoring of data at the assay LLOQ; however, let's confirm there are no weird individual subject profiles by connecting observed data points longitudinally within a subject - in other words, let's make spaghetti plots! 

We will change the central tendency measure to the median and add the spaghetti lines. Data points within an individual value of `grp_var` will be connected by a narrow line when `grp_dv = TRUE`. The default is `grp_var = "ID"`.

```{r plot-dvtime-part-ind, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = "ODV", col_var = "Dose and Food", cent = "median", 
            ylab = "Concentration (ng/mL)", log_y = TRUE, 
            grp_dv = TRUE) +
  facet_wrap(~PART)
```
It does not seem like there are outlier individuals driving the noise in the late terminal phase; therefore, this is almost certainly artifact introduced by data missing due to assay sensitivity and censoring at the lower limit of quantification (LLOQ). 

## Defining imputations for BLQ data

Let's use imputation to assess the potential impact of the data missing due to assay sensitivity. `plot_dvtime` includes some functionality to do this imputation for us using the `loq` and `loq_method` arguments.

The `loq_method` argument species how BLQ imputation should be performed. Options are:

  + `0` : No handling. Plot input dataset `DV` vs `TIME` as is. (default)
  + `1` : Impute all BLQ data at `TIME` <= 0 to 0 and all BLQ data at `TIME` > 0 to 1/2 x `loq`.
      Useful for plotting concentration-time data with some data BLQ on the linear scale
  + `2` : Impute all BLQ data at `TIME` <= 0 to 1/2 x `loq` and all BLQ data at `TIME` > 0 to 1/2 x `loq`. 
  
The `loq` argument species the value of the LLOQ. The `loq` argument must be specified when `loq_method` is `1` or `2`, but can be `NULL` *if* the variable `LLOQ` is present in the dataset. In our case, `LLOQ` is a variable in `plot_data`, so we do not need to specify the `loq` argument (default is `loq = NULL`).

```{r plot-dvtime-blq1, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(plot_data_pk, dv_var = "ODV", col_var = "Dose and Food", cent = "mean",
            ylab = "Concentration (ng/mL)", log_y = TRUE,
            loq_method = 2) +
  facet_wrap(~PART)
```
The same plot is obtained by specifying `loq = 1`

```{r plot-dvtime-blq2, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(plot_data_pk, dv_var = "ODV", col_var = "Dose and Food", cent = "mean",
            ylab = "Concentration (ng/mL)",  log_y = TRUE,
            loq_method = 2, loq = 1) +
  facet_wrap(~PART)
```

A reference line is drawn to denote the LLOQ and all observations with `EVID=0` and `MDV=1` are imputed as LLOQ/2. The numeric value of LLOQ is printed in the legend and a caption is added to indicate the imputation method for BLQ data.

Imputing post-dose concentrations below the lower limit of quantification as 1/2 x LLOQ normalizes the late terminal phase of the concentration-time profile. This is confirmatory evidence for our hypothesis that the noise in the late terminal phase is due to censoring of observations below the LLOQ.

## Dose-normalization

We can also generate dose-normalized concentration-time plots by specifying `dosenorm = TRUE`. 

```{r plot-dvtime-dn, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(plot_data_pk, dv_var = "ODV", col_var = "Dose and Food", cent = "mean",
            ylab = "Dose-normalized Conc. (ng/mL per mg Drug)", log_y = TRUE,
            dosenorm = TRUE) +
  facet_wrap(~PART)
```

When `dosenorm = TRUE`, the variable specified in `dose_var` (default = "DOSE") needs to be present in the input dataset `data`. If `dose_var` is not present in `data`, the function will return an *Error* with an informative error message.

```{r plot-dvtime-dn-nodose, fig.height=4, fig.width=6, message=TRUE, warning=TRUE, error=TRUE}
plot_dvtime(select(plot_data_pk, -DOSE), 
            dv_var = "ODV", col_var = "Dose and Food", cent = "mean",
            ylab = "Dose-normalized Conc. (ng/mL per mg Drug)", log_y = TRUE,
            dosenorm = TRUE) +
  facet_wrap(~PART)
```

Dose-normalization is performed *AFTER* BLQ imputation in the case in which both options are requested. The reference line for the LLOQ will not be plotted when dose-normalized concentration is the dependent variable. 

```{r plot-dvtime-blq-dn, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(plot_data_pk, dv_var = "ODV", col_var = "Dose and Food", cent = "mean",
            ylab = "Dose-normalized Conc. (ng/mL per mg Drug)", log_y = TRUE,
            loq_method = 2, dosenorm = TRUE) +
  facet_wrap(~PART)
```


## Adjusting the Attributes for Points and Lines

The default attributes of data points and lines are controlled by the `theme` argument. The defaults

```{r plot-dvtime-theme, message = FALSE, warning=FALSE}
plot_dvtime_theme()
```

These attributes can be updated by passing a named list to the `theme` argument. Say we want to reduce the linewidth of the error bars and reduce the size of the mean summary points to only visualize the lines. This can be accomplished for an individual plot as follows:

```{r plot-dvtime-part-meansdl-theme, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = "ODV", col_var = "Dose and Food", cent = "mean_sdl", 
            ylab = "Concentration (ng/mL)", log_y = TRUE,
            obs_dv = FALSE, 
            theme = list(linewidth_errorbar = 0.5, size_point_cent = 0.1)) +
  facet_wrap(~PART)
```

One could also globally set a new theme by updating the default using `plot_dvtime_theme` and pass the new theme list object to the `theme` argument. This is useful if generating multiple plots using the same modified theme.

```{r plot-dvtime-part-meansdl-themeset, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
new_theme <- plot_dvtime_theme(list(linewidth_errorbar = 0.5, size_point_cent = 0.1))

plot_dvtime(data = plot_data_pk, dv_var = "ODV", col_var = "Dose and Food", cent = "mean_sdl", 
            ylab = "Concentration (ng/mL)", log_y = TRUE,
            obs_dv = FALSE, 
            theme = new_theme) +
  facet_wrap(~PART)
```

The default error bar width is 2.5% of the maximum nominal time in the dataset. This can be overwritten to a user-specified value using the `width_errorbar` attribute of `plot_dvtime_theme`. This value is passed to the `width` argument of `geom_errorbar`. 

```{r plot-dvtime-part-meansdl-width, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data, dv_var = "ODV", col_var = "Dose and Food", cent = "mean_sdl", 
            ylab = "Concentration (ng/mL)", log_y = TRUE,
            obs_dv = FALSE, 
            theme = list(width_errorbar = 8)) +
  facet_wrap(~PART)
```

# Individual Concentration-time plots

The previous section provides an overview of how to generate population concentration-time profiles by dose using `plot_dvtime`; however, we can also use `plot_dvtime` to generate subject-level visualizations with a little pre-processing of the input dataset.

We can specify `cent = "none"` to remove the central tendency layer when plotting individual subject data. 

```{r plot-dvtime-blq2-indonly, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(plot_data_pk, dv_var = "ODV", col_var = "Dose and Food", cent = "none",
            ylab = "Concentration (ng/mL)", log_y = TRUE,
            grp_dv = TRUE,
            loq_method = 2, loq = 1) +
  facet_wrap(~PART)
```
We can plot an individual subject by filtering the input dataset. This could be extended generate plots for all individuals using `for` loops, `lapply`,  `purrr::map()` functions, or other methods.

```{r plot-dvtime-blq2-indplots, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}

ids <- sort(unique(plot_data_pk$ID)) #vector of unique subject ids
n_ids <- length(ids) #count of unique subject ids
plots_per_pg <- 4
n_pgs <- ceiling(n_ids/plots_per_pg) #Total number of pages needed

plist<- list()
for(i in 1:n_ids){
  plist[[i]] <- plot_dvtime(filter(plot_data_pk, ID == ids[i]), 
                               dv_var = "ODV", cent = "none",
            ylab = "Concentration (ng/mL)", log_y = TRUE,
            grp_dv = TRUE,
            loq_method = 2, loq = 1, show_caption = FALSE) +
  facet_wrap(~PART)+
  labs(title = paste0("ID = ", ids[i], " | Dose = ", unique(plot_data$DOSE[plot_data_pk$ID==ids[i]]), " mg"))+
  theme(legend.position="none")
}

lapply(1:n_pgs, function(n_pg) {
      i <-  (n_pg-1)*plots_per_pg+1
      j <- n_pg*plots_per_pg
      wrap_plots(plist[i:j])
})
```

# Population Response-Concentration Plots

## Overview of `plot_dvconc`

Now let's take a look at the pharmacodynamic (PD) data! We can visualize the response-time profile using `plot_dvtime`, just as we did to visualize the concentration-time profile previously, by filtering our PK/PD dataset to the compartment with the biomarker observations (`CMT=3`).

```{r plot-dvtime-pd, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = filter(plot_data_pd, CMT == 3), dv_var = "ODV", col_var = "Dose and Food", ylab = "Response") 
```

Now let's visualize the response data versus drug concentration!. `pmxhelpr` includes a function for common visualizations of observed response versus drug concentration data in exploratory data analysis: `plot_dvconc`

In our visualizations, we will leverage the following dataset variables: 

  + `ODV`: the original dependent variable (biomarker response) normalized to baseline
  + `CONC` : time-matched, observed drug concentration at the time of biomarker assessment 
  
`plot_dvconc` requires a dependent variable, specified as string via the `dv_var` argument, and an independent variable, specified as a string via the `idv_var` argument. The default values are `"DV"` and `"CONC"`.

```{r plot-dvconc, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT==3), dv_var = "ODV", idv_var = "CONC") 
```

## Color Aesthetic

`plot_dvconc` includes two arguments for controlling the color aesthetic, `col_var` and `col_trend`. If a variable is passed as a string to the `col_var` argument, the data points are colored based on this variable; however, by default `col_trend = FALSE` and the trend line is fit to the totality of the data  without stratifying the trend lines by the variable mapped to the color aesthetic.

```{r plot-dvconc-col, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT == 3), dv_var = "ODV", idv_var = "CONC", 
            col_var = "Dose and Food", col_trend = FALSE) 
```

Trend lines stratified by the variable mapped to the color aesthetic are requested by setting `col_trend = TRUE`.

```{r plot-dvconc-col-trend, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT == 3), dv_var = "ODV", idv_var = "CONC", 
            col_var = "Dose and Food", col_trend = TRUE) 
```

## Central Tendency

There are two trend line types supported by `plot_dvconc`: locally estimated scatter plot smoothing (LOESS) fit and linear regression. The central tendency trend lines visualized are controlled by logical arguments `loess` and `linear`. The default is `loess = TRUE` and `linear = FALSE`

```{r plot-dvconc-col-trend-loess-linear, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT == 3), dv_var = "ODV", idv_var = "CONC", 
            col_var = "Dose and Food", col_trend = FALSE, loess = TRUE, linear = TRUE) 
```
The confidence intervals of the trendlines are suppressed by defaul in order to facilitate visualization of the central tendency and spread of observed data points simultaneously. Confidence intervals can be added to the plot using the logical arguments `se_loess` and `se_linear`.

```{r plot-dvconc-loess-linear-se, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT == 3), dv_var = "ODV", idv_var = "CONC", 
            col_var = "Dose and Food", col_trend = FALSE, 
            loess = TRUE, linear = TRUE, se_loess = TRUE, se_linear = TRUE) 
```
Additional arguments can be passed to `geom_smooth(method = "loess")`, such as increasing the span of the smoothing fit.

```{r plot-dvconc-loess-span, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT == 3), dv_var = "ODV", idv_var = "CONC", 
            col_var = "Dose and Food", col_trend = FALSE, 
            loess = TRUE, linear = FALSE, se_loess = TRUE, se_linear = FALSE, 
            span = 1) 
```
If the color aesthetic is mapped to the trendlines with `col_trend = TRUE`, it will also map to the ribbons defining the 95% CIs.

```{r plot-dvconc-col-trend-se, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT == 3), dv_var = "ODV", idv_var = "CONC", 
            col_var = "Dose and Food", col_trend = TRUE,
            se_loess = TRUE) 
```


# Population Concentration-Time and Response-Time

The wrapper function `plot_dvtime_dual` is provided to support simultaneous plotting of PK and PD profiles for visualization of trends in time. This function takes two  arguments, `dv_var1` and `dv_var2` to specify the variable to plot as the dependent variable for each of the top and bottom panels. It also takes the `dvid_var` argument to specify the variable that contains the identifiers for each dependent variable to be plotted, as well as `dvid_val1` and `dvid_val2` arguments to specify the values of the variable in `dvid_var` to be used identify each set of data for plotting.

The default values for `dv_var1` and `dv_var2` are `"DV"`, the default `dvid_var` is `"CMT"` with default values of `dvid_val1 = 2` and `dvid_val2 = 3`.

```{r plot-dvtime-dual-col, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime_dual(plot_data_pd, dv_var1 = "ODV",dv_var2 = "ODV",
                 dvid_var = "CMT", dvid_val1 = 2, dvid_val2 = 3,
                 col_var = "Dose") 
```

Both y-axis transformations and labels  can be specified separately.

```{r plot-dvtime-dual-col-y12, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime_dual(plot_data_pd, dv_var1 = "ODV", dv_var2 = "ODV",
                 dvid_var = "CMT", dvid_val1 = 2, dvid_val2 = 3,
                 col_var = "Dose",
                 log_y1 = TRUE, ylab1 = "Drug Conc. (ng/mL)", ylab2 = "Response (% Base)") 
```

However, the top panel (DV1) is presumed to apply to PK and the bottom panel (DV2) is presumed to apply to PD. Therefore, BLQ imputation and dose-normalization options can only be applied to the top panel (DV1). 

```{r plot-dvtime-dual-col-y1, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime_dual(plot_data_pd, dv_var1 = "ODV", dv_var2 = "ODV",
                 dvid_var = "CMT", dvid_val1 = 2, dvid_val2 = 3,
                 col_var = "Dose", loq_method = 2, dosenorm = T,dose_var = "DOSE",
                 log_y1 = TRUE, ylab1 = "Dose-norm Drug Conc. (ng/mL)", ylab2 = "Response (% Base)") 
```

Similarly, specification of a reference line at y=`cfb_base` when `cfb=TRUE` is only supported for the bottom panel (DV2).

```{r plot-dvtime-dual-col-y2, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime_dual(plot_data_pd, dv_var1 = "ODV", dv_var2 = "ODV",
                 dvid_var = "CMT", dvid_val1 = 2, dvid_val2 = 3,
                 col_var = "Dose", cfb=TRUE, cfb_base = 100,
                 log_y1 = TRUE, ylab1 = "Drug Conc. (ng/mL)", ylab2 = "Response (% CFB)") 

plot_dvtime_dual(plot_data_pd, dv_var1 = "ODV", dv_var2 = "CFB",
                 dvid_var = "CMT", dvid_val1 = 2, dvid_val2 = 3,
                 col_var = "Dose", cfb=TRUE,
                 log_y1 = TRUE, ylab1 = "Drug Conc. (ng/mL)", ylab2 = "Response (% CFB)") 
```

# Dose-proportionality Assessment: Power Law Regression

Another assessment that is commonly performed for pharmacokinetic data is dose proportionality (e.g., does exposure increase proportionally with dose). This is an important assessment prior to population PK modeling, as it informs whether non-linearity is an important consideration in model development. 

The industry standard approach to assessing dose proportionality is power law regression. Power law regression is based on the following relationship:

$$Exposure = \alpha*(DOSE)^\gamma$$

This power relationship can be transformed to a linear relationship to support quantitative estimation of the power ($\gamma$) via simple linear regression by taking the logarithm of both sides:

$$log(Exposure) = intercept + \gamma*log(DOSE)$$

`NOTE`: Use of natural logarithm and log10 transformations will not impact the assessment of the power and will only shift the intercept. 

This approach facilitates hypothesis testing via assessment of the 95% CI around the power ($\gamma$) estimated from the log-log regression. The null hypothesis is that exposure increases proportionally to dose (e.g., $\gamma=1$) and the alternative hypothesis is that exposure does *NOT* increase proportionally to dose (e.g., $\gamma\ne 1$). 

Interpretation of the relationship is based on the 95% CI of the $\gamma$ estimate as follows:

  + 95% CI includes one (1): exposure increases proportionally to dose
  + 95% CI excludes one (1) & is less than 1: exposure increases less-than-proportionally to dose
  + 95% CI excludes one (1) & is greater than 1: exposure increases greater-than-proportionally to dose
  
This assessment is generally performed based on both maximum concentration (Cmax) and area under the concentration-time curve (AUC). While not a hard and fast rule, some inference can be drawn about which phase of the pharmacokinetic profile is most likely contributing the majority of the non-linearity of exposure with dose. 

  + AUC = *NOT* dose-proportional | Cmax = dose-proportional = elimination phase
  + AUC = dose-proportional | Cmax = *NOT* dose-proportional = absorption phase (rate)
  + AUC = *NOT* dose-proportional | Cmax = *NOT* dose-proportional = absorption phase (extent)
  
These exploratory assessments provide quantitative support for structural PK model decision-making. Practically speaking, non-linearities in absorption rate are rarely impactful, and the modeler is really deciding between dose-dependent bioavailability and concentration-dependent elimination (e.g., Michaelis-Menten kinetics, target-mediated drug disposition [TMDD])
  
## Step 1: Derive NCA Parameters

The first step in performing this assessment is deriving the necessary NCA PK parameters. NCA software (e.g., Phoenix WinNonlin) is quite expensive; however, thankfully there is an excellent R package for performing NCA analyses - `PKNCA`.

Refer to the documentation for the `PKNCA` packge for details. This vignette will not provide a detailed overview of `PKNCA` functions and workflows.

First, let's set the options for our NCA analysis and define the intervals over which we want to obtain the NCA parameters. The `data_sad` dataset internal to `pmxhelpr` is a single ascending dose (SAD) design with a parallel food effect (FE) cohort; therefore, our interval is [0, $\infty$]

```{r nca-set-up}
##Set NCA options
PKNCA.options(conc.blq = list("first" = "keep", 
                              "middle" = unique(data_sad$LLOQ[!is.na(data_sad$LLOQ)]), 
                              "last" = "drop"),
              allow.tmax.in.half.life = FALSE,
              min.hl.r.squared = 0.9)

##Calculation Intervals and Requested Parameters
intervals <-
  data.frame(start = 0,
             end = Inf,
             auclast = TRUE,
             aucinf.obs = TRUE,
             aucpext.obs = TRUE,
             half.life = TRUE,
             cmax = TRUE,
             vz.obs = TRUE, 
             cl.obs = TRUE 
             )  
```

Next, we will set up our dose and concentration objects and perform the NCA using `PKNCA`

```{r nca-output}
#Impute BLQ concentrations to 0 (PKNCA formatting)
data_sad_nca_input <- data_sad %>% 
  mutate(CONC = ifelse(is.na(ODV), 0, ODV), 
         AMT = AMT/1000) #Convert from mg to ug (concentration is ng/mL = ug/L)

#Build PKNCA objects for concentration and dose including relevant strata
conc_obj <- PKNCAconc(filter(data_sad_nca_input, EVID==0), CONC~TIME|ID+DOSE+PART)
dose_obj <- PKNCAdose(filter(data_sad_nca_input, EVID==1), AMT~TIME|ID+PART)
nca_data_obj <- PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_results_obj <- as.data.frame(pk.nca(nca_data_obj))
glimpse(nca_results_obj)
```

The NCA results object output from `PKNCA` is formatted using the variable names in `SDTM` standards for the `PP` domain (Pharmacokinetic Parameters). This NCA output dataset is also available internally within `pmxhelpr` as `data_sad_nca` with a few additional columns specifying units.

```{r nca-data}
glimpse(data_sad_nca)
```

We will need to select the relevant PK parameters from this dataset for input into our power law regression analysis of dose-proportionality. Thankfully, `pmxhelpr` handles the filtering and power law regression in one step with functions for outputting either tables or plots of results!

## Step 2: Perform Power Law Regression

The `df_doseprop` function is a wrapper function which bundles two other `pmxhelpr` functions:

  + `mod_loglog` a function to perform log-log regression which returns a `lm` object
  + `df_loglog` a function to tabulate the power estimate and CI which returns a `data.frame`
  
There are two required arguments to `df_doseprop`.

  + `data` a `data.frame` containing NCA parameter estimates
  + `metrics` a character vector of NCA parameters to evaluate in log-log regression

```{r power-law-table}
power_table <- df_doseprop(data_sad_nca, metrics = c("aucinf.obs", "cmax"))
power_table
```

The table includes the relevant estimates from the power law regression (intercept, standard error, power, lower confidence limit, upper confidence limit), as well as, a logical flag for dose-proportionality and text interpretation.

Based on this assessment, these data appear dose-proportional for both Cmax and AUC! However, we should not include the food effect part of the study in this assessment, as food could also influence these parameters, and confounds the assessment of dose proportionality. The most important thing is to understand the input data!

Let's run it again, but this time only include `Part 1-SAD`.

```{r power-law-table-nofe}
power_table <- df_doseprop(filter(data_sad_nca, PART == "Part 1-SAD"), metrics = c("aucinf.obs", "cmax"))
power_table
```

In this case, the interpretation is unchanged with and without inclusion of the food effect cohort. `df_doseprop` provides two arguments for defining the confidence interval.

  + `method`: method to derive the upper and lower confidence limits. The default is `"normal"`, specifying use of the normal distribution, with `"tdist"` as an alternative, specifying use of the t-distribution. The t-distribution is preferred for analyses with smaller sample sizes
  + `ci`: width of the confidence interval. The default is `0.95` (95% CI) with `0.90` (90% CI) as an alternative

## Step 3: Visualize the Power Law Regression

We can also visualize these data using the `plot_doseprop` function. This function leverages the linear regression option within `ggplot2::geom_smooth()` to perform the log-log regression for visualization and pulls in the functionality of `df_doseprop` to extract the power estimate and CI into the facet label.

The required arguments to `plot_doseprop` are the same as `df_doseprop`!

```{r power-law-plot}
plot_doseprop(filter(data_sad_nca, PART == "Part 1-SAD"), metrics = c("aucinf.obs", "cmax"))
```
