---
title: "Exploratory Analyses of PK/PD Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eda-pkpd-workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will demonstrate `pmxhelpr` functions for exploratory data analysis of dose-concentration-response-time data to inform population PK/PD modeling analysis. This includes package functions to visualize response over time, response over time and concentration over time simultaneously, and response versus concentration.

This vignette will assume familiarity with the `plot_dvtime` function, as described in the Exploratory Analyses of PK Data vignette. This functionality will not be reviewed in detail in this vignette.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(forcats, warn.conflicts = FALSE)
library(Hmisc, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
library(PKNCA, warn.conflicts = FALSE)
```

# Data

The example dataset used in this vignette is based on a single ascending dose (SAD) study of an orally administered drug product with a parallel group food effect (FE) cohort.

## data_sad

Dataset definitions can be viewed by calling `?data_sad`. We can take a quick look at the modeling-ready analysis dataset (`data_sad`)  using `glimpse()` from the `dplyr` package.

```{r data-sad-glimpse}
glimpse(data_sad)
```

The study design consists of two parts. 

  + Part 1 was a sequential single ascending dose study covering a 10 to 400 mg dose range.
  + Part 2 was a parallel 100 mg single dose food effect (FE) study. 

The dataset is formatted for population pharmacokinetic (PopPK) and pharmacokinetic/pharmacodynamic (PK/PD) modeling in NONMEM with the following compartments and data types:

  - `CMT=1`: oral dose events (`EVID=1`) 
  - `CMT=2`: drug concentration observations (`EVID=0`) 
  - `CMT=3`: pharmacodynamic response observations (`EVID=0`) 
  
Dose events are input based on `AMT`. The nominal (e.g. protocol assigned) dose associated with each observation is captured in `DOSE`.

Plasma drug concentration and pharmacodynamic response observations are expressed in multiple units:  

  - `ODV`: original units of the dependent variable [`CMT=2` ng/mL, `CMT=3` % baseline activity]
  - `LDV`: logarithm-transformed drug concentration [`CMT = 2` log(ng/mL), `CMT = 3` % baseline activity]
  - `CFB`: percentage change from baseline [`CMT=2` missing, `CMT=3` % change from baseline]

The dataset also contains multiple variables for time:

  - `TIME`: Actual time since first dose administration [hours]
  - `NTIME`: Nominal time dose event or sample collection per protocol [hours]
  - `NDAY`: Nominal day on study [day]

As in Exploratory Analyses of PK Data, we will define some new variables to pass to the color aesthetic that include counts of unique individuals per level of the grouping variable with `var_addn`.

```{r plot-data, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_data <- data_sad %>% 
  filter(EVID == 0) %>% 
  mutate(`Food Status` = ifelse(FOOD == 0, "Fasted", "Fed"), 
         `Dose and Food` = paste(DOSE, "mg", `Food Status`)) %>% 
  mutate(Dose = var_addn(DOSE, ID, sep = "mg"),
         `Food Status` = var_addn(`Food Status`, ID),
         `Dose and Food` = var_addn(`Dose and Food`, ID)) %>% 
  mutate(Dose = fct_relevel(Dose, "50 mg (n=6)", after = 1), 
         `Dose and Food` = fct_relevel(`Dose and Food`, "50 mg Fasted (n=6)", after = 1))
```

This time, let's filter down only to only PD observations for this exploratory analysis, which is focused on assessing the informational content of the data and trends for PK/PD analysis.

```{r plot-data-cmt, message=FALSE, warning=FALSE}
plot_data_pk <- filter(plot_data, CMT == 2)
plot_data_pd <- filter(plot_data, CMT == 3)
```


# Population Concentration-time and Response-time Plots with `plot_dvtime_dual`

Now let's take a look at the pharmacodynamic (PD) data! We can visualize the response-time profile using `plot_dvtime`, just as we did to visualize the concentration-time profile previously, by filtering our PK/PD dataset to the compartment with the biomarker observations (`CMT=3`).

```{r plot-dvtime-pd, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pd, dv_var = ODV, col_var = `Dose and Food`) +
  labs(y = "Response (% of Baseline)", x = "Time Since First Douse (hours)")
```

We can add a reference line to our plot using the `plot_dvtime` argument `cfb=TRUE`, which by default plots a reference line at y=0. The location of the reference line can be modified by specifying a y-intercept value using the argument`cfb_base`. 

```{r plot-dvtime-pd-ref, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pd, dv_var = ODV,
            col_var = `Dose and Food`,
            cfb = TRUE, cfb_base = 100) +
  labs(y = "Response (% of Baseline)", x = "Time Since First Douse (hours)")
```

We can add a reference line to our plot using the `plot_dvtime` argument `cfb=TRUE`, which by default plots a reference line at y=0. The location of the reference line can be modified by specifying a y-intercept value using the argument`cfb_base`. 

```{r plot-dvtime-pd-ref0, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pd, dv_var = CFB,
            col_var = `Dose and Food`,
            cfb=TRUE) +
   labs(y = "Response (% of Baseline)", x = "Time Since First Douse (hours)")
```


## Overview of `plot_dvtime_dual`

The wrapper function `plot_dvtime_dual` supports simultaneous visualization of the time-course of drug concentration and response. 
This function uses the `patchwork` package to plot the PK and PD profiles in two vertically arranged panels. `plot_dvtime_dual` includes the following arguments to specify elements of the panels:
  
  + `dv_var1` and `dv_var2` arguments to specify the dependent variable (y-axis) to plot in the top (1) and bottom (2) panels. The default is `DV` .
  + `dvid_var` argument to specify the variable that contains the identifiers for each dependent variable to be plotted. The default is `CMT`.
  + `dvid_val1` and `dvid_val2` arguments to specify the values of the variable in `dvid_var` to identify the dependent variables to plot in the top (1) and bottom (2) panel. The defaults are `2` and `3`, respectively.
  
Both `dv_var1` and `dv_var2` arguments use non-standard evaluation and can be passed as bare column names or strings.

```{r plot-dvtime-dual-col, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime_dual(plot_data, dv_var1 = ODV,dv_var2 = ODV,
                 dvid_var = CMT, dvid_val1 = 2, dvid_val2 = 3,
                 col_var = `Dose and Food`) + 
    labs(y = "Response (% of Baseline)", x = "Time Since First Douse (hours)")
```

The following arguments are available to specify y-axis labels and transformations:

  + `log_y1`: Logical specifying if the top panel y-axis should be transformed to a log scale. Default is `FALSE`.
  + `log_y2`: Logical specifying if the bottom panel y-axis should be transformed to a log scale. Default is `FALSE`.

```{r plot-dvtime-dual-col-y12, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime_dual(plot_data, dv_var1 = ODV, dv_var2 = ODV,
                 dvid_var = CMT, dvid_val1 = 2, dvid_val2 = 3,
                 col_var = `Dose and Food`,
                 log_y1 = TRUE)
```

This wrapper function expects that the top panel will display drug concentration (PK) and the bottom panel will display response (PD). Thus, BLQ imputation and dose-normalization options specified using the standard arguments to `plot_dvtime` are only applied to the top panel.

```{r plot-dvtime-dual-col-dosenorm-blq, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime_dual(plot_data, dv_var1 = ODV, dv_var2 = ODV,
                 dvid_var = CMT, dvid_val1 = 2, dvid_val2 = 3,
                 col_var = `Dose and Food`, loq_method = 2, dosenorm = T,dose_var = DOSE,
                 log_y1 = TRUE)
```

Similarly, since function expects that the bottom panel displays response (PD), a reference line at y=`cfb_base` is only added to the bottom panel when `cfb=TRUE`. 

```{r plot-dvtime-dual-col-ref, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime_dual(plot_data, dv_var1 = ODV, dv_var2 = ODV,
                 dvid_var = CMT, dvid_val1 = 2, dvid_val2 = 3,
                 col_var = Dose, cfb=TRUE, cfb_base = 100,
                 log_y1 = TRUE)
```

The default value of `cfb_base` is `0`, which is the reference condition for when we express our response variable in units of change from baseline or percentage change from baseline. We can modify our plot to use change from baseline as the dependent variable for the PD panel but updating the variable passed to `dv_var2`.

```{r plot-dvtime-dual-col-ref-cfb, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime_dual(plot_data, dv_var1 = ODV, dv_var2 = CFB,
                 dvid_var = CMT, dvid_val1 = 2, dvid_val2 = 3,
                 col_var = Dose, cfb=TRUE,
                 log_y1 = TRUE)
```

Finally, we can modify both of the captions separately using the logical arguments `show_caption1` and `show_caption2`. Their default values are `TRUE`. Since both panels of this plot have the same elements, we can turn the caption off in the top panel and print a single caption under the bottom panel to represent both panels in the plot.

```{r plot-dvtime-dual-col-ref-cfb-nocap, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime_dual(plot_data, dv_var1 = ODV, dv_var2 = CFB,
                 dvid_var = CMT, dvid_val1 = 2, dvid_val2 = 3,
                 col_var = Dose, cfb=TRUE,
                 log_y1 = TRUE,
                 show_caption1 = FALSE, show_caption2 = TRUE)
```


# Population Response-Concentration Plots

## Overview of `plot_dvconc`

Now that we have visualized our longitudinal PK and PD profiles, let's visualize the PD data versus drug concentration! `plot_dvconc` is a plotting function for visualizing the relationship between a dependent variable for response and drug concentration.
  
`plot_dvconc` requires the following arguments which specify the variables to be plotted:

  + `dv_var`: character string specifying the dependent variable to map to the y-axis. Default is `DV`.
  + `idv_var`: character string specifying the dependent variable to map to the y-axis. Default is `CONC`.
  
Both arguments use non-standard evaluation and can be passed as bare column names or strings.

```{r plot-dvconc, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT==3), dv_var = ODV, idv_var = CONC) +
  labs(x = "Drug Conc. (ng/mL)", y = "Response (% of Baseline)")
```

Like `plot_dvtime`, we can specify a reference line using `cfb = TRUE`, which is plotted at `cfb_base` (default = 0).

```{r plot-dvconc-ref, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT==3), dv_var = CFB, idv_var = CONC,
            cfb = TRUE) +
  labs(x = "Drug Conc. (ng/mL)", y = "Response (% Change from Baseline)")
```


## Color Aesthetic

`plot_dvconc` includes two arguments for controlling the color aesthetic, `col_var` and `col_trend`. If a variable is passed to the `col_var` argument, the data points are colored based on this variable; however, by default `col_trend = FALSE` and the trend line is fit to the totality of the data  without stratifying the trend lines by the variable mapped to the color aesthetic. The `col_var` argument uses non-standard evaluation and can be passed as a bare column name or as a string.

```{r plot-dvconc-col, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT == 3), dv_var = CFB, idv_var = CONC,
            cfb = TRUE,
            col_var = `Dose and Food`, col_trend = FALSE) +
  labs(x = "Drug Conc. (ng/mL)", y = "Response (% Change from of Baseline)")
```

Trend lines stratified by the variable mapped to the color aesthetic are requested by setting `col_trend = TRUE`.

```{r plot-dvconc-col-trend, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT == 3), dv_var = CFB, idv_var = CONC,
           cfb = TRUE,
           col_var = `Dose and Food`, col_trend = TRUE) +
  labs(x = "Drug Conc. (ng/mL)", y = "Response (% Change from  Baseline)")
```

## Central Tendency

There are two trend line types supported by `plot_dvconc`: locally estimated scatter plot smoothing (LOESS) fit and linear regression. The central tendency trend lines visualized are controlled by logical arguments `loess` and `linear`. The default is `loess = TRUE` and `linear = FALSE`

```{r plot-dvconc-col-trend-loess-linear, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT == 3), dv_var = CFB, idv_var = CONC,
            cfb = TRUE, col_var = `Dose and Food`, col_trend = FALSE, loess = TRUE, linear = TRUE) +
  labs(x = "Drug Conc. (ng/mL)", y = "Response (% Change from Baseline)")
```
The confidence intervals of the trend lines are suppressed by default in order to facilitate visualization of the central tendency and spread of observed data points simultaneously. Confidence intervals can be added to the plot using the logical arguments `se_loess` and `se_linear`.

```{r plot-dvconc-loess-linear-se, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT == 3), dv_var = CFB, idv_var = CONC,
            cfb = TRUE, col_var = `Dose and Food`, col_trend = FALSE,
            loess = TRUE, linear = TRUE, se_loess = TRUE, se_linear = TRUE) +
  labs(x = "Drug Conc. (ng/mL)", y = "Response (% Change from Baseline)")
```

Additional arguments can be passed to `geom_smooth(method = "loess")`, such as increasing the span of the smoothing fit.

```{r plot-dvconc-loess-span, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT == 3), dv_var = CFB, idv_var = CONC,
            cfb = TRUE, col_var = `Dose and Food`, col_trend = FALSE,
            loess = TRUE, linear = FALSE, se_loess = TRUE, se_linear = FALSE,
            span = 1) +
  labs(x = "Drug Conc. (ng/mL)", y = "Response (% Change from Baseline)")
```

If the color aesthetic is mapped to the trendlines with `col_trend = TRUE`, it will also map to the ribbons defining the CIs of the trend lines.

```{r plot-dvconc-col-trend-se, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = filter(plot_data_pd, CMT == 3), dv_var = CFB, idv_var = CONC,
            cfb = TRUE, col_var = `Dose and Food`, col_trend = TRUE,
            se_loess = TRUE) +
  labs(x = "Drug Conc. (ng/mL)", y = "Response (% Change of Baseline)")
```

