---
title: "Getting Started with pmxhelpr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pmxhelpr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

This vignette is a lean tour of every core `pmxhelpr` workflow in a single sitting. Each section ends with a link to the corresponding deep-dive article on the [pmxhelpr website](https://ryancrass.github.io/pmxhelpr/) for users who want more depth.

Exported functions follow a `ReturnType_Purpose` naming convention:

- `plot_*` — returns a `ggplot` object
- `df_*` — returns a `data.frame` (often class-tagged)
- `var_*` — returns a vector (vectorized helpers for use inside `mutate()`)
- `pmx_*` — returns a theme element constructor

The bundled datasets `data_sad` and `data_sad_pkfit` use `ODV` (original DV). Examples below pass `dv_var = "ODV"` or `dv_var = ODV` to highlight the non-standard evaluation (NSE) of dataset column name variables which may be specified as strings or bare names to override defaults in `pmxhelpr` functions. 

```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts = FALSE)
library(forcats, warn.conflicts = FALSE)
library(ggplot2, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
library(mrgsolve, warn.conflicts = FALSE)
```

# Data

## `data_sad`

`data_sad` is a NLME modeling analysis-ready dataset for single ascending dose (SAD) study with a parallel food-effect (FE) cohort. It contains one oral dose input (`EVID=1`, `CMT=1`) and two observation types (`EVID=0`): 

  + Drug concentration in `CMT=2` (original units: ng/mL)
  + Response in `CMT=3` (original units: percentage of baseline)

```{r data-sad}
glimpse(data_sad)
```

The preprocessing step below labels each participant with a dosing regimen and per-group subject count using `var_addn()`. We will also define separate PK and PD sets for use in downstream plotting functions.

```{r data-addn}
data <- data_sad %>%
  mutate(Food = ifelse(FOOD == 1, "Fed", "Fasted"),
         DoseFood = paste(DOSE, "mg x1", Food),
         Regimen = var_addn(DoseFood, ID))
unique(data$Regimen)
```

The resulting factor variable is inspected, and then re-leveled with `forcats::fct_relevel()` to preserve dose order for plotting.

```{r data-relevel}
data <- data %>% 
  mutate(Regimen = fct_relevel(Regimen, "50 mg x1 Fasted (n=6)", after = 1))
unique(data$Regimen)

data_pk <- data %>%
  filter(CMT %in% c(1,2))

data_pd <- data %>% 
  filter(CMT %in% c(1,3))
```

## `data_sad_nca`

`data_sad_nca` contains pharmacokinetic parameters and exposure metrics from a non-compartmental analysis (NCA) of `data_sad` using the `PKNCA` package.

We will filter to Part 1 fasted conditions only for use in dose-proportionality assessment.

```{r data-nca}
glimpse(data_sad_nca)
data_nca_part1 <- filter(data_sad_nca, PART == "Part 1-SAD")
```

## `data_sad_pkfit`

`data_sad_pkfit` is a model output dataset version of `data_sad` (`CMT` 1 and 2) with two additional variables (`PRED` and `IPRED`) appendend to the `data_sad`.  

These variables are derived from the internal PK model `pkmodel`.

```{r vpc-model, message=FALSE, warning=FALSE}
pkmodel <- model_mread_load("pkmodel")
see(pkmodel)
```
We will process this dataset in a manner analogous to `data_sad` for plotting, but adding subject counts to a derived dosing regimen variable.

```{r data-gof}
data_gof <- data_sad_pkfit %>%
  mutate(Food = ifelse(FOOD == 1, "Fed", "Fasted"),
         DoseFood = paste(DOSE, "mg x1", Food),
         Regimen = var_addn(DoseFood, ID)) %>%
  mutate(Regimen = fct_relevel(Regimen, "50 mg x1 Fasted (n=6)", after = 1))
```


# Longitudinal concentration and response with `plot_dvtime()`

`plot_dvtime()` produces a longitudinal observed-versus-time plot with central-tendency overlays, which was can be used for any longitudinal repeated measures continuous variable, including both concentration `CMT = 2`) and response (`CMT = 3`).

```{r plot-dvtime-pk, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
pk_plot <- plot_dvtime(
  data = data_pk,
  dv_var = ODV,
  cent = "mean_sdl",
  col_var = Regimen,
  log_y = TRUE,
  theme = plot_dvtime_theme(obs_point = pmx_point(alpha = 0))
) +
  labs(y = "Concentration (ng/mL)", x = "Time (hours)")
pk_plot
```

```{r plot-dvtime-pd, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
pd_plot <- plot_dvtime(
  data = data_pd,
  dv_var = "CFB",
  cent = "mean_sdl",
  col_var = Regimen,
  theme = plot_dvtime_theme(obs_point = pmx_point(alpha = 0))
) +
  labs(y = "Response (% Change from Baseline)", x = "Time (hours)")
pd_plot
```

The two plots can be composed into a single paneled figure using the `patchwork` package, aligned vertically with the shared time axis.

```{r plot-dvtime-patchwork, fig.width=8, fig.height=7, warning=FALSE, message=FALSE}
(pk_plot / pd_plot) + plot_layout(guides = "collect")
```

See the [Exploratory Analyses of PK and PK/PD Data](eda-pk-pkpd-workflow.html) article for central tendency controls, BLQ imputation moptions, dose-normalization, and other visual controls.

# Response versus concentration with `plot_dvconc()`

`plot_dvconc()` plots a dependent variable against a continuous independent variable. The most common use case is visualizing a biomarker value or change metric of response against drug concentration. Trend line options include both linear (`linear`/`se_linear`) and non-linear (`loess`/`se_loess`) logical toggles to control the central tendencey layer displayed.

A dashed black reference line is drawn at `y = ref` when `ref` is specified.

```{r plot-dvconc, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
plot_dvconc(
  data = filter(data, CMT == 3),
  dv_var = CFB,
  idv_var = CONC,
  ref = 0,
  col_var = Regimen,
  loess = TRUE,
  linear = TRUE
) +
  labs(y = "Response (% Change from Baseline)", x = "Concentration (ng/mL)")
```

By default the trend lines are not grouped by the variable passed to `col_var`; however, this can be toggled on by `col_trend=TRUE`.

```{r plot-dvconc-coltrend, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
plot_dvconc(
  data = filter(data, CMT == 3),
  dv_var = CFB,
  idv_var = CONC,
  ref = 0,
  col_var = Regimen,
  loess = TRUE,
  se_loess = TRUE,
  linear = FALSE,
  col_trend = TRUE
) +
  labs(y = "Response (% Change from Baseline)", x = "Concentration (ng/mL)")
```

See the [Exploratory Analyses of PK and PK/PD Data](eda-pk-pkpd-workflow.html) article for theme customization and other trend customization options.

# Dose-proportionality with `df_doseprop()` and `plot_doseprop()`

`df_doseprop()` fits a log-log regression of exposure metrics (e.g., Cmax and AUC) versus and returns a class-tagged `doseprop_stats` data frame. 

```{r doseprop-stats}
tab <- df_doseprop(data_nca_part1, metrics = c("aucinf.obs", "cmax"))
tab
```

`plot_doseprop()` can directly process an an NCA input dataset (1-stage) or accept a previously process `doseprop_stats` object (2-stage)

  1. One-stage: raw NCA `data.frame` input (`df_doseprop()` called internally)
  
```{r plot-doseprop-1step, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
plot_doseprop(data_nca_part1, metrics = c("aucinf.obs", "cmax"))
```

  2. Two-stage: a `doseprop_stats` object (computed separately with `df_doseprop()`
  
```{r plot-doseprop-2step, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
plot_doseprop(tab)
```

See the [Dose-Proportionality Workflow](doseprop-workflow.html) article for confidence-interval interpretation, multi-metric layouts, and theme customization.


# Model diagnostics with `plot_gof()`

`plot_gof()` produces a population overlay goodness-of-fit (GOF) plot depicting binned central tendency layers of observed values (`dv`), individual predictions (`ipred`), and population predictions (`pred`) over time along with underlying observed data scatter layer (`obs`).

In most cases, these plots will be drawn using model output tables directly from estimation engines (e.g., NONMEM), which will only include predictions at timepoints non-missing and included in parameter estimation (e.g., MDV=0). Our input dataset `data_gof` (derived from `data_sad_pkfit`) includes model predictions from `mrgsim` at all timepoints; therefore, we will fitler on input to mimic this common scenario.

```{r plot-gof, fig.width=8, fig.height=5, warning=FALSE, message=FALSE}
plot_gof(data = filter(data_gof, MDV == 0), dv_var = ODV, log_y = TRUE) +
  facet_wrap(~Regimen) +
  scale_x_continuous(limits = c(0, 168), breaks = c(0, 24, 72, 120, 168))+
  labs(y = "Concentration (ng/mL)", x = "Time (hours)")
```

See the [Goodness-of-Fit Diagnostics](gof-diagnostics.html) article for specifying central tendency, BLQ handling, and theme customization.

# VPC model evaluation with `plot_vpc_cont()` and `plot_vpc_cens()`

The VPC process has been fully built as a workflow within `pmxhelper`, starting from the a fitted model that has been translated to an `mrgsolve` model format (`mrgmod`). 

From a validated model file, the VPC pipeline includes:
    + replication of the input dataset via simulation with `df_mrgsim_replicat()`
    + derivation of within and across trial replicate summary statistics with `df_vpcstats`
    + plot building with `plot_build_vpc()` (both continuous and censored `type'`available)


## Run the simulation with `df_mrgsim_replicat()`

 `df_mrgsim_replicate()` is a wrapper function for `mrgsim_df()`, which uses `lapply()` (or `future.apply::future_lapply()` when `parallel = TRUE` and a corresponding `future::plan()` in place) to iterate the simulation over integers from 1 to the value passed to the argument `replicates`.

There are 3 required arguments to `df_mrgsim_replicate()`

  - `data`, a `data.frame` modeling analysis dataset
  - `model`, a `mrgmod` model object
  - `replicates`, numeric number of replicates to perform.
  
The are optional arguments specifying key dataset variables to be input into the simulation or captured in output. These include: 
  - `dv_var` = DV, dependent variable
  - `time_var` = TIME, actual time variable
  - `ntime_var` = NTIME, nominal time variable
  - `pred_var` = PRED, population prediction variable (fixed effects only)
  - `ipred_var` = IPRED, individual prediction variable (fixed + level 1 random effects)
  - `sim_dv_var` = DV, dependent variable captured in the simulated output (fixed + level 1 and 2 random effects)

```{r vpc-simulate, message=FALSE, warning=FALSE}
simout <- df_mrgsim_replicate(
  data = data_pk,
  model = pkmodel,
  replicates = 100,
  dv_var = "ODV",
  sim_dv_var = DV,
  carry_out = c("DOSE", "FOOD", "BLQ", "LLOQ"),
  recover  = c("PART", "Regimen")
) 
glimpse(simout)
```

## Calculate summary statistics with `df_vpcstats()`

`df_vpcstats` performs the input data validation and summary statistic calculations for the VPC, returning a `pmx_stats`, `vpc_stats` S3 container including `$stats` (`data.frame` of summary statistics), `$obs` observed data for scatter plot overlay, `$config` configuration informatio (e.g., replicates, loq, stratifying variable).

`vpc_stats` objects contain both standard, prediction-correction, and proportion BLQ statistics and may be passed directly to `plot_vpc_cont()` or `plot_vpc_cens()` following a 2-stage workflow. Both plotting functions can also take in the raw simulated output and call `df_vpcstats()` internally for a one-stage workflow; however, the summary calculation computation cost is paid in every plot.

```{r vpc-stats}
vpcstats_obj_part <- df_vpcstats(simout, strat_var = PART)
vpcstats_obj_regimen <- df_vpcstats(simout, strat_var = Regimen)

vpcstats_obj_part
```

## Plotting the continuous data range with `plot_vpc_cont`

`plot_vpc_cont()` builds standard or prediction-corrected VPCs of the continuous, quantifiable range above the LLOQ. 

The example below is a pcVPC stratified by PART. 

```{r plot-vpc-cont-std, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
vpc_regimen <- plot_vpc_cont(
  data = simout,
  strat_var = Regimen
) +
  scale_x_continuous(breaks = seq(0, 168, 24)) +
  scale_y_log10(guide = "axis_logticks") +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)")
vpc_regimen
```

The example below is a pcVPC stratified by PART. 

```{r plot-vpc-cont-predcorr, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
pcvpc_part <- plot_vpc_cont(
  data = simout,
  strat_var = PART,
  pcvpc = TRUE
) +
  scale_x_continuous(breaks = seq(0, 168, 24)) +
  scale_y_log10(guide = "axis_logticks") +
  labs(x = "Time (hours)", y = "Pred-corrected Conc. (ng/mL)")
pcvpc_part
```
## Plotting the censored data range with `plot_vpc_cens`

`plot_vpc_cens()` is the companion diagnostic for the censored portion of the range — the per-bin BLQ proportion. It mirrors `plot_vpc_cont()` but plots `obs_prop_blq` and the empirical confidence band of `sim_prop_blq` across replicates, and requires a LOQ source.

The most relevant censored VPC is that stratified by dose and food status, which are combined in the `Regimen` variable.

```{r plot-vpc-cens-regimen, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
cens_vpc_regimen <- plot_vpc_cens(
  data = simout,
  strat_var = Regimen,
  loq = 1
) +
  scale_x_continuous(breaks = seq(0, 168, 24)) +
  labs(x = "Time (hours)", y = "Proportion BLQ")
cens_vpc_regimen
```

## Adding a panels and legends with `patchwork` and `plot_vpc_legend()`
`plot_vpc_legend()` returns a legend that can be patchworked beneath the VPC panel(s). The four row layout below stacks the continous and censored VPCs with legends. The two-stage plot building from `vpc_stats` objects is also demonstrated, highlighting the potential efficiency of reusing a precomputed object in such a workflow.

```{r plot-vpc-patchwork, fig.width=8, fig.height=7, warning=FALSE, message=FALSE}
vpc_plot_cont <- plot_vpc_cont(vpcstats_obj_regimen) + 
  scale_x_continuous(breaks = seq(0, 168, 24)) +
  scale_y_log10(guide = "axis_logticks") +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)")
vpc_plot_cens <- plot_vpc_cens(vpcstats_obj_regimen) +
  scale_x_continuous(breaks = seq(0, 168, 24)) +
  labs(x = "Time (hours)", y = "Proportion BLQ")
cont_legend <- plot_vpc_legend()
cens_legend <- plot_vpc_legend(type = "cens", shown = plot_vpc_shown(obs_point = FALSE))
vpc_plot_cont / cont_legend / vpc_plot_cens / cens_legend + 
  plot_layout(heights = c(2, 0.5, 1, 0.5))
```

See the [Visual Predictive Check Workflow](vpc-workflow.html) article for stratification controls, BLQ handling options, multi-LLOQ pooling, `shown`/`theme` customization, and pairing pcVPC with cens VPC.

# Theme system overview

Every `plot_*()` function has a paired `plot_*_theme()` factory that returns a named list of default element keys that follow a `datalayer_geom` pattern. Elements are built with typed constructors (`pmx_point()`, `pmx_line()`, `pmx_ribbon()`, `pmx_trend()`, `pmx_errorbar()`, `pmx_style()`, `pmx_color()`) and passed to keys within the factory function. The factory functions merges user-supplied overrides and override only the values passed leaves the rest at their default values.

```{r theme-defaults}
plot_dvtime_theme()

new_dvtime_theme <- plot_dvtime_theme(
    obs_point = pmx_point(alpha = 0),
    cent_line = pmx_line(linewidth = 1.5),
    cent_errorbar = pmx_errorbar(linewidth = 1.5)
  )
```

```{r plot-dvtime-themed, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
plot_dvtime(
  data = data_pk,
  dv_var = ODV,
  cent = "mean_sdl",
  col_var = "Regimen",
  log_y = TRUE,
  theme = new_dvtime_theme
) +
  labs(y = "Concentration (ng/mL)", x = "Time (hours)")
```

See the [Plot Themes and Aesthetics](plot-themes.html) article for the full theme catalog, element-constructor reference, and the class system that backs them.

# Where to go next

- [Exploratory Analyses of PK and PK/PD Data](eda-pk-pkpd-workflow.html)
- [Dose-Proportionality Workflow](doseprop-workflow.html)
- [Goodness-of-Fit Diagnostics](gof-diagnostics.html)
- [Visual Predictive Check Workflow](vpc-workflow.html)
- [Plot Themes and Aesthetics](plot-themes.html)
