---
title: "Getting Started with pmxhelpr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pmxhelpr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

This vignette is a lean tour of every core `pmxhelpr` workflow in a single 
sitting. Each section ends with a link to the corresponding deep-dive article
on the [pmxhelpr website](https://ryancrass.github.io/pmxhelpr/).

Exported functions follow a `ReturnType_Purpose` naming convention:

- `plot_*` — returns a `ggplot` object
- `df_*` — returns a `data.frame` (often class-tagged)
- `var_*` — returns a vector (vectorized helpers for use inside `mutate()`)
- `style_*` — returns a `ggstylekit::style_spec()` theme object

The bundled datasets `data_sad` and `data_sad_pkfit` use `ODV` (original DV). 
Examples below pass `dv_var = "ODV"` or `dv_var = ODV` to highlight the 
non-standard evaluation (NSE) of dataset column name variables which may be 
specified as strings or bare names to override defaults in `pmxhelpr` functions. 

```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts = FALSE)
library(forcats, warn.conflicts = FALSE)
library(ggplot2, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
library(mrgsolve, warn.conflicts = FALSE)
```

# Data

## `data_sad`

`data_sad` is a NLME modeling analysis-ready dataset for single ascending dose
(SAD) study with a parallel food-effect (FE) cohort. It contains one oral dose
input (`EVID=1`, `CMT=1`) and two observation types (`EVID=0`): 

  + Drug concentration in `CMT=2` (original units: ng/mL)
  + Response in `CMT=3` (original units: percentage of baseline)

```{r data-sad}
glimpse(data_sad)
```

The pre-processing step below creates a new variable combining dose, frequency,
and food status into a single dosing regimen variable with per-group subject 
counts appended using `var_addn()`. Separate datasets are also defined for 
exploration of PK and PD data.

```{r data-addn}
data <- data_sad %>%
  mutate(Food = ifelse(FOOD == 1, "Fed", "Fasted"),
         DoseFood = paste(DOSE, "mg x1", Food),
         Regimen = var_addn(DoseFood, ID))

data_pk <- data %>% filter(CMT %in% c(1,2))
data_pd <- data %>% filter(CMT %in% c(1,3))

unique(data$Regimen)
```

## `data_sad_nca`

`data_sad_nca` contains PK parameters and exposure metrics from a 
non-compartmental analysis (NCA) of `data_sad` using the `PKNCA` package.

We will filter to Part 1 fasted conditions only for use in dose-proportionality assessment.

```{r data-nca}
glimpse(data_sad_nca)
data_nca_part1 <- filter(data_sad_nca, PART == "Part 1-SAD")
```

## `data_sad_pkfit`

`data_sad_pkfit` is a model output dataset version of `data_sad` (`CMT`=1, 2)
with two additional variables (`PRED` and `IPRED`) appended.  These variables 
are derived from the internal PK model `pkmodel`.

```{r vpc-model, message=FALSE, warning=FALSE}
pkmodel <- model_mread_load("pkmodel")
see(pkmodel)
```
We will process this dataset in a manner analogous to `data_sad`.

```{r data-gof}
data_gof <- data_sad_pkfit %>%
  mutate(Food = ifelse(FOOD == 1, "Fed", "Fasted"),
         DoseFood = paste(DOSE, "mg x1", Food),
         Regimen = var_addn(DoseFood, ID)) 
```


# Longitudinal concentration and response with `plot_dvtime()`

`plot_dvtime()` produces a longitudinal plot of a repeated measures, continuous
variable versus time. This plotting function can be used for longitudinal 
exploratory analysis of both concentration (`CMT = 2`) and response (`CMT = 3`)
over time, with functionality for central tendency, variability, and BLQ 
handling.

```{r plot-dvtime-pk, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
pk_plot <- plot_dvtime(
  data = data_pk,
  dv_var = ODV,
  cent = "mean_sdl",
  col_var = Regimen,
  log_y = TRUE,
  style = style_dvtime(alphas = c(obs_point = 0))
) +
  labs(y = "Concentration (ng/mL)", x = "Time (hours)")
pk_plot
```

```{r plot-dvtime-pd, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
pd_plot <- plot_dvtime(
  data = data_pd,
  dv_var = "CFB",
  cent = "mean_sdl",
  col_var = Regimen,
  style = style_dvtime(alphas = c(obs_point = 0))
) +
  labs(y = "Response (% Change from Baseline)", x = "Time (hours)")
pd_plot
```
The two plots can be composed into a single paneled figure using the 
`patchwork` package, aligned vertically with the shared time axis.

```{r plot-dvtime-patchwork, fig.width=8, fig.height=7, warning=FALSE, message=FALSE}
(pk_plot / pd_plot) + plot_layout(guides = "collect")
```

See the [Exploratory Analyses of PK and PK/PD Data](eda-pk-pkpd-workflow.html)
article for additional details on exploratory data analyses with `pmxhelpr`.

# Response versus concentration with `plot_dvconc()`

`plot_dvconc()` plots a continuous dependent variable versus a continuous
independent variable. The most common use case is visualizing a pharmacodynamic
response versus drug concentration. Trend line options include both linear
(`linear`/`se_linear`) and non-linear (`loess`/`se_loess`) logical toggles.

A dashed black reference line is drawn at `y = ref` when `ref` is specified.

```{r plot-dvconc, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
plot_dvconc(
  data = filter(data, CMT == 3),
  dv_var = CFB,
  idv_var = CONC,
  ref = 0,
  col_var = Regimen,
  loess = TRUE,
  linear = TRUE
) +
  labs(y = "Response (% Change from Baseline)", x = "Concentration (ng/mL)")
```

By default the trend lines are not grouped by the variable passed to `col_var`;
however, this can be toggled on with `col_trend=TRUE`.

```{r plot-dvconc-coltrend, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
plot_dvconc(
  data = filter(data, CMT == 3),
  dv_var = CFB,
  idv_var = CONC,
  ref = 0,
  col_var = Regimen,
  loess = TRUE,
  se_loess = TRUE,
  linear = FALSE,
  col_trend = TRUE
) +
  labs(y = "Response (% Change from Baseline)", x = "Concentration (ng/mL)")
```

See the [Exploratory Analyses of PK and PK/PD Data](eda-pk-pkpd-workflow.html) 
article for additional details on exploratory data analyses with `pmxhelpr`.

# Dose-proportionality with `df_doseprop()` and `plot_doseprop()`

`df_doseprop()` fits a log-log regression of exposure metrics (e.g., Cmax and 
AUC) versus dose and returns a class-tagged `doseprop_stats` data frame.

```{r doseprop-stats}
tab <- df_doseprop(data_nca_part1, metrics = c("aucinf.obs", "cmax"))
tab
```

`plot_doseprop()` can directly process an NCA input dataset (1-stage) or accept
a previously processed `doseprop_stats` object (2-stage)

  1. One-stage: raw NCA `data.frame` input (`df_doseprop()` called internally)
  
```{r plot-doseprop-1step, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
plot_doseprop(data_nca_part1, metrics = c("aucinf.obs", "cmax"))
```

  2. Two-stage: a `doseprop_stats` object (computed separately with `df_doseprop()`
  
```{r plot-doseprop-2step, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
plot_doseprop(tab)
```

See the [Dose-Proportionality Workflow](doseprop-workflow.html) article for
additional details on dose-proportionality assessment with `pmxhelpr`.

# Model diagnostics with `plot_gof()`

`plot_gof()` produces a population overlay goodness-of-fit (GOF) plot depicting
binned central tendency layers of observed values (`dv`), individual predictions
(`ipred`), and population predictions (`pred`) over time along with the 
underlying observed data (`obs`) scatter.

In most cases, these plots will be drawn using model output tables directly 
from estimation engines (e.g., NONMEM), which will only include predictions at 
non-missing timepoints included in parameter estimation (e.g., MDV=0). The input
dataset `data_gof` (derived from `data_sad_pkfit`) includes model predictions 
from `mrgsolve::mrgsim()` at all timepoints; therefore, we will filter on input
to mimic this common scenario.

```{r plot-gof, fig.width=8, fig.height=5, warning=FALSE, message=FALSE}
plot_gof(data = filter(data_gof, MDV == 0), dv_var = ODV, log_y = TRUE) +
  facet_wrap(~Regimen) +
  scale_x_continuous(limits = c(0, 168), breaks = c(0, 24, 72, 120, 168))+
  labs(y = "Concentration (ng/mL)", x = "Time (hours)")
```

See the [Goodness-of-Fit Diagnostics](gof-diagnostics.html) article for 
additional details on GOF diagnostics.

# VPC model evaluation with `plot_vpc_cont()` and `plot_vpc_cens()`

The VPC workflow is an end to end process within `pmxhelpr`, starting from a 
fitted model that has been translated to an `mrgsolve` model format (`mrgmod`).

From a validated model file, the VPC pipeline includes functionality to:

  1. replicate the input dataset with `df_mrgsim_replicate()`
  2. derive summary statistics within and across replicates with `df_vpcstats()`
  3. build plots with `plot_vpc_cont()` / `plot_vpc_cens()`

## Run the simulation with `df_mrgsim_replicate()`

`df_mrgsim_replicate()` is a wrapper function for `mrgsim::mrgsim_df()`, 
which uses `lapply()` to iterate from 1 to the integer value passed to 
`replicates`. This can be parallelized using `future.apply::future_lapply()` 
when `parallel = TRUE` and a corresponding `future::plan()` is in place.

There are 3 required arguments to `df_mrgsim_replicate()`:

  - `data`, a `data.frame` modeling analysis dataset
  - `model`, a `mrgmod` model object
  - `replicates`, integer number of replicates to perform.
  
The are optional arguments specifying key dataset variables to be input into 
the simulation or captured in output. These include: 

  - `dv_var` = DV, dependent variable
  - `time_var` = TIME, actual time variable
  - `ntime_var` = NTIME, nominal time variable
  - `pred_var` = PRED, population prediction variable
  - `ipred_var` = IPRED, individual prediction variable 
  - `sim_dv_var` = DV, dependent variable captured in the simulated output

```{r vpc-simulate, message=FALSE, warning=FALSE}
simout <- df_mrgsim_replicate(
  data = data_pk,
  model = pkmodel,
  replicates = 100,
  dv_var = "ODV",
  sim_dv_var = DV,
  carry_out = c("DOSE", "FOOD", "BLQ", "LLOQ"),
  recover  = c("PART", "Regimen")
) 
glimpse(simout)
```

## Calculate summary statistics with `df_vpcstats()`

`df_vpcstats` performs the input data validation and summary statistic 
calculations for the VPC, returning a `vpc_stats` S3 container including:

 + `$stats` (`data.frame` of summary statistics)
 + `$obs` observed data for scatter plot overlay
 + `$config` configuration info (e.g., replicates, loq, stratifying variable).

`vpc_stats` objects contain both standard, prediction-correction, and proportion
BLQ statistics and may be passed directly to `plot_vpc_cont()` or 
`plot_vpc_cens()` following a 2-stage workflow. Both plotting functions can 
also take in the raw simulated output and call `df_vpcstats()` internally in a
one-stage workflow.

```{r vpc-stats}
vpcstats_obj_part <- df_vpcstats(simout, strat_var = PART)
vpcstats_obj_regimen <- df_vpcstats(simout, strat_var = Regimen)

vpcstats_obj_part
```

## Plotting the continuous data range with `plot_vpc_cont`

`plot_vpc_cont()` builds standard or prediction-corrected VPCs of the 
continuous, quantifiable range above the LLOQ.

The example below is a standard (non-prediction-corrected) VPC stratified by
`Regimen`.

```{r plot-vpc-cont-std, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
vpc_regimen <- plot_vpc_cont(
  data = simout,
  strat_var = Regimen
) +
  scale_x_continuous(breaks = seq(0, 168, 24)) +
  scale_y_log10(guide = "axis_logticks") +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)")
vpc_regimen
```

The example below is a pcVPC stratified by PART. 

```{r plot-vpc-cont-predcorr, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
pcvpc_part <- plot_vpc_cont(
  data = simout,
  strat_var = PART,
  pcvpc = TRUE
) +
  scale_x_continuous(breaks = seq(0, 168, 24)) +
  scale_y_log10(guide = "axis_logticks") +
  labs(x = "Time (hours)", y = "Pred-corrected Conc. (ng/mL)")
pcvpc_part
```
## Plotting the censored data range with `plot_vpc_cens`

`plot_vpc_cens()` is the companion diagnostic for the censored portion of the 
data range. It mirrors `plot_vpc_cont()` but plots `obs_prop_blq` and the 
non-parametric confidence band of `sim_prop_blq` across replicates. VPC plots 
of the censored data range should be generated prior to prediction-corrected 
VPCs (pcVPCs) to ensure that the model is adequately predictive of the censored
data range prior to generating pcVPCs of the quantifiable data range.

The most relevant censored VPC is that stratified by dose and food status, 
which are combined in the `Regimen` variable.

```{r plot-vpc-cens-regimen, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
cens_vpc_regimen <- plot_vpc_cens(
  data = simout,
  strat_var = Regimen,
  loq = 1
) +
  scale_x_continuous(breaks = seq(0, 168, 24)) +
  labs(x = "Time (hours)", y = "Proportion BLQ")
cens_vpc_regimen
```

## Adding a panels and legends with `patchwork` and `plot_vpc_legend()`

`plot_vpc_legend()` returns a legend that can be combined with the VPC plots
using the `patwork` package. The four row layout below stacks the continuous 
and censored VPCs with their corresponding legends. 

The two-stage plot building from a`vpc_stats` objects is also demonstrated, 
highlighting the potential efficiencies of reusing a pre-computed object.

```{r plot-vpc-patchwork, fig.width=8, fig.height=7, warning=FALSE, message=FALSE}
vpc_plot_cont <- plot_vpc_cont(vpcstats_obj_regimen) + 
  scale_x_continuous(breaks = seq(0, 168, 24)) +
  scale_y_log10(guide = "axis_logticks") +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)")

vpc_plot_cens <- plot_vpc_cens(vpcstats_obj_regimen) +
  scale_x_continuous(breaks = seq(0, 168, 24)) +
  labs(x = "Time (hours)", y = "Proportion BLQ")

cont_legend <- plot_vpc_legend()
cens_legend <- plot_vpc_legend(type = "cens", shown = plot_vpc_shown(obs_point = FALSE))

vpc_plot_cont / cont_legend / vpc_plot_cens / cens_legend + 
  plot_layout(heights = c(2, 0.5, 1, 0.5))
```

See the [Visual Predictive Check Workflow](vpc-workflow.html) article for 
additional details on the VPC toolchain with `pmxhelpr`.

# Plot themes with `ggstylekit`

`pmxhelpr` leverages the functionality of the `ggstylekit` package to control
the styling of plot elements.

Every `plot_*()` function ships with a compaion `style_*()` helper function
that returns a `ggstylekit::style_spec()` object pre-filled with the defaults. 
The defaults styling for each plot type can be visualized by calling the 
function with no arguments

```{r style-defaults}
style_dvtime()
```

Styling is keyed by  *roles* (e.g. `obs_point`, `cent_line`, `cent_errorbar`,
`ref_line`, `loq_line`). The *values* for each key are individual plot 
aesthetics (e.g, `colors`, `shapes`, `sizes`, `linetypes`, `linewidths`, 
`alphas`) with partial overrides merge onto the defaults so setting one role 
leaves the others untouched.

New styles can be defined as an object and recycled across plots. This revised
style object removes the observed points and increases the linewidth of the 
central tendency line and error bars.

```{r style-new}
new_dvtime_style <- style_dvtime(
    alphas     = c(obs_point = 0),
    linewidths = c(cent_line = 1.5, cent_errorbar = 1.5)
  )
new_dvtime_style
```
The 

```{r plot-dvtime-styled, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
plot_dvtime(
  data = data_pk,
  dv_var = ODV,
  cent = "mean_sdl",
  col_var = "Regimen",
  log_y = TRUE,
  style = new_dvtime_style
) +
  labs(y = "Concentration (ng/mL)", x = "Time (hours)")
```

A finished plot object can also be restyled after the fact with
`restyle_plot()`. Additionally, variables can be surfaced and mapped to
aesthetics after the fact with `reveal()`. Both are re-exported from ggstylekit.

See the [Plot Themes and Aesthetics](plot-themes.html)  article for additional 
details on plot themeing with `pmxhelpr`.

# Where to go next

- [Exploratory Analyses of PK and PK/PD Data](eda-pk-pkpd-workflow.html)
- [Dose-Proportionality Workflow](doseprop-workflow.html)
- [Goodness-of-Fit Diagnostics](gof-diagnostics.html)
- [Visual Predictive Check Workflow](vpc-workflow.html)
- [Plot Themes and Aesthetics](plot-themes.html)
