---
title: "Standard Overlay Goodness-of-Fit Diagnostics Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gof-diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will demonstrate `pmxhelpr` functions for generating standard overlay goodness-of-fit model diagnostics for evaluation of longitudinal non-linear mixed effects model fit.

This vignette will assume familiarity with the `data_sad` internal dataset and the `plot_dvtime` function, as described in the Exploratory Analyses of PK Data vignette. These elements will not be reviewed in detail in this vignette.

First, we will load the required packages.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(forcats, warn.conflicts = FALSE)
library(Hmisc, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
```


# PK Model

An example PK model (`pkmodel`) in `mrgmod` format is provided in the internal package library. This can be loaded using the helper function `model_mread_load()`, which wraps `mrgsolve::mread`.

```{r model-pk}
model <- model_mread_load("pkmodel")
mrgsolve::see(model)
```

# Data 

The example dataset used in this vignette (`data_sad`) is based on a single ascending dose (SAD) study of an orally administered drug product with a parallel group food effect (FE) cohort. 

## data_sad_pkfit

`data_sad_pkfit` is a model output dataset version of `data_sad` containing model predictions from `pkmodel` appended to the observed data. We can take a quick look at the dataset using `glimpse()` from the dplyr package. Dataset definitions can also be viewed by calling `?data_sad_pkfit`, as one would to view the documentation for a package function.

```{r data-glimpse}
glimpse(data_sad_pkfit)
```

This dataset contains two additional variables representing model predictions:

+ `PRED`: population model predicted values accounting only for fixed effects (THETAs)
+ `IPRED`: individual model predicted values accounting for fixed effects (THETAs) and level 1 (inter-individual) random effects (ETAs)

Let's derive some additional variables and leverage the functionality of `var_addn` to create a new factor variable including a count of unique individuals in each unique dosing condition for plotting

```{r data-plotdata}
plot_data <- data_sad_pkfit %>% 
  filter(EVID == 0) %>% 
  mutate(`Food Status` = ifelse(FOOD == 0, "Fasted", "Fed"), 
         `Dose and Food` = paste(DOSE, "mg", `Food Status`)) %>% 
  mutate(DoseFood = `Dose and Food`,
         `Food Status` = var_addn(`Food Status`, ID),
         `Dose and Food` = var_addn(`Dose and Food`, ID)) %>% 
  mutate(`Dose and Food` = fct_relevel(`Dose and Food`, "50 mg Fasted (n=6)", after = 1))
```


# Population Overlay Goodness of Fit Plots with `plot_gof`

## Overview

 `pmxhelpr` includes a function for generating population overlay goodness-of-fit (GOF) plots for model evaluation: `plot_gof`.
 
 *NOTE* `plot_gof` inherits dataset variable handling functionality and arguments from `df_mrgsim_replicate` and ports aesthetic controls from `plot_dvtime`.
  
## Specifying Dependent and Independent Variables 

`plot_gof` has 3 arguments that specify the dependent variables to be mapped to the y-axis (`dv_var`, `ipred_var`, `pred_var`) and 2 arguments for the independent time variables to be mapped to the x-axis (`time_var`, `ntime_var`). `ntime_var` is an exact binned version of the x-axis variable for calculation of central tendency statistics. 

The defaults are as follows:
-  `dv_var`: Dependent variable. Default is `DV`.
-  `ipred_var`: Individual predictions. Default is `IPRED`.
-  `pred_var`: Population predictions. Default is `PRED`.
-  `time_var`: Actual time variable. Default is `TIME`.
-  `ntime_var`: Nominal time variable (must be an exact binning variable). Default is `NTIME`.

These arguments accept non-standard evaluation and may be supplied as bare names or strings.

The example dataset `data_sad_pkfit` only differs from these defaults in the variable name for the dependent variable, `"ODV"`. Thus, the most basic population GOF plot can be obtained with:

```{r plot-popgof, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_gof(data = plot_data, dv_var = "ODV") + 
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)")
```

## Applying Dose-normalization

The pooled analysis population includes multiple dose levels in a single plot, which is probably not optimal as a population PK model diagnostic. A better minimal plot representation of these data can be obtained by dose-normalizing and stratifying by study part to separate out the fast and food effect portions:

```{r plot-popgof-dn, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_gof(data = plot_data, dv_var = "ODV", dosenorm = TRUE) +
  facet_wrap(~`Food Status`) +
  labs(y = "Dose-normalized Conc. (ng/mL/mg)", x = "Time Since First Dose (hours)")
```

## Facetting by Study Design (Extrinsic) Factors

Although, generally population goodness-of-fit plots are stratified by study design (extrinsic) factors (e.g., Dose and Food Status) in order to assess the adequacy of model fit in each unique study condition.

```{r plot-popgof-dosegrp, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_gof(data = plot_data, dv_var = "ODV") +
  facet_wrap(~`Dose and Food`, scales = "free") +
  labs(y = "Concentration (ng/mL)")
```

## Specifying Plot Elements and Assigning Colors

There is *ONLY* one new argument to `plot_gof` that is not present in `plot_dvtime`: `output_colors`. This argument is a named vector of pre-defined plot element and color pairs:

  - `PRED= "red"`
  - `IPRED`=`"green"`,
  - `DV`=`"blue"`.
  - `OBS`=`"darkgrey"`

The default plot theme elements are:

 - `pred` - summary statistics for central tendency in the population model predictions (`pred_var`)
 - `ipred` = summary statistics for central tendency in the individual model predictions (`ipred_var`)
 - `obs` = observed data points for the dependent variable (`dv_var`)
 - `dv` = summary statistics for central tendency (+/- variability) in the dependent variable (`dv_var`)

`obs` and `dv` represent the same data and both map back to the variable in `data` specified in `dv_var`. They are differentiated only to allow visual distinction between individual observation concentration data and summary statistics of observed and model-predicted concentration data.

If one wanted to clarify that all the data mapping back to the dataset variable specified in `dv_var` with color, then the default for `obs` color can be updated to blue to align with `dv` via the `theme` argument.

```{r plot-popgof-dosegrp-blue, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_gof(data = plot_data, dv_var = "ODV", theme = list(obs = pmx_style(color = "blue"))) +
  facet_wrap(~`Dose and Food`, scales = "free") +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)")
```

Alternatively, the observed data points could be removed with only the summary metrics plotted. Notice how different this plot appears when only visualizing central tendency!

```{r plot-popgof-dosegrp-noobs, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_gof(data = plot_data, dv_var = "ODV", obs_dv = FALSE) +
  facet_wrap(~`Dose and Food`, scales = "free") +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)")
```


## Specifying the Central Tendency

`plot_gof` inherits its central tendency handling functionality from `plot_dvtime`. Both functions use the `stat_summary` function from `ggplot2` to calculate and plot the central tendency measures and error bars. The summary statistics calculated are specified by the `cent` argument. 
  
In `plot_gof`, variability (SD, IQR) is plotted for the observed data (DV) and only the central tendency measures are calculated and returned for model predictions (PRED, IPRED).

An often overlooked feature of `stat_summary`, is that it calculates the summary statistics *after* any transformations to the data performed by changing the scales. This means that when `scale_y_log10()` is applied to the plot, the data are log-transformed for plotting and the central tendency measure returned with `"mean"` is the *geometric mean*. 

If the `log_y` argument is used to generate semi-log plots along with `show_captions = TRUE`, then the caption will delineate where arithmetic and geometric means are being returned. 

```{r plot-popgof-dosegrp-meansd, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_gof(data = plot_data, dv_var = "ODV", cent = "mean_sdl",
            log_y = TRUE) +
  facet_wrap(~`Dose and Food`, scales = "free") +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)")
```

The caption will *NOT* update if a new axis is added to the plot object outside of `plot_gof` with `scale_y_log10`.

```{r plot-popgof-dosegrp-logscale, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_gof(data = plot_data, dv_var = "ODV", cent = "mean_sdl") +
  facet_wrap(~`Dose and Food`, scales = "free") +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") + 
  scale_y_log10()
```

Transformation of the plot to a semi-log scale (log-scale y-axis only) is recommended to be performed using the `log_y` argument for the following benefits:

  + Includes log tick marks on the y-axis 
  + Updates the caption with the correct central tendency measure if `show_captions = TRUE`. 
  


# Individual Concentration-time plots

The previous section provides an overview of how to generate population overlay GOF plots using `plot_gof`; however, we can also use `plot_gof` to generate subject-level visualizations with a little pre-processing of the input dataset. 

We can plot an individual subject by filtering the input dataset. This could be extended to generate plots for all individuals using `for` loops, `lapply`, `purrr::map()` functions, or other methods.

```{r plot-indgofplots, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}

ids <- sort(unique(plot_data$ID)) #vector of unique subject ids
n_ids <- length(ids) #count of unique subject ids
plots_per_pg <- 4
n_pgs <- ceiling(n_ids/plots_per_pg) #Total number of pages needed

plist<- list()
for(i in 1:n_ids){
  plist[[i]] <- plot_gof(filter(plot_data, ID == ids[i]),
                            dv_var = "ODV",
                            log_y = TRUE,
                            grp_dv = TRUE,
                            show_caption = FALSE) +
  facet_wrap(~DoseFood) +
  labs(title = paste0("ID = ", ids[i]), y = "Concentration (ng/mL)", x = "Time (hours)") +
  theme(legend.position="none")
}

lapply(1:n_pgs, function(n_pg) {
      i <-  (n_pg-1)*plots_per_pg+1
      j <- n_pg*plots_per_pg
      wrap_plots(plist[i:j])
})
```

