---
title: "Plot Themes and Aesthetics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{plot-aesthetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(forcats, warn.conflicts = FALSE)
library(Hmisc, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
```

# Overview

Every `pmxhelpr` plot function (`plot_dvtime`, `plot_dvconc`, `plot_gof`, `plot_vpc_cont`) accepts a `theme` argument that controls the appearance of points, lines, ribbons, error bars, and reference lines. The theme system has two layers:

1. **Element constructors** (`pmx_point`, `pmx_line`, etc.) define *what* to style. Each constructor maps directly to a `ggplot2` geom and only accepts aesthetics relevant to that geom.

2. **Theme factories** (`plot_dvtime_theme`, `plot_gof_theme`, etc.) define *where* those elements go. Each factory returns a named list of default element objects, one per visual layer in the plot. User overrides are merged field-by-field into these defaults.

This design means every plot function uses the same workflow:

```r
# View defaults
plot_dvtime_theme()

# Override specific fields — unspecified fields keep their defaults
plot_dvtime(data, theme = plot_dvtime_theme(ref = pmx_line(linetype = 3)))
```


# Element Constructors

Each constructor creates a named list with a class tag. Only non-NULL fields are stored, so partial overrides are lightweight:

```{r constructor-example}
# Full element with all fields
pmx_point(shape = 16, size = 1.25, alpha = 1, color = "blue")

# Partial override — only the fields you want to change
pmx_point(size = 2)
```

When a user-supplied element is merged into a default, each specified field replaces the corresponding default. Unspecified fields are preserved:

```{r merge-example}
default <- pmx_point(shape = 1, size = 0.75, alpha = 0.5)
override <- pmx_point(size = 2)
pmxhelpr:::merge_element(override, default)
```

The available constructors are:

| Constructor | Controls | Fields |
|---|---|---|
| `pmx_point()` | Point layers (`geom_point`, `stat_summary`) | `shape`, `size`, `alpha`, `color` |
| `pmx_line()` | Line layers (`geom_line`, `geom_hline`) | `linewidth`, `linetype`, `alpha`, `color` |
| `pmx_ribbon()` | Ribbon layers (`geom_ribbon`) | `fill`, `alpha`, `color`, `linetype`, `linewidth` |
| `pmx_errorbar()` | Error bar layers (`geom_errorbar`) | `linewidth`, `linetype`, `alpha`, `width` |
| `pmx_trend()` | Trend lines (`geom_smooth`) | `linewidth`, `linetype`, `color`, `se_color`, `se_alpha` |
| `pmx_style()` | Convenience shortcut (see below) | `color`, `alpha` |


# Theme Factories

## Naming Convention

Theme keys follow a `role_element` pattern where the **role** identifies the data layer and the **element** identifies the geom type:

- `obs_point` — observed data, point geom
- `cent_line` — central tendency, line geom
- `pred_point` — PRED predictions, point geom

When a role uses only one geom type, the key is just the role name (e.g., `ref`, `obs`, `loess`).

## Shared Elements Across Factories

Several element roles appear across multiple theme factories:

| Element | Type | Used in |
|---|---|---|
| `ref` | `pmx_line` | `plot_dvtime_theme`, `plot_gof_theme`, `plot_dvconc_theme` |
| `errorbar` | `pmx_errorbar` | `plot_dvtime_theme`, `plot_gof_theme` |
| `obs` / `obs_point` | `pmx_point` | all four theme factories |

The same constructors and merge logic apply everywhere, so learning the system once covers all plot types.

## Role-Level Shortcuts with `pmx_style()`

When a role has both `_point` and `_line` elements (e.g., `obs_point` and `obs_line`), you can set shared aesthetics on both at once using `pmx_style()`:

```{r style-shortcut}
# These two calls produce the same result:
plot_dvtime_theme(obs_point = pmx_point(alpha = 0.3),
                  obs_line = pmx_line(alpha = 0.3))

plot_dvtime_theme(obs = pmx_style(alpha = 0.3))
```

`pmx_style()` accepts `color` and `alpha` — the two fields shared by both `pmx_point` and `pmx_line`.

When both a style shortcut and an explicit element override are provided, the style is applied first, then the element override wins for any conflicting fields:

```{r style-precedence}
# Style sets color to purple on both point and line,
# then the explicit override changes just the point size
theme <- plot_gof_theme(
  pred = pmx_style(color = "purple"),
  pred_point = pmx_point(size = 3)
)
theme$pred_point  # color = "purple", size = 3
theme$pred_line   # color = "purple", linewidth = 0.75 (default)
```


## Per-Function Defaults

### `plot_dvtime_theme()`

```{r dvtime-defaults}
plot_dvtime_theme()
```

| Key | Type | Role |
|---|---|---|
| `obs_point` | `pmx_point` | Observed data scatter |
| `obs_line` | `pmx_line` | Spaghetti lines connecting observations |
| `cent_point` | `pmx_point` | Central tendency points |
| `cent_line` | `pmx_line` | Central tendency lines |
| `errorbar` | `pmx_errorbar` | Error bars for central tendency |
| `ref` | `pmx_line` | Reference lines (LLOQ, change-from-baseline) |

Shortcuts: `obs`, `cent` accept `pmx_style()`.

### `plot_gof_theme()`

```{r gof-defaults}
plot_gof_theme()
```

| Key | Type | Role |
|---|---|---|
| `obs_point`, `obs_line` | `pmx_point`, `pmx_line` | Observed data (default: darkgrey) |
| `dv_point`, `dv_line` | `pmx_point`, `pmx_line` | DV central tendency (default: blue) |
| `pred_point`, `pred_line` | `pmx_point`, `pmx_line` | PRED central tendency (default: red) |
| `ipred_point`, `ipred_line` | `pmx_point`, `pmx_line` | IPRED central tendency (default: green) |
| `errorbar` | `pmx_errorbar` | Error bars (shared) |
| `ref` | `pmx_line` | Reference lines |

Shortcuts: `obs`, `dv`, `pred`, `ipred` accept `pmx_style()`.

`obs` and `dv` represent the same underlying data (`dv_var`). They are differentiated to allow visual distinction between individual observations and summary statistics.

### `plot_dvconc_theme()`

```{r dvconc-defaults}
plot_dvconc_theme()
```

| Key | Type | Role |
|---|---|---|
| `obs` | `pmx_point` | Observed data points |
| `ref` | `pmx_line` | Reference line |
| `loess` | `pmx_trend` | LOESS trend line |
| `linear` | `pmx_trend` | Linear trend line |

### `plot_vpc_theme()`

```{r vpc-defaults}
plot_vpc_theme()
```

| Key | Type | Role |
|---|---|---|
| `obs` | `pmx_point` | Observed data points |
| `obs_median` | `pmx_line` | Observed median line |
| `obs_ci` | `pmx_line` | Observed CI lines |
| `sim_pi` | `pmx_ribbon` | Simulated prediction interval ribbons |
| `sim_median` | `pmx_ribbon` | Simulated median ribbon |
| `loq` | `pmx_line` | LOQ reference line |
| `bin_sep` | `pmx_line` | Bin separator line |


# Examples

The following examples demonstrate common customization patterns. Each pattern applies to any plot function — only the theme factory and element keys change.

```{r example-data, message=FALSE, warning=FALSE}
plot_data <- data_sad %>%
  filter(CMT == 2) %>%
  mutate(`Dose and Food` = var_addn(
    paste(DOSE, "mg", ifelse(FOOD == 0, "Fasted", "Fed")), ID)) %>%
  mutate(`Dose and Food` = fct_relevel(`Dose and Food`,
                                        "50 mg Fasted (n=6)", after = 1))

gof_data <- data_sad_pkfit %>%
  filter(EVID == 0) %>%
  mutate(`Dose and Food` = var_addn(
    paste(DOSE, "mg", ifelse(FOOD == 0, "Fasted", "Fed")), ID)) %>%
  mutate(`Dose and Food` = fct_relevel(`Dose and Food`,
                                        "50 mg Fasted (n=6)", after = 1))
```

## Inline Override

Pass element overrides directly in the `theme` argument. Only the fields you specify are changed:

```{r example-inline, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data, dv_var = ODV, col_var = `Dose and Food`,
            cent = "mean_sdl", log_y = TRUE, obs_dv = FALSE,
            theme = plot_dvtime_theme(errorbar = pmx_errorbar(linewidth = 0.5),
                                     cent_point = pmx_point(size = 0.1))) +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

## Reusable Theme Object

Create a theme once and pass it to multiple plots:

```{r example-reuse, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
my_theme <- plot_dvtime_theme(errorbar = pmx_errorbar(linewidth = 0.5),
                              cent_point = pmx_point(size = 0.1))

plot_dvtime(data = plot_data, dv_var = ODV, col_var = `Dose and Food`,
            cent = "mean_sdl", log_y = TRUE, obs_dv = FALSE,
            theme = my_theme) +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

## Color Customization with `pmx_style()`

Use `pmx_style()` to change the color of both point and line elements for a role in one call:

```{r example-style, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_gof(data = gof_data, dv_var = "ODV",
         theme = list(obs = pmx_style(color = "blue"))) +
  facet_wrap(~`Dose and Food`, scales = "free") +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)")
```

## Error Bar Width

The default error bar width is 2.5% of the maximum nominal time. Override with the `width` field:

```{r example-ebwidth, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data, dv_var = ODV, col_var = `Dose and Food`,
            cent = "mean_sdl", log_y = TRUE, obs_dv = FALSE,
            theme = plot_dvtime_theme(errorbar = pmx_errorbar(width = 8))) +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

## Custom VPC Color Scheme

Override multiple elements to apply a cohesive color scheme:

```{r vpc-sim-data, message=FALSE, warning=FALSE, include=FALSE}
model <- model_mread_load("pkmodel")
simout <- df_mrgsim_replicate(data_sad, model, replicates = 200, dv_var = "ODV")
```

```{r example-vpc, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
plot_vpc_cont(
  sim = simout,
  pcvpc = TRUE,
  vpc_theme = plot_vpc_theme(obs = pmx_point(color = "#000000"),
                            obs_median = pmx_line(color = "#000000"),
                            obs_ci = pmx_line(color = "#000000"),
                            sim_median = pmx_ribbon(fill = "#3388cc"),
                            sim_pi = pmx_ribbon(fill = "#3388cc")),
  min_bin_count = 5
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_y_log10(guide = "axis_logticks")
```
