---
title: "Plot Themes and Aesthetics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{plot-aesthetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will demonstrate `pmxhelpr` helper functions for updating plot aesthetics across `plot_dvtime`, `plot_dvconc`, `plot_gof`, and `plot_vpc_cont` functions reviewed in detail the other Vignettes for PK and PK/PD exploratory data analysis and VPC workflows.  

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(forcats, warn.conflicts = FALSE)
library(Hmisc, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
library(PKNCA, warn.conflicts = FALSE)
```

# Data

The example dataset used in this vignette is based on a single ascending dose (SAD) study of an orally administered drug product with a parallel group food effect (FE) cohort. 

## data_sad

Dataset definitions can be viewed by calling `?data_sad`. A quick summary of variables and types is printed below.

```{r data-sad-glimpse}
glimpse(data_sad)
```

The study design consists of two parts. 

  + Part 1 was a sequential single ascending dose study covering a 10 to 400 mg dose range.
  + Part 2 was a parallel 100 mg single dose food effect (FE) study. 
  

