---
title: "Plot Themes and Aesthetics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{plot-aesthetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette documents the `pmxhelpr` theme system for controlling plot aesthetics across `plot_dvtime`, `plot_dvconc`, `plot_gof`, and `plot_vpc_cont`. Each plot function accepts a `theme` argument that controls the appearance of points, lines, error bars, and other visual elements.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(forcats, warn.conflicts = FALSE)
library(Hmisc, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
```

# Element Constructors

The theme system is built on element constructors that map directly to `ggplot2` geom aesthetics:

| Constructor | Controls | Fields |
|---|---|---|
| `pmx_point()` | Point layers (`geom_point`, `stat_summary`) | `shape`, `size`, `alpha`, `color` |
| `pmx_line()` | Line layers (`geom_line`, `geom_hline`) | `linewidth`, `linetype`, `alpha`, `color` |
| `pmx_ribbon()` | Ribbon layers (`geom_ribbon`) | `fill`, `alpha`, `color`, `linetype`, `linewidth` |
| `pmx_errorbar()` | Error bar layers (`geom_errorbar`) | `linewidth`, `linetype`, `alpha`, `width` |
| `pmx_trend()` | Trend lines (`geom_smooth`) | `linewidth`, `linetype`, `color`, `se_color`, `se_alpha` |
| `pmx_style()` | Convenience shortcut | `color`, `alpha` |

`pmx_style()` is a convenience wrapper that applies shared aesthetics (color, alpha) to both the point and line elements of a role. It is used with the role-level shortcut arguments in theme factories (e.g., `obs`, `cent`, `dv`).

# Theme Factories

Each plot function has a corresponding theme factory that returns a named list of default element objects. Call with no arguments to view defaults. Pass element overrides to customize.

## `plot_dvtime_theme()`

Controls aesthetics for `plot_dvtime()`.

```{r dvtime-theme-defaults}
plot_dvtime_theme()
```

Theme keys use a `role_element` naming convention:

- `obs_point`, `obs_line` — observed data scatter and spaghetti lines
- `cent_point`, `cent_line` — central tendency points and lines
- `errorbar` — error bars for central tendency
- `ref` — reference lines (LLOQ, change-from-baseline)

Role-level shortcuts `obs` and `cent` accept `pmx_style()` to set shared aesthetics across both point and line:

```{r dvtime-style-shortcut}
# Set alpha for both obs_point and obs_line at once
plot_dvtime_theme(obs = pmx_style(alpha = 0.3))
```

### Example: Customizing an Individual Plot

```{r dvtime-data, message=FALSE, warning=FALSE}
plot_data <- data_sad %>%
  filter(CMT == 2) %>%
  mutate(`Dose and Food` = var_addn(
    paste(DOSE, "mg", ifelse(FOOD == 0, "Fasted", "Fed")), ID)) %>%
  mutate(`Dose and Food` = fct_relevel(`Dose and Food`,
                                        "50 mg Fasted (n=6)", after = 1))
```

Say we want to reduce the linewidth of error bars and reduce the size of the mean summary points to only visualize the lines:

```{r dvtime-theme-inline, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data, dv_var = ODV, col_var = `Dose and Food`,
            cent = "mean_sdl", log_y = TRUE, obs_dv = FALSE,
            theme = plot_dvtime_theme(errorbar = pmx_errorbar(linewidth = 0.5),
                                     cent_point = pmx_point(size = 0.1))) +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

### Example: Reusable Custom Theme

A theme object can be created once and reused across multiple plots:

```{r dvtime-theme-reuse, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
new_theme <- plot_dvtime_theme(errorbar = pmx_errorbar(linewidth = 0.5),
                              cent_point = pmx_point(size = 0.1))

plot_dvtime(data = plot_data, dv_var = ODV, col_var = `Dose and Food`,
            cent = "mean_sdl", log_y = TRUE, obs_dv = FALSE,
            theme = new_theme) +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

### Example: Custom Error Bar Width

The default error bar width is 2.5% of the maximum nominal time. Override with the `width` field of `pmx_errorbar()`:

```{r dvtime-theme-ebwidth, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data, dv_var = ODV, col_var = `Dose and Food`,
            cent = "mean_sdl", log_y = TRUE, obs_dv = FALSE,
            theme = plot_dvtime_theme(errorbar = pmx_errorbar(width = 8))) +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```


## `plot_gof_theme()`

Controls aesthetics for `plot_gof()`. Each output variable has separate point and line elements, plus a `color` field used for the plot legend.

```{r gof-theme-defaults}
plot_gof_theme()
```

Theme keys:

- `obs_point`, `obs_line` — observed data points and spaghetti lines (default: darkgrey)
- `dv_point`, `dv_line` — DV central tendency (default: blue)
- `pred_point`, `pred_line` — PRED central tendency (default: red)
- `ipred_point`, `ipred_line` — IPRED central tendency (default: green)
- `errorbar` — error bars (shared across variables)
- `ref` — reference lines

`obs` and `dv` represent the same underlying data (`dv_var`). They are differentiated to allow visual distinction between individual observations and summary statistics.

Role-level shortcuts (`obs`, `dv`, `pred`, `ipred`) accept `pmx_style()`:

### Example: Updating Output Colors

```{r gof-data, message=FALSE, warning=FALSE}
gof_data <- data_sad_pkfit %>%
  filter(EVID == 0) %>%
  mutate(`Dose and Food` = var_addn(
    paste(DOSE, "mg", ifelse(FOOD == 0, "Fasted", "Fed")), ID)) %>%
  mutate(`Dose and Food` = fct_relevel(`Dose and Food`,
                                        "50 mg Fasted (n=6)", after = 1))
```

Update the `obs` color to blue to align with `dv`:

```{r gof-theme-color, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_gof(data = gof_data, dv_var = "ODV",
         theme = list(obs = pmx_style(color = "blue"))) +
  facet_wrap(~`Dose and Food`, scales = "free") +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)")
```


## `plot_dvconc_theme()`

Controls aesthetics for `plot_dvconc()`.

```{r dvconc-theme-defaults}
plot_dvconc_theme()
```

Theme keys:

- `obs` — observed data points (`pmx_point`)
- `ref` — reference line (`pmx_line`)
- `loess` — LOESS trend line (`pmx_trend`)
- `linear` — linear trend line (`pmx_trend`)


## `plot_vpc_theme()`

Controls aesthetics for `plot_vpc_cont()`.

```{r vpc-theme-defaults}
plot_vpc_theme()
```

Theme keys:

- `obs` — observed data points (`pmx_point`)
- `obs_median` — observed median line (`pmx_line`)
- `obs_ci` — observed CI lines (`pmx_line`)
- `sim_pi` — simulated prediction interval ribbons (`pmx_ribbon`)
- `sim_median` — simulated median ribbon (`pmx_ribbon`)
- `loq` — LOQ reference line (`pmx_line`)
- `bin_sep` — bin separator line (`pmx_line`)

### Example: Custom VPC Color Scheme

```{r vpc-data, message=FALSE, warning=FALSE, include=FALSE}
model <- model_mread_load("pkmodel")
simout <- df_mrgsim_replicate(data_sad, model, replicates = 200, dv_var = "ODV")
```

```{r vpc-theme-custom, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
plot_vpc_cont(
  sim = simout,
  pcvpc = TRUE,
  vpc_theme = plot_vpc_theme(obs = pmx_point(color = "#000000"),
                            obs_median = pmx_line(color = "#000000"),
                            obs_ci = pmx_line(color = "#000000"),
                            sim_median = pmx_ribbon(fill = "#3388cc"),
                            sim_pi = pmx_ribbon(fill = "#3388cc")),
  min_bin_count = 5
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_y_log10(guide = "axis_logticks")
```
