---
title: "Plot Themes and Aesthetics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{plot-themes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will demonstrate `pmxhelpr` functions for viewing and modifying plot elements with package plotting functions including `plot_dvtime()`/`plot_dvconc()` and `plot_gof()`/`plot_vpc_cont()`. This vignette will introduce the theory and naming conventions behind the theme factor functions for each plot and individual element constructors.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(forcats, warn.conflicts = FALSE)
library(Hmisc, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
```



# Data and Model Objects

The example datasets used in this vignette are based around the `data_sad` dataset internal to `pmxhelpr`. This dataset is based on a single ascending dose (SAD) study of an orally administered drug product with a parallel group food effect (FE) cohort and is formatted in a analysis-ready format for non-linear mixed effects (NLME) Population PK or PK/PD modeling. `data_sad_pkfit` is a modified version of `data_sad`, which includes additional PK model predictions appended.

This vignette will assume familiarity with these internal datasets from other vignettes. These elements will not be reviewed in detail in this vignette, which is focused only on theme aesthetic controls across plotting functions.

## data_sad

Dataset definitions can be viewed by calling `?data_sad`. A quick summary of variables and types is printed below.

PART specifies the two study cohorts:

  - Single Ascending Dose (SAD)
  - Food Effect (FE).

```{r data-part}
unique(data_sad$PART)
```

Let's define some variables that may be useful in plotting and filter down to PK or PD relevant records.

```{r data-factors, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
data <- data_sad %>% 
  mutate(`Food Status` = ifelse(FOOD == 0, "Fasted", "Fed"), 
         DoseFood = paste(DOSE, "mg", `Food Status`)) %>% 
  mutate(`Food Status` = var_addn(`Food Status`, ID),
         `Dose and Food` = var_addn(DoseFood, ID)) %>% 
  mutate(`Dose and Food` = fct_relevel(`Dose and Food`, "50 mg Fasted (n=6)", after = 1))

gof_data <- data_sad_pkfit %>%
  mutate(`Food Status` = ifelse(FOOD == 0, "Fasted", "Fed"), 
         DoseFood = paste(DOSE, "mg", `Food Status`)) %>% 
  mutate(`Food Status` = var_addn(`Food Status`, ID),
         `Dose and Food` = var_addn(DoseFood, ID)) %>% 
  mutate(`Dose and Food` = fct_relevel(`Dose and Food`, "50 mg Fasted (n=6)", after = 1))

data_pk <- filter(data, CMT %in% c(1,2))
data_pd <- filter(data, CMT %in% c(1,3))
data_pkfit <- filter(gof_data, CMT %in% c(1,2))
```

An example PK model (`pkmodel`) in `mrgmod` format is provided in the internal package library. This is loaded using the helper function `model_mread_load()`, which wraps `mrgsolve::mread`.

```{r model-pk}
model <- model_mread_load("pkmodel")
```



# Overview

Every `pmxhelpr` plot function (`plot_dvtime()`, `plot_dvconc()`, `plot_gof()`, `plot_vpc_cont()`) accepts a `theme` argument that controls the appearance of points, lines, ribbons, error bars, and reference lines. The theme system has two layers:

1. **Element constructors** (`pmx_point()`, `pmx_line()`, etc.) define *what* to style. Each constructor maps directly to a `ggplot2` geom and only accepts aesthetics relevant to that geom.

2. **Theme factories** (`plot_dvtime_theme()`, `plot_gof_theme()`, etc.) define *where* those elements go. Each factory returns a named list of default element objects, one per visual layer in the plot. User overrides are merged field-by-field into these defaults.

This design means every plot function uses the same workflow:

```{r workflow-example, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
# View defaults
plot_dvtime_theme()
plot_dvtime(data_pk, dv_var = ODV, theme = plot_dvtime_theme())

# Override specific fields — unspecified fields keep their defaults
new_dvtime_theme <- plot_dvtime_theme(cent_line = pmx_line(linetype = "dashed"))
plot_dvtime(data_pk, dv_var = ODV, theme = new_dvtime_theme)
```


# Element Constructors

Each constructor creates a named list with a class tag. Only non-NULL fields are stored, so partial overrides are easy to define by only passing the elements to overide.

```{r constructor-ex1}
# Full element with all fields
pmx_point(shape = 16, size = 1.25, alpha = 1, color = "blue")
```

```{r constructor-ex2}
# Partial override — only the fields you want to change
pmx_point(size = 2)
```

When a user-supplied element is merged into a default, each specified field replaces the corresponding default. Unspecified fields are preserved. This occurs without user input in the plot theme constructor factory functions.

```{r merge-example}
default <- pmx_point(shape = 1, size = 0.75, alpha = 0.5)
override <- pmx_point(size = 2)
pmxhelpr:::merge_element(override, default)
```

The available element constructors are as follows:

  + `pmx_point()`
    + `ggplot` layers:`geom_point`, `stat_summary`
    + function fields: `shape`, `size`, `alpha`, `color`
  + `pmx_line()`
    + `ggplot` layers:`geom_line`, `geom_hline`
    + function fields: `shape`, `size`, `alpha`, `color`
  + `pmx_ribbon()`
    + `ggplot` layers:`geom_ribbon`
    + function fields: `fill`, `alpha`, `color`, `linetype`, `linewidth`
  + `pmx_errorbar()`
    + `ggplot` layers:`geom_errorbar`,
    + function fields: `linewidth`, `linetype`, `alpha`, `width` 
  + `pmx_trend()`
    + `ggplot` layers:`geom_smooth`,
    + function fields: `linewidth`, `linetype`, `color`, `se_color`, `se_alpha`
  + `pmx_style()`
    + Convenience shortcut function
    + function fields:  `alpha`, `color`



# Theme Factories


## Naming Convention

Theme keys follow a `role_element` pattern where the **role** identifies the data layer and the **element** identifies the geom type:

- `obs_point` — observed data, point geom
- `cent_line` — central tendency, line geom
- `ref_line` — reference line, line geom

When a role uses only one geom type, the key uses the role plus the element suffix (e.g., `ref_line`, `loq_line`).

## Shared Elements Across Factories

Several element roles appear across multiple theme factories:

  + Element Key: `ref_line`
      + Element Constructor: `pmx_line`
      + Used in: `plot_dvtime_theme`, `plot_gof_theme`, `plot_dvconc_theme`
  + Element Key: `loq_line`
      + Element Constructor: `pmx_line`
      + Used in: `plot_dvtime_theme`, `plot_gof_theme`, `plot_vpc_theme`
  + Element Key: `cent_errorbar`
      + Element Constructor: `pmx_errorbar`
      + Used in: `plot_dvtime_theme`, `plot_gof_theme`
  + Element Key: `obs_point`
      + Element Constructor: `pmx_point`
      + Used in: `plot_dvtime_theme`, `plot_dvconc` `plot_gof_theme`, `plot_dvconc_theme`

The same constructors and merge logic apply everywhere, so learning the system once covers all plot types.

## Role-Level Shortcuts with `pmx_style()`

When a role has both `_point` and `_line` elements (e.g., `obs_point` and `obs_line`), you can set shared aesthetics on both at once using `pmx_style()`:

```{r style-shortcut}
# These two calls produce the same result:
plot_dvtime_theme(obs_point = pmx_point(alpha = 0.3),
                  obs_line = pmx_line(alpha = 0.3))

plot_dvtime_theme(obs = pmx_style(alpha = 0.3))
```

`pmx_style()` accepts `color` and `alpha` — the two fields shared by both `pmx_point()` and `pmx_line()`.

When both a style shortcut and an explicit element override are provided, the style is applied first, then the element override wins for any conflicting fields:

```{r style-precedence}
# Style sets alpha on both point and line,
# then the explicit override changes just the point size
theme <- plot_dvtime_theme(
  obs = pmx_style(alpha = 0.3),
  obs_point = pmx_point(size = 3)
)
theme$obs_point  # alpha = 0.3, size = 3
theme$obs_line   # alpha = 0.3, linewidth = 0.5 (default)
```


## Defaults in Theme Factories

### Defaults for `plot_dvtime()` in `plot_dvtime_theme()`

```{r dvtime-defaults}
plot_dvtime_theme()
```

The relationship between element key, element constructor, and purpose is defined below:

  + `obs_point` | `pmx_point` | Observed data points 
  + `obs_line` | `pmx_line` | Individual spaghetti lines connecting observations
  + `cent_point` | `pmx_point` | Central tendency points
  + `cent_line` | `pmx_line` | Central tendency lines 
  + `cent_errorbar` | `pmx_errorbar` | Error bars for central tendency 
  + `ref_line` | `pmx_line` | Reference line (e.g., change-from-baseline) 
  + `loq_line` | `pmx_line` | LOQ reference line

Shortcuts: `obs` = `obs_point`/`obs_line`, `cent` = `cent_point`/`cent_line` accept `pmx_style()`.

### Defaults for `plot_gof()` in `plot_gof_theme()`

```{r gof-defaults}
plot_gof_theme()
```

The relationship between element key, element constructor, and purpose is defined below:

 + `obs_point` | `pmx_point` | Observed data points (default: darkgrey)
 + `obs_line` | `pmx_line` | Individual spaghetti lines connecting observations (default: darkgrey)
 + `cent_point` | `pmx_point` | Shared central tendency points for DV, PRED, IPRED |
 + `cent_line` | `pmx_line` | Shared central tendency lines for DV, PRED, IPRED |
 + `cent_errorbar` | `pmx_errorbar` | Shared central tendency error bars 
 + `ref_line` | `pmx_line` | Reference line
 + `loq_line` | `pmx_line` | LOQ reference line
 + `cent_color` | `pmx_color` | Overlay color mapping `dv`, `pred`, `ipred` (defaults: blue, red, green)

Shortcuts: `obs` = `obs_point`/`obs_line`, `cent` = `cent_point`/`cent_line` accept `pmx_style()`. Overlay colors are controlled via `cent_color = pmx_color()`.

### Defaults for `plot_dvconc` in `plot_dvconc_theme()`

```{r dvconc-defaults}
plot_dvconc_theme()
```

The relationship between element key, element constructor, and purpose is defined below:

  + `obs_point` | `pmx_point` | Observed data points 
  + `ref_line` | `pmx_line` | Reference line (e.g., change-from-baseline) 
  + `loess` | `pmx_trend` | LOESS trend line and confidence interval
  + `linear` | `pmx_trend` | Linear trend line and confidence interval
  
Shortcuts using `pmx_style()` are not relevant for this function.

### Defaults for `plot_vpc_cont` in `plot_vpc_theme()`

```{r vpc-defaults}
plot_vpc_theme()
```

The relationship between element key, element constructor, and purpose is defined below:
  + `obs_point` | `pmx_point` | Observed data points
  + `obs_median_line` | `pmx_line` | Observed median line
  + `obs_pi_line` | `pmx_line` | Observed quantile lines
  + `sim_pi_line` | `pmx_line` | Simulated prediction interval lines
  + `sim_pi_ci` | `pmx_ribbon` | Simulated PI confidence interval ribbons
  + `sim_pi_area` | `pmx_ribbon` | Simulated PI area ribbon 
  + `sim_median_line` | `pmx_line` | Simulated median line
  + `sim_median_ci` | `pmx_ribbon` | Simulated median CI ribbon
  + `loq_line` | `pmx_line` | LOQ reference line |

Shortcuts using `pmx_style()` are not relevant for this function.

# Examples

The following examples demonstrate common customization patterns. Each pattern applies to any plot function — only the theme factory and element keys change. These examples will endeavor to cover common use cases for default theme overrides. 

## Make Theme Overrides Inline

Pass element overrides directly in the `theme` argument. Only the fields you specify are changed. This example removes points for observations and central tendency means to emphasize the error bars

```{r example-inline, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = data_pk, dv_var = ODV, col_var = `Dose and Food`,
            cent = "mean_sdl", log_y = TRUE,
            theme = plot_dvtime_theme(obs_point = pmx_point(alpha = 0),
                                     cent_point = pmx_point(size = 0))) +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

## Make Theme Overrides with a Reusable Theme Object

A more common approach to modifying theme elements is to create a reusable theme object that can be passed to a series of plots. The example below produces the same plot output as the inline example above, but can be recycled to additional plots.

```{r example-reuse, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
my_theme <- plot_dvtime_theme(cent_point = pmx_point(size = 0), 
                              obs_point = pmx_point(alpha = 0))

plot_dvtime(data = data_pk, dv_var = ODV, col_var = `Dose and Food`,
            cent = "mean_sdl", log_y = TRUE,
            theme = my_theme) +
  scale_x_continuous(breaks = seq(0, 168, 24))+
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

## Remove Individual Data Points

A common use case for the `theme` argument is to set to `alpha = 0` for data points to remove them from the plot. This example removes points for observations and central tendency means to emphasize the error bars

```{r example-no-pts, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = data_pk, dv_var = ODV, col_var = PART,
            cent = "mean_sdl", dosenorm = TRUE, log_y = TRUE,
            theme = plot_dvtime_theme(obs_point = pmx_point(alpha = 0),
                                     cent_point = pmx_point(size = 0))) +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

## Modifying color / alpha for points / lines simultaneously with `pmx_style()`

Use `pmx_style()` to change the color of both point and line elements for a role in one call. This example uses it to change the color and alpha of the individual spaghetti line colors and the color and alpha of the central tendency line.

This is useful when you want to fade or recolor an entire layer without repeating yourself.

```{r example-style1, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = data_pk, dv_var = ODV, id_var = ID,
            theme = plot_dvtime_theme(obs = pmx_style(color = "grey", alpha = 0.3), 
                                      cent = pmx_style(color = "darkred", alpha = 1)), 
            log_y = TRUE) +
  scale_x_continuous(breaks = seq(0, 168, 24)) +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") 
```


## Modifying Error Bar Width

The default error bar width is 2.5% of the maximum nominal time. Override with the `width` field:

```{r example-eb-width, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = data_pk, dv_var = ODV, col_var = `Dose and Food`,
            cent = "mean_sdl", log_y = TRUE,
            theme = plot_dvtime_theme(obs_point = pmx_point(alpha = 0),
                                     cent_errorbar = pmx_errorbar(width = 10))) +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

## Trend Line Customization with `pmx_trend()`

`pmx_trend()` controls both the trend line and its standard error ribbon. The `color`, `linewidth`, and `linetype` fields style the line itself, while `se_color` and `se_alpha` style the confidence ribbon shown when `se_loess = TRUE` or `se_linear = TRUE`.

An example override of the line color and SE ribbon appearances is provided below.

```{r example-trend, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = data_pd, dv_var = CFB, idv_var = CONC,
            loess = TRUE, se_loess = TRUE, ref = 0,
            theme = plot_dvconc_theme(
              loess = pmx_trend(color = "steelblue", linetype = 1,
                                se_color = "steelblue", se_alpha = 0.2))) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```

Both trend elements can be customized independently. Here the LOESS line is styled as a solid blue line and the linear fit as a dashed red line.

```{r example-trend-both, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = data_pd, dv_var = CFB, idv_var = CONC,
            ref = 0,
            loess = TRUE, linear = TRUE,
            se_loess = TRUE, se_linear = TRUE,
            theme = plot_dvconc_theme(
              loess  = pmx_trend(color = "steelblue", se_color = "steelblue", se_alpha = 0.15),
              linear = pmx_trend(color = "firebrick", linetype = 2,
                                 se_color = "firebrick", se_alpha = 0.15))) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```

## Customize Colors for Central Tendency using `plot_gof_theme()`

The `cent_color` key in `plot_gof_theme()` can be used to change the colors of central tendency lines for PRED, IPRED, and DV. The example below sets the observed data and observed data mean line to black and the IPRED mean line to red with the PRED line suppressed.

```{r example-gof-color, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_gof(data = data_pkfit, dv_var = ODV, log_y = TRUE,
         theme = plot_gof_theme(obs_point = pmx_point(color = "black"), 
                                cent_color = pmx_color(dv = "black", ipred = "red")), 
         shown = plot_gof_shown(pred = FALSE)) +
  scale_x_continuous(breaks = seq(0, 168, 24))+
  labs(x = "Time Since First Dose (hours)", y = "Drug Concentration (ng/mL)") + 
  facet_wrap(~DoseFood)
```

## Customize VPC Color Scheme with `plot_vpc_theme()`

For VPC workflows, it is almost always preferred to set alternative theme objects that can be recycled into different VPC plot strata. 

The example below sets a new VPC theme using a blue / black color schema.

```{r new-vpc-theme, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
new_vpc_theme <- plot_vpc_theme(obs_point = pmx_point(color = "#000000"),
                                obs_median_line = pmx_line(color = "#000000"),
                                obs_pi_line = pmx_line(color = "#000000"),
                                sim_median_ci = pmx_ribbon(fill = "#3388cc"),
                                sim_pi_ci = pmx_ribbon(fill = "#3388cc"), 
                                sim_pi_area = pmx_ribbon(fill = "#3388cc"))
```


```{r vpc-sim-data, message=FALSE, warning=FALSE, include=FALSE}
simout <- df_mrgsim_replicate(data_pk, model, replicates = 100, dv_var = "ODV")
```

```{r example-vpc, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
plot_vpc_cont(
  sim = simout,
  pcvpc = TRUE,
  theme = new_vpc_theme,
  min_bin_count = 4,
  shown = plot_vpc_shown(sim_pi_area = TRUE, sim_pi_ci = FALSE, 
                         obs_median_line = TRUE, obs_pi_line = FALSE, 
                         sim_median_ci = FALSE)
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_y_log10(guide = "axis_logticks")
```
