---
title: "VPC Plots with Exact Bins"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vpc-exact-bins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will review the functionality of `plot_vpc_cont()` for generating VPC plots using exact time bins from datasets containing a nominal time variable.

Many VPC tools use automatic binning algorithms (e.g., Jenks natural breaks, k-means, density) to determine bin intervals from the data. While these are useful for datasets without a pre-defined binning variable, they are not optimized to leverage input datasets with a variable representing exact bin times (e.g., nominal or protocol-specified times).

`plot_vpc_cont()` uses the unique values of the nominal time variable directly as exact bins, ensuring that summary statistics are calculated at the actual protocol-specified sampling times.

First, we will load the required packages.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(mrgsolve, warn.conflicts =  FALSE)
library(withr, warn.conflicts =  FALSE)
```

# Analysis Dataset

Next let's explore the input dataset, `data_sad`. This dataset was generated via simulation from `pkmodel`, an `mrgsolve` model internal to the pmxhelpr package.

We can take a quick look at the dataset using `glimpse()` from the dplyr package. Documentation for the dataset can also be viewed using the R help functionality, just as one would for a function, with `?data_sad()`

```{r data-glimpse}
glimpse(data_sad)
```

This dataset is  formatted for modeling. It contains NONMEM reserved variables (e.g., ID, TIME, AMT, EVID, MDV), as well as, dependent variables of drug concentration in original units (ODV) and natural logarithm transformed units (LDV). In addition to the numeric variables, there are two character variables: USUBJID and PART.

PART specifies the two study cohorts:

  - Single Ascending Dose (SAD)
  - Food Effect (FE).

```{r data-part}
unique(data_sad$PART)
```

This dataset also contains an exact binning variable:

  - Nominal Time (NTIME).

This variable represents the nominal time of sample collection relative to first dose per study protocol whereas Actual Time (TIME) represents the actual time the sample was collected.

```{r data-ntime}
##Unique values of NTIME
unique(data_sad$NTIME)

##Comparison of number of unique values of NTIME and TIME
length(unique(data_sad$NTIME))
length(unique(data_sad$TIME))
```
# PK Model

Luckily for us, someone has already fit a PK model to these data! Let's load `pkmodel` by calling `model_mread_load()` and take a look at it with the `see()` function from the `mrgsolve` package.

```{r model_mread_load}
model <- model_mread_load("pkmodel")
see(model)
```

Unluckily for us, no one has validated this PK model! Therefore, we need to generate some Visual Predictive Checks (VPCs).

# VPC Plot Workflow

## Running the simulation

We will use `df_mrgsim_replicate()` to run the simulation for the VPC. `df_mrgsim_replicate()` is a wrapper function for `mrgsim_df()`, which uses `lapply()` to iterate the simulation over integers from 1 to the value passed to the argument `replicates`.

We can pass `data_sad` and `model` from the previous steps to the `data` and `model` arguments, respectively, and run the simulation for 100 `replicates`. The names of actual and nominal time variables in `data_sad` match the default arguments; however, our dependent variable is named `ODV`, which must be specified in the `dv_var` argument, since it differs from the default (`DV`). This argument uses non-standard evaluation and can be passed as a bare column name or as a string.

We would like to recover the numerical variables `"DOSE"` and `"FOOD"` and the character variable `"PART"` from the input dataset, as we may need these study conditions to stratify our VPC plots. We will request `"BLQ"` and `"LLOQ"` for potential assessment of impact of censoring in the VPCs, as well as, the NONMEM reserved variables `"CMT"`, `"EVID"`, and `"MDV"`. Finally, we will add the argument `obsonly = TRUE`, which is passed to `mrgsim()`, to remove dose records from the simulation output and reduce file size.

```{r df_mrgsim_replicate}
data_sad_pk <- filter(data_sad, CMT %in% c(1,2))

simout <- df_mrgsim_replicate(data = data_sad_pk,
                     model = model,
                     replicates = 100,
                     dv_var = ODV,
                     time_vars = c(TIME = "TIME", NTIME = "NTIME"),
                     output_vars = c(PRED = "PRED", IPRED = "IPRED", DV = "DV"),
                     num_vars = c("CMT", "BLQ", "LLOQ", "EVID", "MDV", "DOSE", "FOOD"),
                     char_vars = c("PART"),
                     obsonly = TRUE)

glimpse(simout)

max(simout$SIM)
```

The maximum value of our replicate count variable (default = `"SIM"`) indicates that the dataset has been replicated 100 times.

Glimpsing the output reveals the following model outputs:

  - `PRED` (population prediction)
  - `IPRED` (individual prediction)
  - `SIMDV` (simulated dependent variable)
  - `OBSDV` (observed dependent variable).

## Why Exact Bins?

Many automatic binning algorithms are designed to determine optimal binning intervals from the data. However, when a dataset contains an exact binning variable (such as nominal time), these algorithms may not reproduce the exact bins. For example, absorption phase timepoints that are close together may be grouped into a single bin, obscuring important pharmacokinetic details.

The presence of exact bins in the data is a common scenario in pharmacometrics, as Clinical Study Protocols specify the times of PK sampling.

## VPC Plots with `pmxhelpr`

`plot_vpc_cont()` uses the unique values of the nominal time variable (`NTIME`) directly as exact bins. This ensures that summary statistics are calculated at each protocol-specified sampling time, preserving the full resolution of the data.

Because we have multiple dose levels in the SAD portion of the study, as well as, a food effect cohort, prediction-correction will be used to plot all the data on a single plot using the argument `pcvpc = TRUE`.

```{r plot-exact, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
vpc_exact <- plot_vpc_cont(
  sim = simout,
  pcvpc = TRUE,
  time_vars = c(TIME = "TIME", NTIME = "NTIME"),
  output_vars = c(PRED = "PRED", IPRED = "IPRED", SIMDV = "SIMDV", OBSDV = "OBSDV"),
  pi = c(0.05, 0.95),
  ci = c(0.05, 0.95)
) +
  labs(x = "Time (hours)", y = "PRED-corrected Concentration (ng/mL)") +
  scale_y_log10(guide = "axis_logticks")

vpc_exact
```

The difference in plotting the observed prediction-corrected data points by `"TIME"` versus `"NTIME"` is negligible for this example dataset, due to the high concordance between `"TIME"` and `"NTIME"` in this example Phase 1 study. However, this difference is often much larger for pooled analyses including later phase clinical studies where plotting the observed points versus actual time will result in a plot that is much more representative of the distribution of times in the model training dataset.

`plot_vpc_cont()` contains the argument `min_bin_count` (default = 1) filters out exact bins with fewer quantifiable observations than the minimum set by this argument. Importantly, the observed data points in these small bins are *still plotted*; however, they do not influence the calculation of summary statistics or summary plot elements (shaded intervals, lines). This provides the greatest fidelity to the data visualized without introducing visual artifacts due to small sample timepoints.

Additionally, because our plot input dataset `simout` was generated using `df_mrgsim_replicate`, the `time_vars` and `output_vars` match the default values, so we do not need to specify these arguments and they can be removed. Furthermore, the `pi` and `ci` arguments we are specifying match the defaults and can be removed.

When setting`min_bin_count = 10`, summary statistics are not plotted for the final two timepoint containing only fewer than 10 quantifiable observations; however, the observations themselves are still plotted.

```{r plot-exact-min-bin-count, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
vpc_exact_bin_gt10 <- plot_vpc_cont(
  sim = simout,
  pcvpc = TRUE,
  min_bin_count = 10
) +
  labs(x = "Time (hours)", y = "PRED-corrected Concentration (ng/mL)") +
  scale_y_log10(guide = "axis_logticks")

vpc_exact_bin_gt10
```

We can also pass stratifying variables to the argument `strat_var` in order to facet our plots by relevant covariate conditions. Currently, only one variable can be passed to this argument.

```{r plot-exact-one-strata, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
vpc_exact_food <- plot_vpc_cont(
  sim = mutate(simout, FOOD_f = factor(FOOD, levels = c(0,1), labels = c("Fasted", "Fed"))),
  strat_var = FOOD_f,
  pcvpc = TRUE,
  min_bin_count = 4
) +
  labs(x = "Time (hours)", y = "PRED-corrected Concentration (ng/mL)") +
  scale_y_log10(guide = "axis_logticks")

vpc_exact_food
```

Because `plot_vpc_cont()` returns a `ggplot2` object, the y-axis can be transformed to log10 scale by adding `scale_y_log10()` as a layer. This approach avoids any data transformation prior to calculating summary statistics, ensuring all quantile calculations are performed on the original scale.
