---
title: "VPC Plots with Exact Bins"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vpc-exact-bins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will review the limitations of the `vpc()` function in the `vpc` package for datasets containing an exact binning variable. This was the motivating reason for development of `plot_vpc_exactbins()`. 

The `vpc`package offers a lot of great functionality to generate VPC plots including, but certainly not limited to: 

1) data processing and visualization steps combined in one function
2) built-in options for automatic binning for datasets without a binned variable
3) built-in option for prediction-correction
4) built-in option for censoring at the lower limit of quantification (LLOQ) 

However, although `vpc()` contains many great options to automatically identify bins in the data, it is not optimized to leverage input datasets with variable a variable representing exact bin times (e.g., nominal, or protocol-specified, times). 

First, we will load the required packages.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(vpc, warn.conflicts =  FALSE)
library(mrgsolve, warn.conflicts =  FALSE)
library(withr, warn.conflicts =  FALSE)
```

Next let's explore the input dataset, `data_sad`. This dataset was generated via simulation from `model`, an `mrgsolve` model internal to the pmxhelpr package. 

We can take a quick look at the dataset using `glimpse()` from the dplyr package.

```{r data-glimpse}
glimpse(data_sad)
```

We can see that this dataset is already formatted for modeling. It contains NONMEM reserved variables (e.g., ID, TIME, AMT, EVID, MDV), as well as, dependent variables of drug concentration in original units (ODV) and natural logarithm transformed units (LDV). In addition to the numeric variables, there are two character variables: USUBJID and PART. 

PART specifies the two study cohorts: 

  - Single Ascending Dose (SAD)
  - Food Effect (FE).

```{r data-part}
unique(data_sad$PART)
```

This dataset also contains an exact binning variable: 

  - Nominal Time (NTIME). 

This variable represents the nominal time of sample collection relative to first dose per study protocol whereas Actual Time (TIME) represents the actual time the sample was collected.

```{r data-ntime}
##Unique values of NTIME
unique(data_sad$NTIME)

##Comparison of number of unique values of NTIME and TIME
length(unique(data_sad$NTIME))
length(unique(data_sad$TIME))
```

Let's visualize the data. First, we will derive a factor variable from DOSE to pass to the color aesthetic, as well as, a dependent variable (DV) with BLQ values imputed to 1/2 x LLOQ.

```{r data-plotdata}
plot_data <- data_sad %>% 
  filter(EVID == 0) %>% 
  mutate(`Dose (mg)` = factor(DOSE), 
         DV = case_when(BLQ == 0 ~ ODV, 
                        BLQ %in% c(-1, 1) ~ LLOQ/2, 
                        .default = NA_real_), 
         DNDV = ODV/DOSE)
```

Now let's plot the data using ODV colored by DOSE and faceted by PART. The concentration-time profiles increase with dose with some potential impact of censoring at the LLOQ in the late terminal phase.

```{r plot-conc-time, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
ggplot(aes(x = TIME, y = ODV, col = `Dose (mg)`), data = plot_data)+
  geom_point(shape = 1, alpha = 0.5)+
  stat_summary(fun = "mean", geom = "point", 
               aes(x = NTIME, y = ODV, col = `Dose (mg)`), inherit.aes = FALSE) +
  stat_summary(fun = "mean", geom = "line", 
               aes(x = NTIME, y = ODV, col = `Dose (mg)`), inherit.aes = FALSE) +
  geom_hline(yintercept = unique(plot_data$LLOQ), linetype = "dashed")+
  scale_x_continuous(breaks = seq(0,168,24))+
  scale_y_log10()+
  facet_wrap(~PART)+
  labs(y = "Concentration (ng/mL)", x = "Time Since Dose Administration (hours)")+
  theme_bw()
```

Imputing post-dose concentrations below the lower limit of quantification as 1/2 x LLOQ normalizes the late terminal phase of the concentration-time profile. This suggests that the artifact in the late terminal phase is due to censoring of observations below the LLOQ, which impacts lower doses more than higher doses.

```{r plot-conc-time-blq, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
ggplot(aes(x = TIME, y = DV, col = `Dose (mg)`), data = plot_data)+
  geom_point(shape = 1, alpha = 0.5)+
  stat_summary(fun = "mean", geom = "point", 
               aes(x = NTIME, y = DV, col = `Dose (mg)`), inherit.aes = FALSE) +
  stat_summary(fun = "mean", geom = "line", 
               aes(x = NTIME, y = DV, col = `Dose (mg)`), inherit.aes = FALSE) +
  geom_hline(yintercept = unique(plot_data$LLOQ), linetype = "dashed")+
  scale_x_continuous(breaks = seq(0,168,24))+
  scale_y_log10()+
  facet_wrap(~PART)+
  labs(y = "Concentration (ng/mL)", x = "Time Since Dose Administration (hours)")+
  theme_bw()
```

Luckily for us, someone has already fit a PK model to these data! Let's load `model` by calling `model_mread_load()` and take a look at it with the `see()` function from the `mrgsolve` package.

```{r model_mread_load}
model <- model_mread_load("model")
see(model)
```

Unluckily for us, no one has validated this PK model! Therefore, we need to generate some Visual Predictive Checks (VPCs). 

We will use `df_mrgsim_replicate()` to run the simulation for the VPC. `df_mrgsim_replicate()` is a wrapper function for `mrgsim_df()`, which uses `lapply()` to iterate the simulation over integers from 1 to the value passed to the argument `replicates`.

We can pass `data_sad` and `model` from the previous steps to the `data` and `model` arguments, respectively, and run the simulation for 100 `replicates`. The names of actual and nominal time variables in `data_sad` match the default arguments; however, our dependent variable is named `"ODV"`, which must be specified in the output_vars argument. 

**Note: Currently output variables `IPRED` and `DV` are required in `model` capture statement**

We would like to recover the numerical variables `"DOSE"` and `"FOOD"` and the character variable `"PART"` from the input dataset, as we may need these study conditions to stratify our VPC plots. We will request `"BLQ"` and `"LLOQ"` for potential assessment of impact of censoring in the VPCs, as well as, the NONMEM reserved variables `"CMT"`, `"EVID"`, and `"MDV"`. Finally, we will add the argument `obsonly = TRUE`, which is passed to `mrgsim()`, to remove dose records from the simulation output and reduce file size.

```{r df_mrgsim_replicate}
simout <- df_mrgsim_replicate(data = data_sad, 
                     model = model, 
                     replicates = 100, 
                     time_vars = c(TIME = "TIME", NTIME = "NTIME"),
                     output_vars = c(PRED = "PRED", IPRED = "IPRED", DV = "ODV"),
                     num_vars = c("CMT", "BLQ", "LLOQ", "EVID", "MDV", "DOSE", "FOOD"),
                     char_vars = c("PART"),
                     obsonly = TRUE)

glimpse(simout)

max(simout$SIM)
```

The maximum value of our replicate count variable (default = `"SIM"`) indicates that the dataset has been replicated 100 times. 

Glimpsing the output reveals the following model outputs: 

  - `PRED` (population prediction)
  - `IPRED` (individual prediction)
  - `SIMDV` (simulated dependent variable)
  - `OBSDV` (observed dependent variable). 

Now that we have run the simulation, we generate our VPC plots using `vpc()`.The documentation for `vpc()` (https://vpc.ronkeizer.com/binning.html) describes the following binning methods:

  - `time`: Divide bins equally over time (or whatever independent variable is used). Recommended only when there is no observable clustering in the independent variable.
  - `data`: Divide bins equally over the amount of data ordered by independent variable. Recommended only when data are for nominal timepoints and all datapoints are available.
  - `density`: Divide bins based on data-density, i.e. place the bin-separators at nadirs in the density function. An approximate number of bins can be specified, but it is not certain that the algorithm will strictly use the specified number. More info in `?auto_bin()`.
  - `jenks`: Default and recommended method. Jenk's natural breaks optimization, similar to K-means clustering.
  - `kmeans`: K-means clustering.
  - `pretty`, `quantile`, `hclust`, `sd`, `bclust`, `fisher`. Methods provided by the classInt package, see the package help for more information.

We can visualize the bins assigned by the various binning approaches in `vpc()` using by assigning the plot as an object , let's call it `plot_object`, and calling `plot_object$data`.

Because we have multiple dose levels in the SAD portion of the study, as well as, a food effect cohort, prediction-correction will be used to plot all the data on a single plot using the argument `pred_corr = TRUE`.

```{r plot-jenks, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
vpc_jenks <- vpc(
  sim = simout, 
  obs = filter(simout, SIM == 1), 
  bins = "jenks", #default
  n_bins = "auto", 
  sim_cols = list(dv = "SIMDV", idv = "NTIME", pred = "PRED"),
  obs_cols = list(dv = "OBSDV", idv = "NTIME", pred = "PRED"),
  pred_corr = TRUE,
  pi = c(0.05, 0.95),
  ci = c(0.05, 0.95),
  show = list(obs_dv = TRUE),
  log_y = TRUE,
  xlab = "Time (hours)",
  ylab = "PRED-corrected Concentration (ng/mL)"
) +
  scale_x_continuous(breaks = seq(0,168,24))

vpc_jenks
```

This plot looks pretty good! The default binning in `vpc::vpc()` produce a very nice looking plot. However, all of the binning methods internal to `vpc()` are designed to determine binning intervals from the data. Currently, there is no method to use *exact bins contained in the data* in place of *bin intervals determined from the data*. The presence of exact bins in the data is a common scenario in pharmacometrics, as Clinical Study Protocols usually specify specific Study Visits corresponding ot each sample, which are associated with a nominal (protocol-specified) time of collection.

We can use the `df_nobsbin()` function to calculate and return a summary data.frame containing the unique exact bin times, count of non-missing observations (EVID=0 & MDV=0), and count of missing (EVID=0 & MDV=1) observations.

```{r bins-jenks}
##Exact bins in the input data
df_nobsbin(data_sad, bin_var = "NTIME")

##Bin midpoints and boundaries determined by vpc() using bins = "jenks"
distinct(select(vpc_jenks$data, bin_mid, bin_min, bin_max))
```

The `bin_mid` variable is where`vpc()` will plot the summary statistics calculated for the observed and simulated data. 

We can clearly see that the default `bins = "jenks"` method does not reproduce the exact bins in the observed dataset, *even when passing nominal, rather than actual, time as the independent variable (`idv`)*. How about the other binning methods native to `vpc()`? Let's take a look at `bins = "pretty"` next.

```{r plot-pretty, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
vpc_pretty <- vpc(
  sim = simout, 
  obs = filter(simout, SIM == 1), 
  bins = "pretty", 
  n_bins = "auto", 
  sim_cols = list(dv = "SIMDV", idv = "NTIME", pred = "PRED"),
  obs_cols = list(dv = "OBSDV", idv = "NTIME", pred = "PRED"),
  pred_corr = TRUE,
  pi = c(0.05, 0.95),
  ci = c(0.05, 0.95),
  show = list(obs_dv = TRUE),
  log_y = TRUE,
  xlab = "Time (hours)",
  ylab = "PRED-corrected Concentration (ng/mL)"
) +
  scale_x_continuous(breaks = seq(0,168,24))

vpc_pretty
```

The vpc plot produced with `bins = "pretty` *is* fairly pretty; however, again we can see that the binning is not true to the exact bins in our dataset, especially at the earlier absorption phase time-points, which are being largely binned together.

```{r bins-pretty}
##Exact bins in the input data
df_nobsbin(data_sad, bin_var = "NTIME")

##Bin midpoints and boundaries determined by vpc() using bins = "pretty"
distinct(select(vpc_pretty$data, bin_mid, bin_min, bin_max))
```

The `bins = "kmeans"` option produces yet another reasonable plot; however, like `bins = "pretty"`, it groups many of the absorption phase timepoints together and does not include the last three sampling times with quantifiable observations in simulated intervals. 

```{r plot-kmeans, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
vpc_kmeans <- vpc(
  sim = simout, 
  obs = filter(simout, SIM == 1), 
  bins = "kmeans", 
  n_bins = "auto", 
  sim_cols = list(dv = "SIMDV", idv = "NTIME", pred = "PRED"),
  obs_cols = list(dv = "OBSDV", idv = "NTIME", pred = "PRED"),
  pred_corr = TRUE,
  pi = c(0.05, 0.95),
  ci = c(0.05, 0.95),
  show = list(obs_dv = TRUE),
  log_y = TRUE,
  xlab = "Time (hours)",
  ylab = "PRED-corrected Concentration (ng/mL)"
) +
  scale_x_continuous(breaks = seq(0,168,24))

vpc_kmeans
```

```{r bins-kmeans}
##Exact bins in the input data
df_nobsbin(data_sad, bin_var = "NTIME")

##Bin midpoints and boundaries determined by vpc() using bins = "kmeans"
distinct(select(vpc_kmeans$data, bin_mid, bin_min, bin_max))
```

The native binning method `density` attempts to bin the data by finding the nadir in the density function. In this case, we can try and inform the binning algorithm on how many bins we *expect* in the data. Let's pass the length of the vector of unique `NTIME` values in our dataset to the `n_bins` argument and see if the `density` approach can find the correct bins.

```{r plot-density, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
vpc_density <- vpc(
  sim = simout, 
  obs = filter(simout, SIM == 1), 
  bins = "density", 
  n_bins = length(unique(simout$NTIME)), 
  sim_cols = list(dv = "SIMDV", idv = "NTIME", pred = "PRED"),
  obs_cols = list(dv = "OBSDV", idv = "NTIME", pred = "PRED"),
  pred_corr = TRUE,
  pi = c(0.05, 0.95),
  ci = c(0.05, 0.95),
  show = list(obs_dv = TRUE),
  log_y = TRUE,
  xlab = "Time (hours)",
  ylab = "PRED-corrected Concentration (ng/mL)"
) +
  scale_x_continuous(breaks = seq(0,168,24))

vpc_density
```

Whomp Whomp. `bins = "density"`, like the previous methods evaluated, grouped most of the absorption phase into a single bin.

```{r bins-density}
##Exact bins in the input data
df_nobsbin(data_sad, bin_var = "NTIME")

##Bin midpoints and boundaries determined by vpc() using bins = "density"
distinct(select(vpc_density$data, bin_mid, bin_min, bin_max))
```

This leaves us with two methods that divide the data equally into bins over the range of values in the data: `bins = "data"` (equal data density in each bin) and `bins = "time"` (equal bin width in time).

```{r plot-data-time, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
vpc_data <- vpc(
  sim = simout, 
  obs = filter(simout, SIM == 1), 
  bins = "data", 
  sim_cols = list(dv = "SIMDV", idv = "NTIME", pred = "PRED"),
  obs_cols = list(dv = "OBSDV", idv = "NTIME", pred = "PRED"),
  pred_corr = TRUE,
  pi = c(0.05, 0.95),
  ci = c(0.05, 0.95),
  show = list(obs_dv = TRUE),
  log_y = TRUE,
  xlab = "Time (hours)",
  ylab = "PRED-corrected Concentration (ng/mL)"
) +
  scale_x_continuous(breaks = seq(0,168,24))

vpc_time <- vpc(
  sim = simout, 
  obs = filter(simout, SIM == 1), 
  bins = "time", 
  sim_cols = list(dv = "SIMDV", idv = "NTIME", pred = "PRED"),
  obs_cols = list(dv = "OBSDV", idv = "NTIME", pred = "PRED"),
  pred_corr = TRUE,
  pi = c(0.05, 0.95),
  ci = c(0.05, 0.95),
  show = list(obs_dv = TRUE),
  log_y = TRUE,
  xlab = "Time (hours)",
  ylab = "PRED-corrected Concentration (ng/mL)"
) +
  scale_x_continuous(breaks = seq(0,168,24))

vpc_data
vpc_time
```

These methods are designed to equalize data density in each bin, which  does not produce binning consistent with the exact bins in our dataset.

```{r bins-data-time}
##Exact bins in the input data
df_nobsbin(data_sad, bin_var = "NTIME")

##Bin midpoints and boundaries determined by vpc() using bins = "data"
distinct(select(vpc_data$data, bin_mid, bin_min, bin_max))

##Bin midpoints and boundaries determined by vpc() using bins = "time"
distinct(select(vpc_time$data, bin_mid, bin_min, bin_max))
```

The pmxhelpr function `plot_vpc_exactbins()` is a wrapper function for `vpc()`, which is optimized for input datasets containing exact bins. This wrapper passes the the unique exact bins (e.g., nominal times) in the input dataset as bin boundaries with the addition of `Inf` to the end of the vector to ensure that the final exact bin is included, rather than set only as a boundary. 

This functionality can be reproduced using `vpc()` by passing a vector of unique exact bins to `bins` with the addition of `Inf` as follows:

- `bins = c(sort(unique(simout$NTIME)), Inf)`

```{r plot-vpc-exact, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
exact_bins <- c(sort(unique(simout$NTIME)), Inf)

vpc_exact_ntime <- vpc(
  sim = simout, 
  obs = filter(simout, SIM == 1), 
  bins = exact_bins,
  sim_cols = list(dv = "SIMDV", idv = "NTIME", pred = "PRED"),
  obs_cols = list(dv = "OBSDV", idv = "NTIME", pred = "PRED"),
  pred_corr = TRUE,
  pi = c(0.05, 0.95),
  ci = c(0.05, 0.95),
  show = list(obs_dv = TRUE),
  log_y = TRUE,
  xlab = "Time (hours)",
  ylab = "PRED-corrected Concentration (ng/mL)"
)+
  scale_x_continuous(breaks = seq(0,168,24))

vpc_exact_ntime
```

However, this workaround only works if `idv` is set to the nominal time variable (`idv = "NTIME`) in both `sim_cols` and `obs_cols`, which removes the option to plot the observationsby actual time (`"TIME"`). 

`plot_vpc_exactbins()` gets around this limitation by plotting the observed data in a separate layer on top of the plot object returned by `vpc()`, including prediction-correction of those observed points if `pcvpc = TRUE` (also passed along to the `pred_corr` argument of `vpc()`).

```{r plot-exact, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
vpc_exact <- plot_vpc_exactbins(
  sim = simout, 
  pcvpc = TRUE,
  time_vars = c(TIME = "TIME", NTIME = "NTIME"),
  output_vars = c(PRED = "PRED", IPRED = "IPRED", SIMDV = "SIMDV", OBSDV = "OBSDV"),
  pi = c(0.05, 0.95),
  ci = c(0.05, 0.95),
  log_y = TRUE,
  xlab = "Time (hours)",
  ylab = "PRED-corrected Concentration (ng/mL)"
)+
  scale_x_continuous(breaks = seq(0,168,24))

vpc_exact
```

The difference in plotting the observed prediction-corrected data points by `"TIME"` versus `"NTIME"` is negligible for this example dataset, due to the high concordance between `"TIME"` and `"NTIME"` in this example Phase 1 study. However, this difference is often much larger for pooled analyses including later phase clinical studies where plotting the observed points versus actual time will result in a plot that is much more representative of the distribution of times in the model training dataset.

When exploring the bins in the input and output datasets using `plot_vpc_exactbins()`, we now see that they are consistent! Huzzah!

```{r bins-exact}
##Exact bins in the input data
df_nobsbin(data_sad, bin_var = "NTIME")

##Bin midpoints and boundaries determined by plot_vpc_exactbins()
distinct(select(vpc_exact$data, bin_mid, bin_min, bin_max))

```

`plot_vpc_exactbins()` also contains an argument built around a function we introduced earlier (`df_nobsbin()`). The argument `min_bin_count` (default = 1) filters out exact bins with fewer quantifiable observations than the minimum set by this argument. Importantly, the observed data points in these small bins are *still plotted*; however, they do not influence the calculation of summary statistics or summary plot elements (shaded intervals, lines). This provides the greatest fidelity to the data visualized without introducing visual artifacts due to small sample timepoints.

Additionally, because our plot input dataset `simout` was generated using `df_mrgsim_replicate`, the `time_vars` and `output_vars` match the default values, so we do not need to specify these arguments and they can be removed. Furthermore, the `pi` and `ci` arguments we are specifying match the default 90% PI and CI in `vpc()` and can be removed.

When setting`min_bin_count = 10`, summary statistics are not plotted for the final two timepoint containing only fewer than 10 quantifiable observations; however, the observations themselves are still plotted.

```{r plot-exact-min-bin-count, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
vpc_exact_bin_gt10 <- plot_vpc_exactbins(
  sim = simout, 
  pcvpc = TRUE,
  min_bin_count = 10,
  log_y = TRUE,
  xlab = "Time (hours)",
  ylab = "PRED-corrected Concentration (ng/mL)"
) +
  scale_x_continuous(breaks = seq(0,168,24))

vpc_exact_bin_gt10
```

We can also pass stratifying variables to the argument `strat_var` in order to facet our plots by relevant covariate conditions. The stratification variables specified in `strat_var` are also passed to the `stratify` argument of `vpc()`, in order to facet the resulting plots. Currently, only one variable can be passed to this argument.

```{r plot-exact-one-strata, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
vpc_exact_food <- plot_vpc_exactbins(
  sim = mutate(simout, FOOD_f = factor(FOOD, levels = c(0,1), labels = c("Fasted", "Fed"))), 
  strat_var = "FOOD_f",
  pcvpc = TRUE,
  log_y = TRUE,
  xlab = "Time (hours)",
  ylab = "PRED-corrected Concentration (ng/mL)"
) +
  scale_x_continuous(breaks = seq(0,168,24))

vpc_exact_food
```

