---
title: "VPC Plot Aesthetics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vpc-plot-aesthetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will review the functionality for updating aesthetic elements of VPC plots generated using `vpc()` with the `new_vpc_theme()` function  from the `vpc` package. 

`plot_legend()` is a helper plotting function that creates a legend for plots generated using `vpc()`. These legends can then be merged with the VPC plot into a single plot object using the `patchwork` package.

Let's get started. First, we will load the required packages.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(vpc, warn.conflicts =  FALSE)
library(mrgsolve, warn.conflicts =  FALSE)
library(withr, warn.conflicts =  FALSE)
library(patchwork, warn.conflicts = FALSE)
```

Next, let's load use the internal data and model objects from `pmxhelpr` and `df_mrgsim_replicate` to run the simulation.

```{r sim}
data <- data_sad
model <- model_mread_load("model")

simout <- df_mrgsim_replicate(data = data, model = model,replicates = 100, 
                     time_vars = c(TIME = "TIME", NTIME = "NTIME"),
                     output_vars = c(PRED = "PRED", IPRED = "IPRED", DV = "ODV"),
                     num_vars = c("CMT", "BLQ", "LLOQ", "EVID", "MDV", "DOSE", "FOOD"),
                     char_vars = c("PART"),
                     obsonly = TRUE)
```

Now let's plot all data together in a prediction-corrected VPC (pcVPC). There is only single dose administration in this dataset, thus we are able to pool across doses and food conditions with prediction-correction. We will set `min_bin_count = 5` to avoid the simulated intervals extending to the final timepoint with only a single observation.

```{r plot-vpc-pcvpc, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_pc <- plot_vpc_exactbins(
  sim = simout, 
  time_vars = c(TIME = "TIME", NTIME = "NTIME"),
  output_vars = c(PRED = "PRED", IPRED = "IPRED", SIMDV = "SIMDV", OBSDV = "OBSDV"),
  log_y = TRUE,
  xlab = "Time (hours)",
  ylab = "Concentration (ng/mL)",
  pcvpc = TRUE,
  min_bin_count = 5
)

vpc_pc
```

The default elements shown in the plot are inherited from `vpc()`. The `shown` argument can be provided to `plot_vpc_exactbins()` and passed on to the `show` argument of `vpc()`. 

The options are as follows:

```{r vpc-plot-shown}
shown_list <- list(obs_dv = TRUE, obs_ci = TRUE,
                   pi = FALSE, pi_as_area = FALSE, pi_ci = TRUE,
                   obs_median = TRUE, sim_median =FALSE, sim_median_ci = TRUE)
print(shown_list)
```

Similarly, the default aesthetics for this plot are inherited from the defaults in the `vpc` package, which can be viewed by running `new_vpc_theme()` with no arguments.

```{r vpc-theme-update}
vpc_theme_list <- new_vpc_theme()
print(vpc_theme_list)
```

Now, suppose we want to change the default aesthetics of the VPC plot. This can be accomplished by passing a named list of elements to update to the function `new_vpc_theme()`. The named list object generated from this function can then be passed to the `theme` argument in `vpc_plot_exactbins`, which is also an alias for the `vpc_theme` argument in `vpc()`.

Let's say we prefer a more classical VPC color schema, based only on basic red-blue-green color brewer. We will make the following changes:
  - Observed points blue
  - Observed median and simulated 90% CI of the median red 
  - Observed 5th and 95th and simulated 90% CI of the 5th and 95th blue

```{r vpc-theme-set}
vpc_theme <- new_vpc_theme(
  list(
  obs_color = rgb(0,0,1),
  
  obs_median_color = rgb(1,0,0),
  obs_ci_color = rgb(0,0,1),
  
  sim_median_fill = rgb(1,0,0),
  sim_pi_fill = rgb(0,0,1)
  )
)
```

Now, let's take a look at our plot prior plot with the new color schema.

```{r plot-vpc-pcvpc-new-theme, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_pc <- plot_vpc_exactbins(
  sim = simout, 
  time_vars = c(TIME = "TIME", NTIME = "NTIME"),
  output_vars = c(PRED = "PRED", IPRED = "IPRED", SIMDV = "SIMDV", OBSDV = "OBSDV"),
  log_y = TRUE,
  xlab = "Time (hours)",
  ylab = "Concentration (ng/mL)",
  pcvpc = TRUE,
  min_bin_count = 5
) 

vpc_pc
```

This is nice; however, I do not like the grey background and I would prefer only major y-axis grid lines without any x-axis grid lines. Conveniently, we can just add `theme()` elements on top of the plot object using the standard `ggplot2` approach.

```{r plot-vpc-pcvpc-new, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_pc_new <- plot_vpc_exactbins(
  sim = simout, 
  time_vars = c(TIME = "TIME", NTIME = "NTIME"),
  output_vars = c(PRED = "PRED", IPRED = "IPRED", SIMDV = "SIMDV", OBSDV = "OBSDV"),
  log_y = TRUE,
  xlab = "Time (hours)",
  ylab = "Concentration (ng/mL)",
  pcvpc = TRUE,
  min_bin_count = 5,
  theme = vpc_theme
) +
  theme(panel.background = element_rect(fill = "white", linewidth = 0.5, color = "black"),
        panel.grid.minor.y = element_blank(), 
        panel.grid.minor.x = element_blank(), 
        panel.grid.major.x = element_blank())

vpc_pc_new
```

Okay, now we have gotten the plot aesthetics where we want them; however, there is one other element we may like to include in the figure to make it more easily interpreted in isolation - a legend. 

`pmxhelpr` provides a useful helper function for this purpose, `plot_legend()`. The arguments to `plot_legend()` are as follows:
    - `ci` simulated confidence interval plotted. Default is c(0.05, 0.95).
    - `pi` prediction intervals plotted.  Default is c(0.05, 0.95).
    - `lloq` label for lower limit of quantification in the plot legend.
    - `update` list containing the plot elements to be updated. Default is set by `vpc::new_vpc_theme()`.
    - `shown` Named list of logicals specifying which layers to include on the plot. 
    
To obtain a legend for a plot using default aesthetics, simply run plot_legend() without any arguments specified.

```{r vpc_pc_legend, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_pc_legend <- plot_legend()
vpc_pc_legend
```

Now we have a `ggplot` object legend for our first plot!

To generate one for our second plot with updated aesthetics, let's pass the same named list object (`vpc_theme`) to the `update` argument.

```{r vpc_pc_new_legend, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_pc_new_legend <- plot_legend(update = vpc_theme)
vpc_pc_new_legend
```

Okay, now that we have our legend plot objects, let's combine them with the VPC plot objects into a single plot object with the `patchwork` package.

```{r plot-vpc-pcvpc-w-legend, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_pc_wleg <- vpc_pc + vpc_pc_legend + plot_layout(heights = c(2.5,1))
vpc_pc_wleg
```

```{r plot-vpc-pcvpc-new-w-legend, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_pc_new_wleg <- vpc_pc_new + vpc_pc_new_legend + plot_layout(heights = c(2.5,1))
vpc_pc_new_wleg
```

