---
title: "Visual Predictive Check Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vpc-workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will demonstrate `pmxhelpr` functions for simulation-based visual predictive check (VPC) and prediction-corrected visual predictive check (pcVPC) evaluation of models fit to concentration-time data.  

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(forcats, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(mrgsolve, warn.conflicts =  FALSE)
library(withr, warn.conflicts =  FALSE)
library(patchwork, warn.conflicts = FALSE)
```



# Data 

The example dataset used in this vignette (`data_sad`) is based on a single ascending dose (SAD) study of an orally administered drug product with a parallel group food effect (FE) cohort. This vignette will assume familiarity with the `data_sad` internal dataset and the `plot_dvtime()` function, as described in the [Exploratory Analyses of PK Data](eda-pk-workflow.html) vignette. These elements will not be reviewed in detail in this vignette.

## data_sad

Dataset definitions can be viewed by calling `?data_sad`. A quick summary of variables and types is printed below.

```{r data-glimpse}
glimpse(data_sad)
```

PART specifies the two study cohorts:

  - Single Ascending Dose (SAD)
  - Food Effect (FE).

```{r data-part}
unique(data_sad$PART)
```

This dataset also contains an exact binning variable:

  - Nominal Time (NTIME).

This variable represents the nominal time of sample collection relative to first dose per study protocol whereas Actual Time (TIME) represents the actual time the sample was collected.

```{r data-ntime}
##Unique values of NTIME
unique(data_sad$NTIME)

##Comparison of number of unique values of NTIME and TIME
length(unique(data_sad$NTIME))
length(unique(data_sad$TIME))
```

Let's define some variables that may be useful in plotting and filter down to PK relevant records.

```{r data-factors, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
data_sad_pk <- data_sad %>% 
  filter(CMT %in% c(1,2)) %>% 
  mutate(`Food Status` = ifelse(FOOD == 0, "Fasted", "Fed"), 
         DoseFood = paste(DOSE, "mg", `Food Status`)) %>% 
  mutate(`Food Status` = var_addn(`Food Status`, ID),
         `Dose and Food` = var_addn(DoseFood, ID)) %>% 
  mutate(`Dose and Food` = fct_relevel(`Dose and Food`, "50 mg Fasted (n=6)", after = 1))
```



# PK Model

An example PK model (`pkmodel`) in `mrgmod` format is provided in the internal package library. This can be loaded using the helper function `model_mread_load()`, which wraps `mrgsolve::mread`.

```{r model-pk}
model <- model_mread_load("pkmodel")
```

We can take a look at the model code using `mrgsolve::see`

```{r model-see}
mrgsolve::see(model)
```



# VPC Workflow

## Overview

 `pmxhelpr` includes a function for generating VPC and pcVPC plots for model evaluation: `plot_vpc_cont()`.

Many VPC tools use automatic binning algorithms (e.g., Jenks natural breaks, k-means, density) to determine bin intervals from the data. While these are useful for datasets without a pre-defined binning variable, they are not optimized to leverage input datasets with a variable representing exact bin times (e.g., nominal or protocol-specified times). Additionally, when a dataset contains an exact binning variable, these algorithms may not reproduce the exact bins. For example, absorption phase timepoints that are close together may be grouped into a single bin, obscuring important pharmacokinetic details.

The presence of exact bins in the data is a common scenario in pharmacometrics, as Clinical Study Protocols specify the times of PK sampling. `plot_vpc_cont()` uses the unique values of the nominal time variable directly as exact bins, ensuring that summary statistics are calculated at visualized at the protocol-specified sampling times.

 `plot_vpc_cont()` aesthetic control arguments are not reviewed in detail in this vignette. See the [Plot Themes and Aesthetics](plot-themes.html) vignette for details on customizing plot appearance.


## Running the simulation with `df_mrgsim_replicate()`

### Overview

`pmxhelpr` includes a function for running multiple replicates of a data set via simulation: `df_mrgsim_replicate()`

 `df_mrgsim_replicate()` is a wrapper function for `mrgsim_df()`, which uses `lapply()` to iterate the simulation over integers from 1 to the value passed to the argument `replicates`.

There are 3 required arguments to `df_mrgsim_replicate()`

  - `data`, data.frame modeling analysis dataset
  - `model`, `mrgmod` model object
  - `replicates`, numeric number of replicates to perform.
  
The are optional arguments specifying key dataset variables to be input into the simulation or captured in output. These include: 

  - `dv_var` = DV, dependent variable
  - `time_var` = TIME, actual time variable
  - `ntime_var` = NTIME, nominal time variable
  - `pred_var` = PRED, population prediction variable (fixed effects only)
  - `ipred_var` = IPRED, individual prediction variable (fixed + level 1 random effects)
  - `sim_dv_var` = DV, dependent variable captured in the simulated output (fixed + level 1 and 2 random effects)
  
These argument uses non-standard evaluation and can be passed as a bare column name or as a string.

We can pass `data_sad_pk` and `model`  and run the simulation for 100 `replicates`. The names of actual and nominal time variables in `data_sad` match the default arguments; however, our dependent variable is named `ODV`, which must be specified in the `dv_var` argument, since it differs from the default (`DV`). 

```{r df_mrgsim_replicate}
simout <- df_mrgsim_replicate(data = data_sad_pk,
                     model = model,
                     replicates = 100,
                     dv_var = ODV)

glimpse(simout)
max(simout$SIM)
```

The simulation has appended some key variables with standardized names that will allow us to take advantage of default arguments in later pipeline functions:
  - `PRED` (population prediction)
  - `IPRED` (individual prediction)
  - `SIMDV` (simulated dependent variable)
  - `OBSDV` (observed dependent variable).
  
These variable standard names are output regardless of the input name. For example, we can see from the `glimpse()` output that the simulation also outputs `"Y"`, which represents a simulated version of DV. We will still receive the same standardized output variable names if we change argument form `sim_dv_var`, as the variables are renamed to VPC workflow standards on output

```{r df_mrgsim_replicate-y}
simout <- df_mrgsim_replicate(data = data_sad_pk,
                     model = model,
                     replicates = 100,
                     dv_var = ODV, 
                     sim_dv_var = Y)

glimpse(simout)
```

### Changing the Replicate Counter

The maximum value of our replicate count variable (default = `"SIM"`) indicates that the dataset has been replicated 100 times. The variable name output for the count of replicates can be specified using the `irep_name` argument.

This argument uses non-standard evaluation and can be supplied as a bare name or string

```{r df_mrgsim_replicate-irep}
simout <- df_mrgsim_replicate(data = data_sad_pk,
                     model = model,
                     replicates = 100,
                     dv_var = ODV, 
                     irep_name = IREP)

max(simout$IREP)
```


### Capturing Additional Variables
  
The user may want to recover additional variables from the input dataset that may be needed to stratify or pre-process the simulation output for VPC plots. 

`df_mrgsim_replicate` includes two arguments for this:

  - `num_vars` character vector of variable names passed to `carry_out` argument of `mrgsim_df()`. 
  - `char_vars` character vector of variable names passed to `recover` argument of `mrgsim_df()`
  
*NOTE* CMT, EVID, MDV, TIME, and NTIME are allways carried out automatically and returned

```{r df_mrgsim_replicate-capture}
simout <- df_mrgsim_replicate(data = data_sad_pk,
                     model = model,
                     replicates = 100,
                     dv_var = ODV, 
                     num_vars = c("DOSE", "FOOD", "BLQ", "LLOQ"), 
                     char_vars = c("PART", "DoseFood"))

colnames(simout)
```

### Setting the seed

The unique seed can be set using the `seed` argument. The default is `1234567889`.

```{r df_mrgsim_replicate-seed}
simout <- df_mrgsim_replicate(data = data_sad_pk,
                     model = model,
                     replicates = 100,
                     dv_var = ODV, 
                     seed = 42)

glimpse(simout)
```

### Additional arguments passed to `mrgsim_df()`

Additional arguments can be passed to `mrgsim_df()` via the elipses `...`. One argument that may sometimes be useful is `obsonly = TRUE`. This argument removes dose records from the simulation output and reduce file size.

```{r df_mrgsim_replicate-obsonly}
simout <- df_mrgsim_replicate(data = data_sad_pk,
                     model = model,
                     replicates = 100,
                     dv_var = ODV, 
                     num_vars = c("DOSE", "FOOD", "BLQ", "LLOQ"), 
                     char_vars = c("PART", "DoseFood"))

nrow(simout)

simout <- df_mrgsim_replicate(data = data_sad_pk,
                     model = model,
                     replicates = 100,
                     dv_var = ODV, 
                     num_vars = c("DOSE", "FOOD", "BLQ", "LLOQ"), 
                     char_vars = c("PART", "DoseFood"), 
                     obsonly = TRUE)

nrow(simout)
```


## VPC Plots with `plot_vpc_cont()`

`pmxhelpr` contains a function to produce VPC and pcVPC plots for models fit to longitudinal, continuous repeated measures data: `plot_vpc_cont()` 

This function is designed around the concept of actual and nominal time variables, with the latter used to bin data together to calculate and display summary statistics across simulation replicates.

`plot_vpc_cont` has 1 required argument: `sim`, the input simulated dataset for the VPC procedure.

```{r plot-vpc, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_vpc_cont(
  sim = simout,
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_x_continuous(breaks = seq(0,168,24))+
  scale_y_log10(guide = "axis_logticks")
```

A decent, if slightly misleading, plot! Simulated 90% (`ci` = c(0.05, 0.95)) confidence intervals print at the median (50th) and  5th and 95th percentiles (`pi` = c(0.05, 0.95)) and corresponding quantiles of the observed are shown as lines with observed data points overlaid.

`plot_vpc_cont` censors quantiles of the observed data displayed as lines when the corresponding quantile of observations is BLQ. This is done automatically if the variable `LLOQ` is present in the input simulated dataset `sim`. This behavior is suppressed when prediction-correction is applied with `pcvpc = TRUE`.

### Prediction-correction

Because this study includes multiple dose levels, it is not appropriate to visualize the data together in a single VPC plot. One must either: 1) Facet by dose, 2) prediction-correct 

```{r plot-vpc-pc, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_vpc_cont(
  sim = simout,
  pcvpc = TRUE
) +
  labs(x = "Time (hours)", y = "Pred-corrected Conc. (ng/mL)") +
  scale_x_continuous(breaks = seq(0,168,24))+
  scale_y_log10(guide = "axis_logticks")
```

Note that the overlay of the LLOQ is now suppressed with prediction-correction.

### Stratifying plots

The other option to prediction-correction is to stratify by confounding factors. Dose is generally the minimum stratification for a VPC plot. The variable for stratification *MUST* be passed to the argument `strat_var` and not added later only as a `facet_wrap`/`facet_grid` call to the returned object.

Let's take a look at our prediction-corrected VPC stratified by Food Status.

```{r plot-vpc-pc-food, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_vpc_cont(
  sim = mutate(simout,FoodStatus = ifelse(FOOD == 1, "Fed", "Fasted")),
  pcvpc = TRUE,
  strat_var = FoodStatus
) +
  labs(x = "Time (hours)", y = "Pred-corrected Conc. (ng/mL)") +
  scale_x_continuous(breaks = seq(0,168,24))+
  scale_y_log10(guide = "axis_logticks")

```

Now let's look at a regular VPC stratified by the interaction of dose and food status

```{r plot-vpc-dosefood, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_vpc_cont(
  sim = simout,
  strat_var = DoseFood
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_x_continuous(breaks = seq(0,168,24))+
  scale_y_log10(guide = "axis_logticks")
```

*NOTE* Because `plot_vpc_cont` performs BOTH the summary statistic calculation and the plotting, one cannot simply add a `facet_wrap` layer with the stratification after calling `plot_vpc_cont`. This will result in the binned observed quantiles and simulated intervals plotted reflecting the unstratified data across all conditions with only the observed points stratified.

```{r plot-vpc-dosefood-facet, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_vpc_cont(
  sim = simout
) +
  facet_wrap(~DoseFood)+
  labs(x = "Time (hours)", y = "Pred-corrected Conc. (ng/mL)") +
  scale_x_continuous(breaks = seq(0,168,24))+
  scale_y_log10(guide = "axis_logticks")
```

### Dropping small sample bins

There may be nominal time bins that contain only a small number of observations that skew the simulated intervals. `plot_vpc_cont()` includes the argument `min_bin_count` (default = 1), which filters out exact bins with fewer quantifiable observations than the minimum set by this argument. Importantly, the observed data points in these small bins are *still plotted*; however, they do not influence the calculation of summary statistics or summary plot elements (shaded intervals, lines). This provides the greatest fidelity to the data visualized without introducing visual artifacts due to small sample timepoints.

When setting`min_bin_count = 10`, summary statistics are not plotted for the final two timepoint containing only fewer than 10 quantifiable observations; however, the observations themselves are still plotted.

```{r plot-vpc-min-bin-count, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_vpc_cont(
  sim = simout,
  pcvpc = TRUE,
  min_bin_count = 10
) +
  labs(x = "Time (hours)", y = "PRED-corrected Concentration (ng/mL)") +
  scale_x_continuous(breaks = seq(0,168,24))+
  scale_y_log10(guide = "axis_logticks")
```

### BLQ handling

BLQ data are often ignored in population PK modeling (i.e., set as missing; M1 method) in population. This is unlikely to introduce appreciable bias in model parameter estimation, provided the absolute frequency of BLQ data is low (e.g., <10-20%) and the pattern of BLQ censoring is in line with the expectations by dose and time.

However, the handling of BLQ data in VPCs may impact the interpretation of these graphical model diagnostics.

Let's generate some VPC plots to demonstrate the impact of different approaches to BLQ handling to the visual assessment of the adequacy of model fit and the true underlying PK profile. We will continue using the `simout` object generated above.

Let's start by focusing on evaluating the model fit of the 100 mg dose level we have been exploring across Part1-SAD and Part2-FE.

```{r sim-input}
sim100 <- simout %>%
  filter(DOSE == 100)
```

There are two potential approaches processing and plotting the VPC:

  1. `Exclude BLQ`: remove missing observations (`MDV=1`) in the observed input data before passing to the simulation (e.g., filter out `MDV=1 and EVID=0`)
  2. `Censor Observed Quantiles`: simulate all timepoints and set quantiles of the observed data to `NA`, if the sample representing that quantile of the observed is BLQ (`MDV=1`) in the bin.

Let's plot using the default arguments to `plot_vpc_cont()`. The default behavior is `Exclude BLQ`, which filters out `MDV=1`.

```{r plot-vpc-blq-drop, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
vpc_blq_drop <- plot_vpc_cont(
  sim = filter(sim100, MDV==0),
  pcvpc = TRUE,
  strat_var = PART
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_x_continuous(breaks = seq(0,168,24))+
  scale_y_log10(guide = "axis_logticks")

vpc_blq_drop
```

The observed median appears consistent with the exploratory concentration-time profiles generated earlier excluding data below the lower limit of quantification (i.e., plotting `"ODV"`). Notice that the increasing trend in the observed quantiles at later timelines coincides with completely overlapping simulated confidence intervals for all quantiles. This is due to the decreasing sample size with time due to exclusion of timepoints where `MDV=1`.

Let's instead try specifying a new argument `loq` to `plot_vpc_cont()`.

`loq` is the numeric value of the lower limit of quantification in the units of the dependent variable. When specified, observed quantiles are computed using censored quantile estimation, where values below the LLOQ (including missing values) are treated as left-censored. If the quantile of the observed data falls below the LLOQ, it is returned as `NA`.

```{r plot-vpc-blq-obs-cens, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
vpc <- plot_vpc_cont(
  sim = sim100,
  strat_var = PART,
  loq = 1
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_x_continuous(breaks = seq(0,168,24))+
  scale_y_log10(guide = "axis_logticks")

vpc
```

Now we see a red horizontal line depicting the LLOQ (1 ng/mL). Notice also the increase in the y-axis range, the change in the shape of the observed quantiles, and the greater resolution in separation between the confidence intervals.

Specifying `loq` assumes that `MDV=1` represents *only* BLQ samples in the input dataset for the plot (`sim`) when `EVID=0`. The function sets the `MDV` variable in `sim` to zero (0) so that all timepoints are included.

  - In the calculation of the summary statistics of the observed data, the quantiles are calculated across all data per bin and taken as `NA` when that quantile of observations is below `loq`.
  - In the calculation of summary statistics of the simulated data, the simulated confidence intervals are based on all data without any censoring, which is representative of the model-predicted "true" underlying PK profile in the absence of real-world assay limitations.

Specifying `loq` is the preferred method for plotting VPC diagnostics with the `pmxhelpr` package. This is not the default method due to the all-too-common case where the assay LLOQ is not known to the analyst. 

Additionally, this method of accounting for the censoring of data below the LLOQ by specifying the `loq` argument should only be applied to Visual Predictive Checks (`pcvpc = FALSE`) and should *not* be applied to Prediction-corrected Visual Predictive Checks (`pcvpc = TRUE`).

### Adjusting the elements displayed

The default elements shown in the plot are controlled by the `shown` argument in `plot_vpc_cont()`.

The default is as follows:

```{r vpc-plot-shown}
plot_vpc_shown()
```

The components of the list correspond to the following vpc plot elements:

-   Observed points: `obs_point`
-   Observed quantile lines: `obs_pi_line`
-   Observed median line: `obs_med_line`
-   Simulated prediction interval lines: `sim_pi_line`
-   Simulated prediction interval CI: `sim_pi_ci`
-   Simulated prediction interval area: `sim_pi_area`
-   Simulated median line: `sim_median_line`
-   Simulated median CI: `sim_median_ci`

One or more elements to be updated from the defaults above can be passed as a list to the argument `shown`. Any elements not specified in `shown` will inherit the defaults.

For example, the 90% prediction interval (i.e., 5th to 95th percentiles) can be visualized in place of CIs of each of the 5th and 95th percentiles with the observed and simulated medians shown as follows:

```{r plot-vpc-pcvpc-90pi, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
plot_vpc_cont(
  sim = simout,
  pcvpc = TRUE,
  min_bin_count = 5,
  shown = plot_vpc_shown(obs_pi_line = FALSE, sim_pi_ci = FALSE, sim_median_ci = FALSE, sim_median_line = TRUE, sim_pi_area = TRUE)
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_x_continuous(breaks = seq(0,168,24))+
  scale_y_log10(guide = "axis_logticks")
```

Now, let's say we want to remove the observed data points from the plot above to better visualize the observed quantile lines relative to their corresponding simulated confidence intervals. This can be accomplished with `shown`.

```{r plot-vpc-pcvpc-noobs, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
 plot_vpc_cont(
  sim = simout,
  pcvpc = TRUE,
  min_bin_count = 5,
  shown = plot_vpc_shown(obs_point = FALSE)
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_x_continuous(breaks = seq(0,168,24))+
  scale_y_log10(guide = "axis_logticks")
```

We could also take this one step further and only look at the median and the simulated confidence interval of the median, to closely interrogate central tendency. This is common for VPC strata which have few observations, leading to inadequate sample size to discriminate between the confidence intervals of the median and the extremes. This is common scenario when evaluating VPC plots stratified by individual study arms in early phase trials.

```{r plot-vpc-pcvpc-noobs-medonly, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
 plot_vpc_cont(
  sim = simout,
  pcvpc = TRUE,
  min_bin_count = 5,
  shown = plot_vpc_shown(obs_point = FALSE, obs_pi_line = FALSE, sim_pi_ci = FALSE)
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_x_continuous(breaks = seq(0,168,24))+
  scale_y_log10(guide = "axis_logticks")
```

### Adjusting the Plot Theme with `plot_vpc_theme()`

The default aesthetics for VPC plots are controlled via `plot_vpc_theme()`. See the [Plot Themes and Aesthetics](plot-themes.html) vignette for details on the theme system, element constructors, and examples of customizing VPC aesthetics.

```{r vpc-theme}
plot_vpc_theme()
```

We can define an alternative theme using a blue / grey color schema using this function and the constructor helpers.

```{r vpc-new-theme}
vpc_new_theme <- plot_vpc_theme(
  obs_point = pmx_point(color = "#000000"),
  obs_med_line = pmx_line(color = "#000000"),
  obs_pi_line = pmx_line(color = "#000000"),
  sim_median_ci = pmx_ribbon(fill = "#3388cc"),
  sim_pi_ci = pmx_ribbon(fill = "#3388cc")
)
```

Regenerating our BLQ quantile censored plot stratified by part with the new theme yields the following plot

```{r plot-vpc-blq-obs-cens2, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
vpc2 <- plot_vpc_cont(
  sim = sim100,
  strat_var = PART,
  loq = 1,
  theme = vpc_new_theme
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_x_continuous(breaks = seq(0,168,24))+
  scale_y_log10(guide = "axis_logticks")

vpc2
```

### Accessing VPC Summary Statistics

By default, `plot_vpc_cont()` returns a `ggplot` object. To access the underlying summary statistics used to build the plot, set `vpcstats = TRUE`. This returns a `data.frame` containing the observed and simulated quantiles, confidence intervals, and bin information.

```{r vpc-stats, message=FALSE, warning=FALSE}
stats <- plot_vpc_cont(
  sim = simout,
  strat_var = DoseFood,
  vpcstats = TRUE
)

head(stats)
```


### Modifying Axes and Labels

Because `plot_vpc_cont()` returns a `ggplot2` object, the y-axis can be transformed to log10 scale by adding `scale_y_log10()` as a layer. This approach avoids any data transformation prior to calculating summary statistics, ensuring all quantile calculations are performed on the original scale.

Similarly, the axis labels can be controlled with `labs` and the x-axis breaks can be set to more reasonable values for time units using `scale_x_continuous()`.



## VPC Plot Legends with `plot_vpc_legend()`

This section will review the functionality for generating legends for VPC plots with `plot_vpc_legend()`

`plot_vpc_legend()` is a helper plotting function that creates a legend for VPC plots. These legends can then be merged with the VPC plot into a single plot object using the `patchwork` package.

### Defaults Legend

To obtain a legend for a plot using default aesthetics, simply run `plot_vpc_legend()` without any arguments specified.

```{r vpc-legend, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_legend <- plot_vpc_legend()
vpc_legend
```

The legend can then be combined with the ggplot object returned from `plot_vpc_cont` into a single plot object with the `patchwork` package.

```{r plot-vpc-pcvpc-w-legend, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_wleg <- vpc + vpc_legend + plot_layout(heights = c(2.5,1))
vpc_wleg
```

### Updating the Legend Elements

The legend can be updated to remove or add elements based on `shown` argument common to `plot_vpc_cont` and `plot_vpc_legend`. The defaults list can be viewed and updated globally to be passed to all plots in a VPC workflow using `plot_vpc_shown()`

```{r vpc-legend-medonly}
plot_vpc_legend(shown = plot_vpc_shown(obs_pi_line = FALSE, sim_pi_ci = FALSE))
```

```{r vpc-legend-shown-medonly}
shown_elements <- plot_vpc_shown(obs_pi_line = FALSE, sim_pi_ci = FALSE)
plot_vpc_legend(shown = shown_elements)
```

### Updating the Legend Theme

The legend can also be updated with the updated VPC plot theme elements. This is easiest to do by setting a new theme object, which we did previously when we created `vpc_new_theme` with `plot_vpc_theme()`. We can use this same theme list object to generat the VPC plot with `plot_vpc_cont()` and the VPC legend with `plot_vpc_legend()`.

```{r vpc-legend2, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_new_legend <- plot_vpc_legend(update = vpc_new_theme)
vpc_new_legend
```

The legend can then be combined with the ggplot object returned from `plot_vpc_cont` into a single plot object with the `patchwork` package.

```{r plot-vpc-pcvpc-w-legend2, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_wleg2 <- vpc2 + vpc_new_legend + plot_layout(heights = c(2.5,1))
vpc_wleg2
```
