---
title: "VPC Plot Aesthetics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vpc-plot-aesthetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette will review the functionality for updating aesthetic elements of VPC plots generated using `plot_vpc_exactbins()` with the `plot_vpc_theme()` function.

`plot_vpclegend()` is a helper plotting function that creates a legend for VPC plots. These legends can then be merged with the VPC plot into a single plot object using the `patchwork` package.

Let's get started. First, we will load the required packages.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(mrgsolve, warn.conflicts =  FALSE)
library(withr, warn.conflicts =  FALSE)
library(patchwork, warn.conflicts = FALSE)
```

Next, let's load use the internal data and model objects from `pmxhelpr` and `df_mrgsim_replicate` to run the simulation.

```{r sim}
data <- data_sad %>%
  filter(CMT %in% c(1,2))
model <- model_mread_load("pkmodel")

simout <- df_mrgsim_replicate(data = data, model = model,replicates = 100,
                     dv_var = "ODV",
                     time_vars = c(TIME = "TIME", NTIME = "NTIME"),
                     output_vars = c(PRED = "PRED", IPRED = "IPRED", DV = "DV"),
                     num_vars = c("CMT", "BLQ", "LLOQ", "EVID", "MDV", "DOSE", "FOOD"),
                     char_vars = c("PART"),
                     obsonly = TRUE)
```

Now let's plot all data together in a prediction-corrected VPC (pcVPC). There is only single dose administration in this dataset, thus we are able to pool across doses and food conditions with prediction-correction. We will set `min_bin_count = 5` to avoid the simulated intervals extending to the final timepoint with only a single observation.

```{r plot-vpc-pcvpc, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_pc <- plot_vpc_exactbins(
  sim = simout,
  pcvpc = TRUE,
  min_bin_count = 5
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_y_log10(guide = "axis_logticks")

vpc_pc
```

# Plot Elements

## Default Elements

The default elements shown in the plot are controlled by the `shown` argument in `plot_vpc_exactbins()`.

The default is as follows:

```{r vpc-plot-shown}
plot_vpc_shown()
```

The components of the list correspond to the following vpc plot elements:

-   Observed points: `obs_dv`
-   Observed quantiles: `obs_ci`
-   Simulated inter-quantile range: `pi`
-   Simulated inter-quantile area: `pi_as_area`
-   Simulated Quantile CI: `pi_ci`
-   Observed Median: `obs_median`
-   Simulated Median: `sim_median`
-   Simulated Median CI: `sim_median_ci`

One or more elements to be updated from the defaults above can be passed as a list to the argument `shown`. Any elements not specified in `shown` will inherit the defaults.

## Adjusting Elements

For example, we may want to visualize the 90% prediction interval (i.e., 5th to 95th percentiles) derived from the simulation, rather than the confidence intervals of the 5th and 95th percentiles independently. In this case, we want to also see how close the simulated median falls relative to the observed median.

Let's see how this can be down using `shown`!

```{r plot-vpc-pcvpc-90pi, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}

vpc_pc_pi <- plot_vpc_exactbins(
  sim = simout,
  time_vars = c(TIME = "TIME", NTIME = "NTIME"),
  output_vars = c(PRED = "PRED", IPRED = "IPRED", SIMDV = "SIMDV", OBSDV = "OBSDV"),
  pcvpc = TRUE,
  min_bin_count = 5,
  shown = plot_vpc_shown(obs_ci = FALSE, pi_ci = FALSE, sim_median_ci = FALSE, sim_median = TRUE, pi_as_area = TRUE)
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_y_log10(guide = "axis_logticks")

vpc_pc_pi
```

Now, let's say we want to remove the observed data points from the plot above to better visualize the observed quantile lines relative to their corresponding simulated confidence intervals. We can do this as follows.

```{r plot-vpc-pcvpc-noobs, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}

vpc_pc_noobs <- plot_vpc_exactbins(
  sim = simout,
  time_vars = c(TIME = "TIME", NTIME = "NTIME"),
  output_vars = c(PRED = "PRED", IPRED = "IPRED", SIMDV = "SIMDV", OBSDV = "OBSDV"),
  pcvpc = TRUE,
  min_bin_count = 5,
  shown = plot_vpc_shown(obs_dv = FALSE)
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_y_log10(guide = "axis_logticks")

vpc_pc_noobs
```

We could also take this one step further and only look at the median and the simulated confidence interval of the median, to closely interrogate central tendency. This is common for VPC strata which have few observations, leading to inadequate sample size to discriminate between the confidence intervals of the median and the extremes. This is common scenario when evaluating VPC plots stratified by individual study arms in early phase trials.

```{r plot-vpc-pcvpc-noobs-medonly, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}

vpc_pc_noobs_medonly <- plot_vpc_exactbins(
  sim = simout,
  time_vars = c(TIME = "TIME", NTIME = "NTIME"),
  output_vars = c(PRED = "PRED", IPRED = "IPRED", SIMDV = "SIMDV", OBSDV = "OBSDV"),
  pcvpc = TRUE,
  min_bin_count = 5,
  shown = plot_vpc_shown(obs_dv = FALSE, obs_ci = FALSE, pi_ci = FALSE)
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_y_log10(guide = "axis_logticks")

vpc_pc_noobs_medonly
```

# Plot Aesthetics

## Default Aesthetics

Similarly, the default aesthetics for vpc plots in pmxhelpr use the classical red-blue-green brewer color schema. The defaults for `pmxhelpr` can be visualized by specifying the `plot_vpc_theme()` function with no arguments.

```{r vpc-theme-pmxhelpr}
pmxhelpr_theme_list <- plot_vpc_theme()
print(pmxhelpr_theme_list)
```

## Adjusting Aesthetics

Now, suppose we want to change the default aesthetics of the VPC plot, without changing what is being shown. This can be accomplished by passing element overrides to `plot_vpc_theme()`. The theme object generated from this function can then be passed to the `vpc_theme` argument in `plot_vpc_exactbins`.

Let's say we prefer a grey-blue color scheme. We can customize it by passing element constructors to `plot_vpc_theme()`:

```{r plot-vpc-customtheme, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_pc_customtheme <- plot_vpc_exactbins(
  sim = simout,
  time_vars = c(TIME = "TIME", NTIME = "NTIME"),
  output_vars = c(PRED = "PRED", IPRED = "IPRED", SIMDV = "SIMDV", OBSDV = "OBSDV"),
  pcvpc = TRUE,
  vpc_theme = plot_vpc_theme(obs = pmx_vpc_point(color = "#000000"),
                            obs_median = pmx_vpc_line(color = "#000000"),
                            obs_ci = pmx_vpc_line(color = "#000000"),
                            sim_median = pmx_vpc_ribbon(fill = "#3388cc"),
                            sim_pi = pmx_vpc_ribbon(fill = "#3388cc")),
  min_bin_count = 5
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_y_log10(guide = "axis_logticks")

vpc_pc_customtheme
```

## Adding Layers

Let's say we would like to visualize the major x-axis grid lines to help locate data corresponding to each sampling timepoint. Conveniently, `plot_vpc_exactbins` returns a `ggplot2` plot object, which we can modify by adding layers just like like any other `ggplot2` object.

```{r plot-vpc-pcvpc-xgrid, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_pc_xgrid <- plot_vpc_exactbins(
  sim = simout,
  time_vars = c(TIME = "TIME", NTIME = "NTIME"),
  pcvpc = TRUE,
  min_bin_count = 5,
  vpc_theme = plot_vpc_theme(obs = pmx_vpc_point(color = "#000000"),
                            obs_median = pmx_vpc_line(color = "#000000"),
                            obs_ci = pmx_vpc_line(color = "#000000"),
                            sim_median = pmx_vpc_ribbon(fill = "#3388cc"),
                            sim_pi = pmx_vpc_ribbon(fill = "#3388cc"))
) +
  labs(x = "Time (hours)", y = "Concentration (ng/mL)") +
  scale_y_log10(guide = "axis_logticks") +
  theme(panel.grid.major.x = element_line())

vpc_pc_xgrid
```

# VPC Plot Legends

## Defaults Legend

Okay, now we have gotten the plot aesthetics where we want them; however, there is one other element we may like to include in the figure to make it more easily interpreted in isolation - a legend.

`pmxhelpr` provides a useful helper function for this purpose, `plot_vpclegend()`. To obtain a legend for a plot using default aesthetics, simply run `plot_vpclegend()` without any arguments specified.

```{r vpc_pc_legend, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_pc_legend <- plot_vpclegend()
vpc_pc_legend
```

Now we have a `ggplot` object legend for our first plot!

To generate one with updated aesthetics to match the custom theme, let's pass the named list to the `update` argument.

```{r vpc_pc_new_legend, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_pc_new_legend <- plot_vpclegend(update = list(obs_color = "#000000",
                                                   obs_median_color = "#000000",
                                                   obs_ci_color = "#000000",
                                                   sim_median_fill = "#3388cc",
                                                   sim_pi_fill = "#3388cc"))
vpc_pc_new_legend
```

Okay, now that we have our legend plot objects, let's combine them with the VPC plot objects into a single plot object with the `patchwork` package.

```{r plot-vpc-pcvpc-w-legend, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_pc_wleg <- vpc_pc + vpc_pc_legend + plot_layout(heights = c(2.5,1))
vpc_pc_wleg
```

```{r plot-vpc-pcvpc-customtheme-w-legend, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_pc_new_wleg <- vpc_pc_xgrid + vpc_pc_new_legend + plot_layout(heights = c(2.5,1))
vpc_pc_new_wleg
```
