---
title: "Dose-Proportionality Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dose-Proportionality Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

This vignette demonstrates the `pmxhelpr` dose-proportionality assessment workflow using power law (log-log) regression of exposure versus dose. 

The pipeline mirrors the [VPC workflow](vpc-workflow.html): a stats function (`df_doseprop()`) returns a class-tagged object, a public renderer (`plot_build_doseprop()`) draws plots from that object, and a top-level wrapper (`plot_doseprop()`) bundles both for one-shot use.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
```

# Background

Dose proportionality assessment evaluates whether exposure increases proportionally with dose. The standard approach is power law regression on the relationship

$$Exposure = \alpha \cdot DOSE^\gamma$$

which linearizes via log transformation to

$$\log(Exposure) = intercept + \gamma \cdot \log(DOSE)$$

Hypothesis testing proceeds via the confidence interval around the power estimate ($\gamma$): the null is dose-proportional ($\gamma = 1$); the alternative is non-proportional ($\gamma \ne 1$). Interpretation rules of thumb:

  + CI includes 1: dose-proportional
  + CI excludes 1, upper bound below 1: less than dose-proportional
  + CI excludes 1, lower bound above 1: greater than dose-proportional

This assessment is typically performed on both Cmax and AUC. The combination informs which phase of the PK profile is contributing most to non-linearity (absorption rate, absorption extent, or elimination). 

# Data

This vignette uses the internal `data_sad_nca` dataset, which is the NCA parameter output for the `data_sad` SAD + food-effect study. The dataset is formatted in the SDTM `PP` (Pharmacokinetic Parameters) domain conventions, with one row per subject per parameter.

```{r data-glimpse}
glimpse(data_sad_nca)
```

For dose-proportionality assessment we want to focus on the SAD cohort and exclude the food-effect arm — food can also influence exposure and may confound dose-proportionality interpretation.

```{r data-prep}
data_nca1 <- filter(data_sad_nca, PART == "Part 1-SAD")
unique(data_nca1$PART)
```


# Computing summary statistics with `df_doseprop()`

The required arguments are:

  + `data` — a `data.frame` of NCA parameters (typically `PKNCA::pk.nca()` output)
  + `metrics` — character vector of exposure metric names to evaluate
  
```{r df-doseprop-obj}
dose_prop_obj <- df_doseprop(data_nca1, metrics = c("aucinf.obs", "cmax"))
dose_prop_obj
```
 
Optional arguments must be specified if input dataset defaults differ from SDTM formatting and `PKNCA` defaults:

  + `metric_name_var`: variable in `data` containing metric names requested in `metrics`, default `PPTESTCD`.
  + `metric_value_var`: variable in `data` containing exposure metric values, default `PPORRES`.
    
The default argument for the numeric dose variable `dose_var` is `DOSE`.

```{r df-doseprop-stat}
stats <- dose_prop_obj$stats
stats
```

The `stats` slot holds the per-metric estimates from the log log regression, which contains all the information needed to format an output table.

## Adjusting the summary statistics

The summary statistics calculated from the log-log regression of exposure versus dose can be customized using the following arguments:

  + `method`: distribution used to calculate the confidence interval: Default is `"normal"`. Alternative is `"tdist"`, which may be preferred for analysis with a small sample size where normality cannot be assumed (e.g, < 30).
  + `ci`: confidence interval to return expressed as a proportion. Default is `"0.9"` based on the CI recommended for the bioequivalence assessment.
  + `sigdigits`: number of significant digits to return in outputs
  
The example below modifies the previous calculation by requesting the `tdist` method, 95%, and rounding to only 2 significant digits.
  
```{r df-doseprop-obj-95ci-tdist}
dose_prop_obj_95ci_tdist <- df_doseprop(data_nca1, 
                                        metrics = c("aucinf.obs", "cmax"), 
                                        method = "tdist", 
                                        ci = 0.95, 
                                        sigdigits = 2)
dose_prop_obj_95ci_tdist
```

# Plotting with `plot_doseprop()`

`plot_doseprop()` is the top-level wrapper that bundles `df_doseprop()` and `plot_build_doseprop()`. The required arguments are the same as `df_doseprop()`. The output is a `ggplot` object with one panel per metric, where the facet label includes the metric name and the `PowerCI` text.

```{r plot-raw, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_doseprop(data_nca1, metrics = c("aucinf.obs", "cmax"))
```

The default appearance uses open circles at `alpha = 0.7` for the observation points and a black regression line with a grey SE ribbon. This matches the package design language for exploratory plots. To restore filled black circles with full opacity, override `obs_point` via the `theme` argument (see [Theming](#theming) below).

## Adjusting the summary statistics

The summary statistics calculated from the log-log regression of exposure versus dose can be customized using the following arguments:

  + `method`: distribution used to calculate the confidence interval: Default is `"normal"`. Alternative is `"tdist"`, which may be preferred for analysis with a small sample size where normality cannot be assumed (e.g, < 30).
  + `ci`: confidence interval to return expressed as a proportion. Default is `"0.9"` based on the CI recommended for the bioequivalence assessment.
  + `sigdigits`: number of significant digits to return in outputs
 
```{r plot-doseprop-obj-95ci-tdist, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_doseprop(data_nca1, 
              metrics = c("aucinf.obs", "cmax"),
              method = "tdist", 
              ci = 0.95, 
              sigdigits = 2)

```

## Reusing precomputed stats

`plot_doseprop()` also accepts a `doseprop_stats` object directly. This skips the regression refit and is useful when you want to inspect the stats object and then render the same data multiple times — for example, with and without the SE ribbon, or with different aesthetic overrides — without paying the regression cost again.

```{r plot-precomputed, fig.height=4, fig.width=6, warning=FALSE}
# With the SE ribbon (default)
plot_doseprop(dose_prop_obj)

# Without the SE ribbon
plot_doseprop(dose_prop_obj, se = FALSE)
```

When `data` is a precomputed `doseprop_stats` result, the pipeline arguments (`metrics`, `metric_name_var`, `metric_value_var`, `dose_var`, `method`, `ci`, `sigdigits`) **cannot** be honored — those decisions were already made when `df_doseprop()` ran, and re-passing them on the cached path would silently shadow the original values. Passing any of them aborts with a message pointing the caller back at `df_doseprop()`. Only plot-only arguments (`theme`, `se`) are accepted on this path; to change a regression setting, re-run `df_doseprop()` and pass the new result.



# Inspecting the `doseprop_stats` object

`df_doseprop()` performs the per-metric log-log regression and returns a `doseprop_stats`  [`pmx_stats`][is_pmx_stats] object with three slots:

  + `stats` (one row per metric), 
  + `obs` (the observation rows used for the scatter overlay)
  + `config` (the regression configuration). Downstream plot builders read directly from these slots to render without a refit.

The returned object is class-tagged `c("doseprop_stats", "pmx_stats")`, which provides three S3 methods designed for interactive inspection.

`print()` shows a focused summary — object dimensions, the regression configuration values (`metric_name_var`, `metric_value_var`, `dose_var`, `ci`, `method`), the count of observation rows attached for the scatter overlay, and the per-metric stats body.

```{r print-stats}
print(dose_prop_obj)
```

`summary()` shares the header but condenses the body to one line per metric using the `PowerCI` and `Interpretation` columns. This is the most compact way to scan the assessment.

```{r summary-stats}
summary(dose_prop_obj)
```

`as.data.frame()` returns the `stats` slot as a plain `data.frame` — useful when downstream tooling expects a flat tabular result. The observation overlay and run configuration remain accessible via `dose_prop_obj$obs` and `dose_prop_obj$config`.

```{r as-data-frame-stats}
plain <- as.data.frame(dose_prop_obj)
class(plain)
dose_prop_obj$config$metric_name_var
```

Use `is_doseprop_stats()` to test class membership programmatically — for example, in user code that should accept either a precomputed `doseprop_stats` object or a raw NCA data.frame. Pass `strict = TRUE` to additionally validate structural integrity.

```{r predicate-stats}
is_doseprop_stats(dose_prop_obj)
is_doseprop_stats(plain)  # FALSE -- coerced to a plain data.frame
```


# Styling

Plot aesthetics are customized via the `style` argument and the `style_doseprop()` preset, which returns a `ggstylekit::style_spec()`. The relevant roles are:

  + `obs_point` — observation point aesthetics
  + `linear` — regression line (`colors`/`linewidths`/`linetypes`) and SE ribbon (base `line_fill`, ribbon alpha via the `linear` entry of `alphas`)

```{r style-defaults}
style_doseprop()
```

A typical override:

```{r plot-styled, fig.height=4, fig.width=6, warning=FALSE}
plot_doseprop(
  dose_prop_obj,
  style = style_doseprop(
    shapes    = c(obs_point = 19),
    colors    = c(obs_point = "black", linear = "firebrick"),
    alphas    = c(obs_point = 0.8, linear = 0.2),
    sizes     = c(obs_point = 1),
    line_fill = "firebrick")
)
```

For a deeper treatment of the styling system — `style_spec()` fields, per-role maps, and post-hoc `restyle_plot()` / `reveal()` — see the [Plot Themes and Aesthetics](plot-themes.html) vignette.

# See also

  + [PK and PK/PD EDA workflow](eda-pk-pkpd-workflow.html) — exploratory analysis of continuous longitudinal concentration-time data, response-time, and response-concentration data.
  + [Plot Themes and Aesthetics](plot-themes.html) — element constructors, theme factories, and class system for customizing plot output.
