---
title: "Exploratory Analyses of PK/PD Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eda-pkpd-workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.retina = 1,
  dpi = 96,
  out.width = "100%"
)
```

This vignette will demonstrate `pmxhelpr` functions for exploratory data analysis of longitudinal dose-concentration-response data to inform population PK/PD modeling analysis. This includes package functions to visualize response over time and response versus concentration.

This vignette will assume familiarity with the `data_sad` internal dataset and the `plot_dvtime()` function, as described in the [Exploratory Analyses of PK Data](eda-pk-workflow.html) vignette. These elements will not be reviewed in detail in this vignette.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(forcats, warn.conflicts = FALSE)
library(Hmisc, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
library(PKNCA, warn.conflicts = FALSE)
```

# Data

The example dataset used in this vignette (`data_sad`) is based on a single ascending dose (SAD) study of an orally administered drug product with a parallel group food effect (FE) cohort.

As in [Exploratory Analyses of PK Data](eda-pk-workflow.html), we will define some new variables to pass to the color aesthetic that include counts of unique individuals per level of the grouping variable with `var_addn()`.

```{r plot-data, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_data <- data_sad %>% 
  filter(EVID == 0) %>% 
  mutate(`Food Status` = ifelse(FOOD == 0, "Fasted", "Fed"), 
         `Dose and Food` = paste(DOSE, "mg", `Food Status`)) %>% 
  mutate(Dose = var_addn(DOSE, ID, sep = "mg"),
         `Food Status` = var_addn(`Food Status`, ID),
         `Dose and Food` = var_addn(`Dose and Food`, ID)) %>% 
  mutate(Dose = fct_relevel(Dose, "50 mg (n=6)", after = 1), 
         `Dose and Food` = fct_relevel(`Dose and Food`, "50 mg Fasted (n=6)", after = 1))
```

This time, let's filter down to generate separate PK and PD observation datasets for this exploratory analysis.

```{r plot-data-cmt, message=FALSE, warning=FALSE}
plot_data_pk <- filter(plot_data, CMT == 2)
plot_data_pd <- filter(plot_data, CMT == 3)
```


# Population Response-time Plots with `plot_dvtime()`

Longitudinal pharmacodynamic response data can be visualized over time using `plot_dvtime()`. This function is designed to visualize a dependent variable versus time and can be used for PK and/or PD data!

```{r plot-dvtime-pd, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pd, dv_var = ODV, col_var = `Dose and Food`) +
  labs(y = "Response (% of Baseline)", x = "Time Since First Dose (hours)")
```

## Specifying a Reference Value for Change Metrics

A horizontal reference line can be added by specifying the `ref` argument with a numeric y-intercept value. For example, `ref = 100` adds a reference line at y = 100:

```{r plot-dvtime-pd-ref, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pd, dv_var = ODV,col_var = `Dose and Food`,
            ref = 100) +
  labs(y = "Response (% of Baseline)", x = "Time Since First Dose (hours)")
```

When the response variable is expressed as change from baseline, `ref = 0` provides the appropriate reference line:

```{r plot-dvtime-pd-ref0, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pd, dv_var = CFB,col_var = `Dose and Food`,
            ref = 0) +
   labs(y = "Response (% of Baseline)", x = "Time Since First Dose (hours)")
```

## Combining PK and PD Panels

Simultaneous visualization of the time-course of drug concentration and response can be achieved by combining two `plot_dvtime()` calls with the `patchwork` package. This approach provides full control over the content and aesthetics of each panel.

```{r plot-dvtime-pkpd-panels, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
p_pk <- plot_dvtime(data = plot_data_pk, dv_var = ODV, col_var = `Dose and Food`, 
                    log_y = TRUE) +
  labs(x = "Time Since First Dose (hours)", y = "Concentration (ng/mL)")

p_pd <- plot_dvtime(data = plot_data_pd, dv_var = ODV, col_var = `Dose and Food`,
                    ref = 100) +
  labs(x = "Time Since First Dose (hours)", y = "Response (% of Baseline)")

p_pk / p_pd + plot_layout(guides = "collect")
```

Each panel can be customized independently. For example, BLQ imputation can be applied to the PK panel while the PD panel uses change from baseline with its own central tendency specification with captions removed from both panels.

```{r plot-dvtime-pkpd-custom, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
p_pk <- plot_dvtime(data = plot_data_pk, dv_var = ODV,col_var = Dose, 
                    log_y = TRUE,loq_method = 2, show_caption = FALSE) +
  labs(x = "Time Since First Dose (hours)", y = "Concentration (ng/mL)")

p_pd <- plot_dvtime(data = plot_data_pd, dv_var = CFB,
                    col_var = Dose, cent = "median", show_caption = FALSE) +
  labs(x = "Time Since First Dose (hours)", y = "Response (% Change from Baseline)")

p_pk / p_pd + plot_layout(guides = "collect")
```

# Population Response-Concentration Plots with `plot_dvconc()`

## Overview

 `plot_dvconc()` is a plotting function intended to help with visualization of the relationship between a dependent variable for response and drug concentration.
 
## Specifying Depenent and Independent Variables

`plot_dvconc()` has 2 arguments that specify the dependent variable to be mapped to the y-axis `dv_var` and the independent variable to be mapped to the x-axis (`idv_var`), The defaults are as follows:

  - `dv_var` = DV, character string specifying the dependent variable to map to the y-axis. 
  - `idv_var`= CONC, character string specifying the dependent variable to map to the x-axis.
  
Both arguments use non-standard evaluation and can be passed as bare column names or strings.

```{r plot-dvconc, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = ODV, idv_var = CONC) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% of Baseline)")
```

## Specifying Change from Baseline

Like `plot_dvtime()`, we can add a horizontal reference line by specifying `ref` with a numeric y-intercept.

```{r plot-dvconc-ref, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC,
            ref = 0) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```

## Specifying the Color Aesthetic

`plot_dvconc()` includes two arguments for controlling the color aesthetic, `col_var` and `col_trend`. If a variable is passed to the `col_var` argument, the data points are colored based on this variable; however, by default `col_trend = FALSE` and the trend line is fit to the totality of the data  without stratifying the trend lines by the variable mapped to the color aesthetic. The `col_var` argument uses non-standard evaluation and can be passed as a bare column name or as a string.

```{r plot-dvconc-col, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC,
            ref = 0,
            col_var = `Dose and Food`, col_trend = FALSE) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```

Trend lines stratified by the variable mapped to the color aesthetic are requested by setting `col_trend = TRUE`.

```{r plot-dvconc-col-trend, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC,ref = 0,
           col_var = `Dose and Food`, col_trend = TRUE) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```

## Specifying the Central Tendency

There are two trend line types supported by `plot_dvconc()`

  - locally estimated scatter plot smoothing (LOESS) fit
  - linear regression. 
  
The central tendency trend lines visualized are controlled by logical arguments `loess` and `linear`. The default is `loess = TRUE` and `linear = FALSE`

```{r plot-dvconc-col-trend-loess-linear, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC, ref = 0, 
            col_var = `Dose and Food`, col_trend = FALSE, loess = TRUE, linear = TRUE) +
   labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```

The confidence intervals of the trend lines are suppressed by default in order to facilitate visualization of the central tendency and spread of observed data points simultaneously. Confidence intervals can be added to the plot using the logical arguments `se_loess` and `se_linear`.

```{r plot-dvconc-loess-linear-se, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC,ref = 0, 
            col_var = `Dose and Food`, col_trend = FALSE,
            loess = FALSE, linear = TRUE, se_loess = FALSE, se_linear = TRUE) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```

Additional arguments can be passed to `geom_smooth(method = "loess")`, such as increasing the span of the smoothing fit.

```{r plot-dvconc-loess-span, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC,ref = 0, 
            col_var = `Dose and Food`, col_trend = FALSE,
            loess = TRUE, linear = FALSE, se_loess = TRUE, se_linear = FALSE,
            span = 1) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```
If the color aesthetic is mapped to the trendlines with `col_trend = TRUE`, it will also map to the ribbons defining the CIs of the trend lines.

```{r plot-dvconc-col-trend-se, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC,ref = 0, 
            col_var = `Dose and Food`, col_trend = TRUE,se_loess = TRUE) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```


## Adjusting the Plot Theme with `plot_dvconc_theme()`

The default aesthetics for Response-Concentration plots are controlled via `plot_dvconc_theme()`. See the [Plot Themes and Aesthetics](plot-themes.html) vignette for details on the theme system, element constructors, and examples of customizing Response-Concentration aesthetics.

```{r theme}
plot_dvconc_theme()
```

Say we want to update the color of the trend line and standard error. This can be done by defining a new theme and passing that to the `theme` argument of `plot_dvconc()`.

```{r new-theme}
dvconc_new_theme <- plot_dvconc_theme(
  loess = pmx_trend(color = "darkred", se_color = "darkred")
)
```

```{r plot-dvconc-new-theme, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC,ref = 0, 
            col_var = `Dose and Food`, col_trend = FALSE,
            loess = TRUE, linear = FALSE, se_loess = TRUE, se_linear = FALSE,
            theme = dvconc_new_theme) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```



