---
title: "Exploratory Analyses of PK and PK/PD Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploratory Analyses of PK and PK/PD Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

This vignette will demonstrate `pmxhelpr` functions for exploratory data analysis of dose-concentration-time data to inform population PK modeling analysis. This includes package functions to visualize concentration-time data and to assess dose-proportionality of exposure.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(forcats, warn.conflicts = FALSE)
library(Hmisc, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
library(PKNCA, warn.conflicts = FALSE)
```

# Data

The example dataset used in this vignette is based on a single ascending dose (SAD) study of an orally administered drug product with a parallel group food effect (FE) cohort. 

## data_sad

Dataset definitions can be viewed by calling `?data_sad`. A quick summary of variables and types is printed below.

```{r data-sad-glimpse}
glimpse(data_sad)
```

The study design consists of two parts. 

  + Part 1 was a sequential single ascending dose study covering a 10 to 400 mg dose range.
  + Part 2 was a parallel 100 mg single dose food effect (FE) study. 

```{r data-sad-part}
distinct(data_sad, DOSE, PART, FOOD)
```

The dataset is formatted for population pharmacokinetic (PopPK) and pharmacokinetic/pharmacodynamic (PK/PD) modeling in NONMEM with the following compartments and data types:

  - `CMT=1`: oral dose events (`EVID=1`) 
  - `CMT=2`: drug concentration observations (`EVID=0`) 
  - `CMT=3`: pharmacodynamic response observations (`EVID=0`) 
  
Dose events are input based on `AMT`. The nominal (e.g. protocol assigned) dose associated with each observation is captured in `DOSE`.

Plasma drug concentration and pharmacodynamic response observations are expressed in multiple units:  
  - `ODV`: original units of the dependent variable [`CMT=2` ng/mL, `CMT=3` % baseline activity]
  - `LDV`: log-transformed drug concentration [`CMT = 2` log(ng/mL), `CMT = 3` % baseline activity]
  - `CFB`: percentage change from baseline [`CMT=2` missing, `CMT=3` % change from baseline]

Time is expressed as actual time of dose input or observation and nominal time of observations:

  - `TIME`: Actual time since first dose administration [hours]
  - `NTIME`: Nominal time dose event or sample collection per protocol [hours]
  - `NDAY`: Nominal day on study [day]

The nominal time variables are *exact binning variables*, which are useful for grouping data for exploratory data analysis or model evaluation.

```{r data-sad-ntime}
##Unique values of time variables
unique(data_sad$NDAY)
unique(data_sad$NTIME)
```

## var_addn

Now let's pre-process these data for visualization. Often, it is useful to assign study design variables to the color aesthetic in plots to visualize data stratified by these study design elements. One way to extract even more information from the plots is to include the sample size within each level of the variable mapped to the color aesthetic. `pmxhelpr` provides a useful vectorized helper function for this purpose: `var_addn()`

The helper function `var_addn()` takes two vector variables as arguments: a grouping variable (`grp_var`) and an identifier variable (`id_var`). The function counts unique values of `id_var` per level of `grp_var` and returns a factor vector with group labels appended with counts of unique identifiers. It can be used inside `mutate()` to transform columns in place. 

A string constant separator can be added between the values in `grp_var` and the count of `id_var` using the optional `sep` argument. A common use is to specify the dose units when linking dose and count of unique individuals assigned to that dose level.

Let's define some new variables and count unique subjects in each for use in plotting. Useful study design variables for stratifying plots include: dose (DOSE), food status (FOOD), and their interaction.

```{r plot-data, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_data <- data_sad %>% 
  filter(EVID == 0) %>% 
  mutate(`Food Status` = ifelse(FOOD == 0, "Fasted", "Fed"), 
         `Dose and Food` = paste(DOSE, "mg", `Food Status`)) %>% 
  mutate(Dose = var_addn(DOSE, ID, sep = "mg"),
         `Food Status` = var_addn(`Food Status`, ID),
         `Dose and Food` = var_addn(`Dose and Food`, ID))

unique(plot_data$Dose)
unique(plot_data$`Food Status`)
unique(plot_data$`Dose and Food`)
```
Unfortunately, often our factors will not be ordered in ascending numerical order. In this case, the 50 mg dose group is sorted at the end. We can use the `forcats` package from the `tidyverse` to quickly reorder our new variables

```{r plot-data-reorder, message=FALSE, warning=FALSE}
plot_data <- plot_data %>% 
  mutate(Dose = fct_relevel(Dose, "50 mg (n=6)", after = 1), 
         `Dose and Food` = fct_relevel(`Dose and Food`, "50 mg Fasted (n=6)", after = 1))

unique(plot_data$Dose)
unique(plot_data$`Dose and Food`)
```

Finally, let's filter down only to PK observations for this exploratory analysis, which is focused on assessing the informational content of the data and trends for population PK analysis.

```{r plot-data-cmt, message=FALSE, warning=FALSE}
plot_data_pk <- filter(plot_data, CMT == 2)
plot_data_pd <- filter(plot_data, CMT == 3)
```



# Population Concentration-time Plots with `plot_dvtime()`


## Overview

 `pmxhelpr` includes a function for common visualizations of observed concentration-time data in exploratory data analysis: `plot_dvtime()`

Let's try passing in our PK dataset with no other arguments. 

```{r plot-dvtime-fail, fig.height=4, fig.width=6, error=TRUE, message=TRUE}
plot_dvtime(plot_data_pk)
```
The function with no arguments other than the input dataset returns an error with a helpful error message letting us know the additional argument required to be specified.

## Specifying Dependent and Independent Variables 

`plot_dvtime()` has 3 arguments that specify the dependent variable to be mapped to the y-axis: 

  - `dv_var` = DV (default), the dependent variable to be mapped to the y-axis
  - `time_var` = TIME (default), the independent variable to be mapped to the x-axis
  - `ntime_var` = NTIME (default), the exact bin version of the x-axis variable for calculation of summary statistics
  
These arguments all support non-standard evaluation and these arguments may be passed as bare names or strings. Since `data_sad` has the dependent variable mapped to the name `ODV`, this name must be passed to the function.

```{r plot-dvtime-pass, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(plot_data_pk, dv_var = ODV)
```

Now the function returns a plot! Conveniently, the time variables in `data_sad` had names aligned with the default arguments and did not have to be specified as arguments; however, this may not always be the case. 

The same variable name can be passed to both `time_var` and `ntime_var` arguments for the special case where only nominal times are known.

```{r plot-dvtime-nom, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(plot_data_pk, dv_var = ODV, time_var = NTIME, ntime_var = NTIME)
```

## Suppressing the Caption

A caption prints by default describing the plot elements. The caption can be suppressed by specifying `show_caption = FALSE`.

```{r plot-dvtime-nocap, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = ODV, show_caption = FALSE) 
```

## Setting Axis Scales, Breaks, and Labels

 `plot_dvtime()` includes one argument (`log_y`), which performs a log10 transformation of the y-axis with some additional formatting benefits over manually adding the layer to the returned object with `scale_y_log10()`.

  + Includes log tick marks on the y-axis 
  + Updates the caption with the correct central tendency measure if `show_captions = TRUE`. 

`plot_dvtime()` uses the `stat_summary()` function from `ggplot2` to calculate and plot the central tendency measures and error bars. An often overlooked feature of `stat_summary()`, is that it calculates the summary statistics *after* any transformations to the data performed by changing the scales. This means that when `scale_y_log10()` is applied to the plot, the data are log-transformed for plotting and the central tendency measure returned when requesting `"mean"` from `stat_summary()` is the *geometric mean*. If the `log_y` argument is used to generate semi-log plots along with `show_captions = TRUE`, then the caption will  delineate where arithmetic and geometric means are being returned. 

```{r plot-dvtime-logy, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = ODV, log_y = TRUE) 
```

The x-axis breaks and axis labels can always be customized by adding `scale_x_continuous()` and/or `labs()` layers to the returned `ggplot` object. The default for ggplot is to break the x-axis in units divisible by 5, which do not often map well to time units and should routinely be overwritten.

```{r plot-dvtime-breaks, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = ODV, log_y = TRUE) + 
  scale_x_continuous(breaks = seq(0, 168, 24)) + 
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)")
```

## Specifying the Color Aesthetic

The color aesthetic can be mapped to a dataset variable using the `col_var` argument. This argument uses non-standard evaluation and can be passed as a bare column name or as a string. 

This can be demonstrated using the `Dose and Food` variable we defined earlier with sample size counts using `var_addn()`.

```{r plot-dvtime-col, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = ODV, log_y = TRUE, 
            col_var = `Dose and Food`) + 
  scale_x_continuous(breaks = seq(0, 168, 24)) + 
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)")
```

## Specifying the Central Tendency

The argument `cent` specifies the method of calculating the central tendency (+/- variability) within levels of the variable passed to the `col_var`. The default is `cent = "mean"`; however, note that the calculation is performed *after* any transformations to the data and this option will return the geometric mean when `log_y=TRUE`. 

If the `log_y` argument is used to generate semi-log plots along with `show_captions = TRUE`, then the caption will automatically delineate where arithmetic and geometric means are being returned. 

```{r plot-dvtime-part-log, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = ODV, col_var = `Dose and Food`, cent = "mean") +
  scale_x_continuous(breaks = seq(0, 168, 24)) + 
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)")
```

`plot_dvtime()` returns a `ggplot` object which we can be further modified using native `ggplot` functions. Therefore, we can facet by PART by simply adding in another layer to our `ggplot` object.

```{r plot-dvtime-part, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = ODV, col_var = `Dose and Food`, cent = "mean",
            log_y = TRUE) +
  scale_x_continuous(breaks = seq(0, 168, 24)) + 
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

The plot can be simplified to show mean +/- standard deviation by specifying `cent = "mean_sdl"`. The observed points just add noise when combined with error bars and can be hidden by setting `obs_point = pmx_point(alpha = 0)` in the theme.

```{r plot-dvtime_part-meansdl, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = ODV, col_var = `Dose and Food`, 
            cent = "mean_sdl",log_y = TRUE,
            theme = plot_dvtime_theme(obs_point = pmx_point(alpha = 0))) +
  scale_x_continuous(breaks = seq(0, 168, 24)) + 
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

For data with high variability visualized on a linear y-axis, the error bars may drop below zero. In this case, the upper error bar only may be requested by specifying `cent = "mean_sdl_upper"`.

```{r plot-dvtime_part-meansdl-upper, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = ODV, col_var = `Dose and Food`, 
            cent = "mean_sdl_upper",
            theme = plot_dvtime_theme(obs_point = pmx_point(alpha = 0))) +
  scale_x_continuous(breaks = seq(0, 168, 24)) + 
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

For small samples or non-normally (or log-normally) distributed data, the median and inter-quartile range (IQR) error bars may be requested by specifying  `cent = "median_iqr"`.

```{r plot-dvtime_part-medianiqr, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = ODV, col_var = `Dose and Food`, 
            cent = "median_iqr",log_y = TRUE,
            theme = plot_dvtime_theme(obs_point = pmx_point(alpha = 0))) +
  scale_x_continuous(breaks = seq(0, 168, 24)) + 
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

## Connecting Repeated Measures at the Individual Level

Data points may be connected longitudinally within an individual by specifying `id_var` (e.g., `id_var = ID`). The grouping variable inherits the variable mapped to the color aesthetic with `col_var`. The central tendency may be removed by specifying `cent ="none"` or the median may be plotted by specifying `cent="median"`.

This argument supports non-standard evaluation, and may be passed as bare names or strings.

```{r plot-dvtime-part-ind, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = ODV, col_var = `Dose and Food`, cent = "median",
            log_y = TRUE,id_var = ID) +
  scale_x_continuous(breaks = seq(0, 168, 24)) + 
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

## Handling of BLQ Data

`plot_dvtime()` includes functionality to automatically handle imputation of PK data censored due to assay limitations. This requires specification of two arguments: `loq` and `loq_method`.

The `loq_method` argument specifies how BLQ imputation should be performed. Options are:

  + `0` or `"none"`: No handling. Plot input dataset `DV` vs `TIME` as is. (default)
  + `1` or `"postdose"`: Impute all BLQ data at `TIME` <= 0 to 0 and all BLQ data at `TIME` > 0 to 1/2 x `loq`.
      Useful for plotting concentration-time data with some data BLQ on a linear scale
  + `2` or `"all"`: Impute all BLQ data to 1/2 x `loq`.
      Useful for plotting concentration-time data with some data BLQ on a log scale
  
The `loq` argument specifies the numeric value of the LLOQ. The `loq` argument must be specified when `loq_method` is non-zero, but can be `NULL` *if* the variable `LLOQ` is present in the dataset and the value of `loq` will be inherited from the data.

```{r plot-dvtime-blq1, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(plot_data_pk, dv_var = ODV, col_var = `Dose and Food`, 
            cent = "mean",log_y = TRUE,loq_method = "all") +
  scale_x_continuous(breaks = seq(0, 168, 24)) + 
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

The same plot is obtained by specifying `loq_method = 2` and `loq = 1`

```{r plot-dvtime-blq2, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(plot_data_pk, dv_var = ODV, col_var = `Dose and Food`, 
            cent = "mean", log_y = TRUE,loq_method = 2, loq = 1) +
  scale_x_continuous(breaks = seq(0, 168, 24)) + 
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

A reference line is drawn to denote the LLOQ and all observations with `EVID=0` and `MDV=1` are imputed as LLOQ/2. The numeric value of LLOQ is printed in the legend and a caption is added to indicate the imputation method for BLQ data.

## Applying Dose-normalization

`plot_dvtime()` also has functionality to  generate dose-normalized concentration-time plots by specifying `dosenorm = TRUE`. 

```{r plot-dvtime-dn, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(plot_data_pk, dv_var = ODV, col_var = `Dose and Food`, cent = "mean",
            log_y = TRUE,
            dosenorm = TRUE) +
  scale_x_continuous(breaks = seq(0, 168, 24)) + 
  labs(y = "Dose-normalized Conc. (ng/mL/mg)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

When `dosenorm = TRUE`, the variable specified in `dose_var` (default = DOSE) needs to be present in the input dataset `data`. If `dose_var` is not present in `data`, the function will return an *Error* with an informative error message. This argument uses non-standard evaluation and can be passed as a bare column name or as a string.

```{r plot-dvtime-dn-nodose, fig.height=4, fig.width=6, message=TRUE, warning=TRUE, error=TRUE}
plot_dvtime(select(plot_data_pk, -DOSE),
            dv_var = ODV, col_var = `Dose and Food`, 
            cent = "mean", log_y = TRUE,dosenorm = TRUE) +
  labs(y = "Dose-normalized Conc. (ng/mL/mg)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

Dose-normalization is performed *AFTER* BLQ imputation in the case in which both options are requested. The reference line for the LLOQ will not be plotted when dose-normalized concentration is the dependent variable. 

```{r plot-dvtime-blq-dn, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(plot_data_pk, dv_var = ODV, col_var = `Dose and Food`, cent = "mean",
            log_y = TRUE,
            loq_method = 2, dosenorm = TRUE) +
  labs(y = "Dose-normalized Conc. (ng/mL/mg)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

## Adjusting the Plot Theme with `plot_dvtime_theme()`

The default aesthetics for Response-Time plots are controlled via `plot_dvtime_theme()`. See the [Plot Themes and Aesthetics](plot-themes.html) vignette for details on the theme system, element constructors, and examples of customizing Response-Time aesthetics.

```{r theme}
plot_dvtime_theme()
```

Say we want to update the errorbar cap width to be more visible in our prior geometric mean +/- geometric SD plot with the interaction of dose and food passed to the color aesthetic. This can be done by defining a new theme and passing that to the `theme` argument of `plot_dvtime()`.

```{r new-theme}
dvtime_new_theme <- plot_dvtime_theme(
  obs_point = pmx_point(alpha = 0),
  cent_errorbar  = pmx_errorbar(width = 10)
)
```

```{r plot-dvtime-blq-dn-new-theme, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pk, dv_var = ODV, col_var = `Dose and Food`, 
            cent = "mean_sdl",log_y = TRUE,
            theme = dvtime_new_theme) +
  scale_x_continuous(breaks = seq(0, 168, 24)) + 
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)
```

# Individual Concentration-time plots with `plot_dvtime()`

The previous section provides an overview of how to generate population concentration-time profiles by dose using `plot_dvtime()`; however, we can also use `plot_dvtime()` to generate subject-level visualizations with a little pre-processing of the input dataset.

We can specify `cent = "none"` to remove the central tendency layer when plotting individual subject data.

```{r plot-dvtime-blq2-indonly, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(plot_data_pk, dv_var = ODV, col_var = `Dose and Food`, 
            cent = "none",log_y = TRUE, id_var = ID,
            loq_method = 2, loq = 1) +
  labs(y = "Concentration (ng/mL)") +
  facet_wrap(~PART)
```
We can plot an individual subject by filtering the input dataset. This could be extended to generate plots for all individuals using `for` loops, `lapply()`,  `purrr::map()` functions, or other methods.

```{r plot-dvtime-blq2-indplots, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}

ids <- sort(unique(plot_data_pk$ID)) #vector of unique subject ids
n_ids <- length(ids) #count of unique subject ids
plots_per_pg <- 4
n_pgs <- ceiling(n_ids/plots_per_pg) #Total number of pages needed

plist<- list()
for(i in 1:n_ids){
  plist[[i]] <- plot_dvtime(filter(plot_data_pk, ID == ids[i]),
                               dv_var = ODV, cent = "none",
            log_y = TRUE,
            id_var = ID,
            loq_method = 2, loq = 1, show_caption = FALSE) +
  labs(y = "Concentration (ng/mL)", x = "Time Since First Dose (hours)") +
  facet_wrap(~PART)+
  labs(title = paste0("ID = ", ids[i], " | Dose = ", unique(plot_data_pk$DOSE[plot_data_pk$ID==ids[i]]), " mg"))+
  theme(legend.position="none")
}

lapply(1:n_pgs, function(n_pg) {
      i <-  (n_pg-1)*plots_per_pg+1
      j <- n_pg*plots_per_pg
      wrap_plots(plist[i:j])
})
```



# Population Response-time Plots with `plot_dvtime()`

Longitudinal pharmacodynamic response data can also be visualized over time using `plot_dvtime()`. This function is designed to visualize a dependent variable versus time and can be used for PK and/or PD data!

```{r plot-dvtime-pd, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pd, dv_var = ODV, col_var = `Dose and Food`) +
  labs(y = "Response (% of Baseline)", x = "Time Since First Dose (hours)")
```

## Specifying a Reference Value for Change Metrics

A horizontal reference line can be added by specifying the `ref` argument with a numeric y-intercept value. For example, `ref = 100` adds a reference line at y = 100:

```{r plot-dvtime-pd-ref, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pd, dv_var = ODV,col_var = `Dose and Food`,
            ref = 100) +
  labs(y = "Response (% of Baseline)", x = "Time Since First Dose (hours)")
```

When the response variable is expressed as change from baseline, `ref = 0` provides the appropriate reference line:

```{r plot-dvtime-pd-ref0, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data = plot_data_pd, dv_var = CFB,col_var = `Dose and Food`,
            ref = 0) +
   labs(y = "Response (% of Baseline)", x = "Time Since First Dose (hours)")
```

## Combining PK and PD Panels

Simultaneous visualization of the time-course of drug concentration and response can be achieved by combining two `plot_dvtime()` calls with the `patchwork` package. This approach provides full control over the content and aesthetics of each panel.

```{r plot-dvtime-pkpd-panels, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
p_pk <- plot_dvtime(data = plot_data_pk, dv_var = ODV, col_var = `Dose and Food`, 
                    log_y = TRUE) +
  labs(x = "Time Since First Dose (hours)", y = "Concentration (ng/mL)")

p_pd <- plot_dvtime(data = plot_data_pd, dv_var = ODV, col_var = `Dose and Food`,
                    ref = 100) +
  labs(x = "Time Since First Dose (hours)", y = "Response (% of Baseline)")

p_pk / p_pd + plot_layout(guides = "collect")
```

Each panel can be customized independently. For example, BLQ imputation can be applied to the PK panel while the PD panel uses change from baseline with its own central tendency specification with captions removed from both panels.

```{r plot-dvtime-pkpd-custom, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
p_pk <- plot_dvtime(data = plot_data_pk, dv_var = ODV,col_var = Dose, 
                    log_y = TRUE,loq_method = 2, show_caption = FALSE) +
  labs(x = "Time Since First Dose (hours)", y = "Concentration (ng/mL)")

p_pd <- plot_dvtime(data = plot_data_pd, dv_var = CFB,
                    col_var = Dose, cent = "median", show_caption = FALSE) +
  labs(x = "Time Since First Dose (hours)", y = "Response (% Change from Baseline)")

p_pk / p_pd + plot_layout(guides = "collect")
```

# Population Response-Concentration Plots with `plot_dvconc()`

## Overview

 `plot_dvconc()` is a plotting function intended to help with visualization of the relationship between a dependent variable for response and drug concentration.
 
## Specifying Dependent and Independent Variables

`plot_dvconc()` has 2 arguments that specify the dependent variable to be mapped to the y-axis `dv_var` and the independent variable to be mapped to the x-axis (`idv_var`). The defaults are as follows:

  - `dv_var` = DV, character string specifying the dependent variable to map to the y-axis.
  - `idv_var`= CONC, character string specifying the independent variable to map to the x-axis.
  
Both arguments use non-standard evaluation and can be passed as bare column names or strings.

```{r plot-dvconc, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = ODV, idv_var = CONC) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% of Baseline)")
```

## Specifying Change from Baseline

Like `plot_dvtime()`, we can add a horizontal reference line by specifying `ref` with a numeric y-intercept.

```{r plot-dvconc-ref, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC,
            ref = 0) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```

## Specifying the Color Aesthetic

`plot_dvconc()` includes two arguments for controlling the color aesthetic, `col_var` and `col_trend`. If a variable is passed to the `col_var` argument, the data points are colored based on this variable; however, by default `col_trend = FALSE` and the trend line is fit to the totality of the data without stratifying the trend lines by the variable mapped to the color aesthetic. 

The `col_var` argument uses non-standard evaluation and can be passed as a bare column name or as a string.

```{r plot-dvconc-col, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC,
            ref = 0,
            col_var = `Dose and Food`, col_trend = FALSE) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```

Trend lines stratified by the variable mapped to the color aesthetic are requested by setting `col_trend = TRUE`.

```{r plot-dvconc-col-trend, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC,ref = 0,
           col_var = `Dose and Food`, col_trend = TRUE) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```

## Specifying the Central Tendency

There are two trend line types supported by `plot_dvconc()`

  - locally estimated scatter plot smoothing (LOESS) fit
  - linear regression. 
  
The central tendency trend lines visualized are controlled by logical arguments `loess` and `linear`. The default is `loess = TRUE` and `linear = FALSE`

```{r plot-dvconc-col-trend-loess-linear, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC, ref = 0, 
            col_var = `Dose and Food`, col_trend = FALSE, loess = TRUE, linear = TRUE) +
   labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```

The confidence intervals of the trend lines are suppressed by default in order to facilitate visualization of the central tendency and spread of observed data points simultaneously. Confidence intervals can be added to the plot using the logical arguments `se_loess` and `se_linear`.

```{r plot-dvconc-loess-linear-se, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC,ref = 0, 
            col_var = `Dose and Food`, col_trend = FALSE,
            loess = FALSE, linear = TRUE, se_loess = FALSE, se_linear = TRUE) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```

Additional arguments can be passed to `geom_smooth(method = "loess")`, such as increasing the span of the smoothing fit.

```{r plot-dvconc-loess-span, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC,ref = 0, 
            col_var = `Dose and Food`, col_trend = FALSE,
            loess = TRUE, linear = FALSE, se_loess = TRUE, se_linear = FALSE,
            span = 1) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```
If the color aesthetic is mapped to the trendlines with `col_trend = TRUE`, it will also map to the ribbons defining the CIs of the trend lines.

```{r plot-dvconc-col-trend-se, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC,ref = 0, 
            col_var = `Dose and Food`, col_trend = TRUE,se_loess = TRUE) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```


## Adjusting the Plot Theme with `plot_dvconc_theme()`

The default aesthetics for Response-Concentration plots are controlled via `plot_dvconc_theme()`. See the [Plot Themes and Aesthetics](plot-themes.html) vignette for details on the theme system, element constructors, and examples of customizing Response-Concentration aesthetics.

```{r theme-dvconc}
plot_dvconc_theme()
```

Say we want to update the color of the trend line and standard error. This can be done by defining a new theme and passing that to the `theme` argument of `plot_dvconc()`.

```{r new-theme-dvconc}
dvconc_new_theme <- plot_dvconc_theme(
  loess = pmx_trend(color = "darkred", se_color = "darkred")
)
```

```{r plot-dvconc-new-theme, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(data = plot_data_pd, dv_var = CFB, idv_var = CONC,ref = 0,
            loess = TRUE, linear = FALSE, se_loess = TRUE, se_linear = FALSE,
            theme = dvconc_new_theme) +
  labs(x = "Drug Concentration (ng/mL)", y = "Response (% Change from Baseline)")
```


# See also

  + [Dose-proportionality Assessment workflow](doseprop-workflow.html) — statistical assessment of dose-proportionality of exposure using power law (log-log) regression
  + [Plot Themes and Aesthetics](plot-themes.html) — element constructors, theme factories, and class system for customizing plot output.
