---
title: "Plot Themes and Aesthetics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot Themes and Aesthetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

This vignette demonstrates how to control the appearance of the `pmxhelpr` plotting functions — `plot_dvtime()`, `plot_dvconc()`, `plot_doseprop()`, `plot_gof()`, `plot_vpc_cont()`, and `plot_vpc_cens()`. `pmxhelpr` styles its plots with the [ggstylekit](https://github.com/A2-ai/ggstylekit) package, which gives a set of packages one consistent way to style the plots they produce.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(ggstylekit)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(Hmisc, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
```



# Data and Model Objects

The example datasets used in this vignette are based around the `data_sad` dataset internal to `pmxhelpr`. This dataset is based on a single ascending dose (SAD) study of an orally administered drug product with a parallel group food effect (FE) cohort and is formatted in a analysis-ready format for non-linear mixed effects (NLME) Population PK or PK/PD modeling. `data_sad_pkfit` is a modified version of `data_sad`, which includes additional PK model predictions appended.

This vignette will assume familiarity with these internal datasets from other vignettes. These elements will not be reviewed in detail in this vignette, which is focused only on aesthetic controls across plotting functions.

## data_sad

Dataset definitions can be viewed by calling `?data_sad`.

Let's define some variables that may be useful in plotting and filter down to PK or PD relevant records.

```{r data-factors, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
data <- data_sad %>%
  mutate(`Food Status` = ifelse(FOOD == 0, "Fasted", "Fed"),
         DoseFood = paste(DOSE, "mg", `Food Status`)) %>%
  mutate(`Dose and Food` = var_addn(DoseFood, ID))

gof_data <- data_sad_pkfit %>%
  mutate(`Food Status` = ifelse(FOOD == 0, "Fasted", "Fed"),
         DoseFood = paste(DOSE, "mg", `Food Status`)) %>%
  mutate(`Dose and Food` = var_addn(DoseFood, ID))

data_pk <- filter(data, CMT %in% c(1,2))
data_pd <- filter(data, CMT %in% c(1,3))
data_pkfit <- filter(gof_data, CMT %in% c(1,2))
data_nca <- filter(data_sad_nca, PART == "Part 1-SAD")
doseprop_data_nca <- df_doseprop(data_nca, metrics = c("aucinf.obs", "cmax"), metric_name_var = "PPTESTCD")
```

An example PK model (`pkmodel`) in `mrgmod` format is provided in the internal package library. This is loaded using the helper function `model_mread_load()`, which wraps `mrgsolve::mread`.

```{r model-pk}
model <- model_mread_load("pkmodel")
```


# Overview

Every `plot_*()` function has a paired **style preset** — `style_dvtime()`, `style_gof()`, `style_dvconc()`, `style_doseprop()`, and `style_vpc()` — that returns a `ggstylekit::style_spec()` pre-filled with that plot family's defaults. Pass one to the plot function's `style` argument:

```{r overview, eval=FALSE}
plot_dvtime(data_pk, dv_var = ODV, style = style_dvtime())
```

Styling is keyed by each plot's **roles**. A role names a layer of the plot — for example `obs_point` (observed points), `obs_line` (spaghetti lines), `cent_point` / `cent_line` / `cent_errorbar` (central tendency), `ref_line`, and `loq_line`. The `style_spec()` per-series maps (`colors`, `fill`, `shapes`, `sizes`, `linetypes`, `linewidths`, `alphas`) are keyed by role.

A finished plot can also be adjusted after the fact with `ggstylekit::restyle_plot()`, or a hidden covariate surfaced with `ggstylekit::reveal()` — both covered at the end of this vignette.


# Viewing and overriding defaults

Call a preset with no arguments to print its defaults:

```{r view-defaults}
style_dvtime()
```

Override a field by passing it to the preset. Per-series maps **merge entry-wise** onto the defaults, so setting one role leaves the others unchanged:

```{r override-defaults}
# Hide the observed points and thicken the central-tendency line;
# every other role keeps its default.
style_dvtime(
  alphas     = c(obs_point = 0),
  linewidths = c(cent_line = 1.5)
)
```

Fields that are not per-series maps (e.g. `title`, `xlabel`, `logy`, `theme`) replace their default wholesale.


# The `style_spec`

A `style_spec()` bundles three kinds of control:

* **Per-series maps**, keyed by role: `colors` (line/point color), `fill` (ribbon fill), `shapes`, `sizes`, `linetypes`, `linewidths`, `alphas`.
* **Base entity aesthetics** applied to a whole geom family: `point_color`, `point_shape`, `line_color`, `line_linewidth`, `line_fill` (the fill of a `geom_smooth` SE ribbon), and so on.
* **Plot-level polish**: `title`, `xlabel`, `ylabel`, `logx`/`logy`, `xlims`/`ylims`, legend controls, faceting, and a ggplot2 `theme`.

pmxhelpr sets a shared house theme (`theme_bw()` with the minor and vertical-major gridlines removed; a white panel for VPCs) via each preset's `theme` field.

> **Note:** the error-bar cap width is a builder argument (`errorbar_width` on `plot_dvtime()` / `plot_gof()`), not a `style_spec` field. It defaults to 2.5% of the maximum nominal time.


# Roles by plot family

| Preset | Roles |
|---|---|
| `style_dvtime()` | `obs_point`, `obs_line`, `cent_point`, `cent_line`, `cent_errorbar`, `ref_line`, `loq_line` |
| `style_gof()` | as `dvtime`, plus the overlay `colors` keyed by label: `OBS`, `DV`, `PRED`, `IPRED` |
| `style_dvconc()` | `obs_point`, `ref_line`, `loess`, `linear` |
| `style_doseprop()` | `obs_point`, `linear` |
| `style_vpc()` | `obs_point`, `obs_median_line`, `obs_pi_line`, `sim_pi_line`, `sim_median_line`, `loq_line` (lines); `sim_pi_ci`, `sim_pi_area`, `sim_median_ci` (ribbon `fill`) |


# Examples

## Inline override

```{r ex-inline, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(
  data = data_pk, dv_var = ODV, col_var = `Dose and Food`,
  cent = "mean_sdl", log_y = TRUE,
  style = style_dvtime(alphas = c(obs_point = 0))
) +
  labs(y = "Concentration (ng/mL)", x = "Time (hours)")
```

## Reusable style object

Build a style once and reuse it across plots:

```{r ex-reusable, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
my_style <- style_dvtime(
  alphas     = c(obs_point = 0),
  linewidths = c(cent_line = 1.5, cent_errorbar = 1.5)
)

plot_dvtime(data_pk, dv_var = ODV, col_var = `Dose and Food`,
            cent = "mean_sdl", log_y = TRUE, style = my_style) +
  labs(y = "Concentration (ng/mL)", x = "Time (hours)")
```

## Error bar cap width

Width is a builder argument, not part of the style:

```{r ex-ebwidth, eval=FALSE}
plot_dvtime(data_pk, dv_var = ODV, cent = "mean_sdl",
            style = my_style, errorbar_width = 10)
```

## Trend line customization (`plot_dvconc()`)

The trend line color is the `loess`/`linear` entry of `colors`; its SE-ribbon fill is the base `line_fill`:

```{r ex-trend, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(
  data = data_pd, dv_var = CFB, idv_var = CONC, ref = 0,
  loess = TRUE, se_loess = TRUE,
  style = style_dvconc(colors = c(loess = "darkred"), line_fill = "darkred")
) +
  labs(x = "Concentration (ng/mL)", y = "Response (% CFB)")
```

## GOF overlay colors (`plot_gof()`)

The DV / PRED / IPRED (and OBS) overlays are colored by the `colors` map keyed by those labels:

```{r ex-gof, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_gof(
  data = data_pkfit, dv_var = ODV, cent = "mean", log_y = TRUE,
  style = style_gof(colors = c(OBS = "grey60", DV = "black",
                               PRED = "#3388cc", IPRED = "firebrick"))
) +
  labs(y = "Concentration (ng/mL)", x = "Time (hours)")
```

## VPC color scheme (`plot_vpc_cont()`)

Line/point colors live in `colors`; ribbon fills live in `fill`:

```{r ex-vpc, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
sim <- df_mrgsim_replicate(data = data_pk, model = model, replicates = 20,
                           dv_var = ODV, carry_out = c("LLOQ", "FOOD"))

vpc_style <- style_vpc(
  colors = c(obs_point = "#000000", obs_median_line = "#000000",
             obs_pi_line = "#000000"),
  fill   = c(sim_median_ci = "#3388cc", sim_pi_ci = "#3388cc")
)

plot_vpc_cont(sim, loq = 1, style = vpc_style)
```


# Post-hoc styling: `restyle_plot()` and `reveal()`

Because the plots are finished ggstylekit objects, they can be adjusted after they are built — without re-running the plot function.

`ggstylekit::restyle_plot()` merges style fields onto a finished plot (per-series maps merge entry-wise):

```{r restyle, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
p <- plot_dvtime(data_pk, dv_var = ODV, cent = "mean")
restyle_plot(p, colors = c(obs_point = "darkgreen"))
```

`ggstylekit::reveal()` maps a covariate that is present in the plot's data but not currently shown onto a graphical channel (or a facet):

```{r reveal, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
reveal(p, FOOD, as = "colors")
```


# Inspecting styles

Print a style to see its set fields, and validate it with `ggstylekit::validate_style_spec()`:

```{r inspect}
style_vpc()

validate_style_spec(style_dvtime())
```
