---
title: "Plot Styling and Aesthetics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot Styling and Aesthetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

This vignette demonstrates how to control the visual aesthetics of the 
`pmxhelpr` plotting functions. `pmxhelpr` styles its plots with `ggstylekit`, 
which provides a uniform approach to styling plots.

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(ggstylekit)
library(dplyr, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(Hmisc, warn.conflicts = FALSE)
library(patchwork, warn.conflicts = FALSE)
```



# Data and Model Objects

The example datasets used in this vignette are based around the `data_sad` 
dataset internal to `pmxhelpr`. This dataset is based on a single ascending 
dose (SAD) study of an orally administered drug product with a parallel group 
food effect (FE) cohort and is formatted in a analysis-ready format for 
non-linear mixed effects (NLME) modeling. `data_sad_pkfit` is a modified 
version of `data_sad`, with PK model predictions appended.

This vignette will assume familiarity with the structure of these datasets 
from other vignettes. 


## data_sad

Dataset definitions can be viewed by calling `?data_sad`.

Let's define some variables that may be useful in plotting and filter 
down to PK or PD relevant records.

```{r data-factors, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
data <- data_sad %>%
  mutate(`Food Status` = ifelse(FOOD == 0, "Fasted", "Fed"),
         DoseFood = paste(DOSE, "mg", `Food Status`)) %>%
  mutate(`Dose and Food` = var_addn(DoseFood, ID))

gof_data <- data_sad_pkfit %>%
  mutate(`Food Status` = ifelse(FOOD == 0, "Fasted", "Fed"),
         DoseFood = paste(DOSE, "mg", `Food Status`)) %>%
  mutate(`Dose and Food` = var_addn(DoseFood, ID))

data_pk <- filter(data, CMT %in% c(1,2))
data_pd <- filter(data, CMT %in% c(1,3))
data_pkfit <- filter(gof_data, CMT %in% c(1,2))
data_nca <- filter(data_sad_nca, PART == "Part 1-SAD")
doseprop_data_nca <- df_doseprop(data_nca, metrics = c("aucinf.obs", "cmax"), metric_name_var = "PPTESTCD")
```

An example PK model (`pkmodel`) in `mrgmod` format is provided in the internal
package library. This is loaded using the helper function `model_mread_load()`,
which wraps `mrgsolve::mread()`.

```{r model-pk}
model <- model_mread_load("pkmodel")
```



# Plot Styling with `ggstylekit` in `pmxhelpr`

`pmxhelpr` re-exports four functions from `ggstylekit`:

  + `reveal()`: map a new data variable to an aesthetic or facet, then re-style
  + `restyle_plot()`: adjust a plot style after it has been styled.
  + `combine_styled_plots()`: combine styled plots and collect legends
  + `legend_spec()`: describe one legend, for the `legends` style field
  
Additionally, `pmxhelpr` includes **style preset** functions which return a
`ggstylekit::style_spec()` object for each `plot_*()` function containing the 
defaults for the plotting family. These functions 

  + `style_dvtime()`/`plot_dvtime()`, 
  + `style_gof()`/`plot_gof()` , 
  + `style_dvconc()`/`plot_dvconc()`, 
  + `style_doseprop()`/`plot_doseprop()`, 
  + `style_vpc()` / `plot_vpc_cont()` / `plot_vpc_cens()` 
  
These style presents are passed to the `style` argument of plotting functions 
and use the following syntax:`style_<fn>(<map> = c(<role> = <value>))`.

Unique **roles** in `pmxhelpr` plot families can be passed to 
**per-series maps** in the style presents. These elements are overridden based 
on the **roles** specified, with remaining **roles** inheriting  the defaults.

```{r overview, eval=FALSE}
plot_dvtime(data_pk, dv_var = ODV, style = style_dvtime())
```



# Basics of Plot Styling with Style Specs


## Defaults

Call a style preset function with no arguments to print its defaults.

```{r view-defaults}
style_dvtime()
```


## Maps within the `style_spec`

The `style_spec()` object bundles several types of *maps*:

  + **Per-series maps**: apply to multiple plot layers specific to `pmxhelpr` 
  plot families:
  `colors`, `fill`, `linetypes`, `alphas`, `shapes`, `sizes`,`linewidths`
  
  + **Across-geom maps**: apply to an entire geom family: 
  `point_color`, `point_shape`, `point_size`, `point_shape`,
  `line_color`, `line_alpha`, `line_linetype`,`line_linewidth`, `line_fill`,
  `bar_fill`, `bar_color`, `bar_alpha`, `bar_linewidth`,
  `area_fill`, `area_color`, `area_alpha`, `area_linewidth`,
  `box_fill`, `box_color`, `box_alpha`, `box_linewidth`,
  `fill_alpha`
  
  + **Labels and axes**: apply to labels and axes:
  `title`, `xlabel`/`ylabel`, `xlims`/`ylims`, `logx`/`logy`, 
  `xbreaks`/`ybreaks`, `xminor_breaks`/`yminor_breaks`, 
  `xtick_labels`/`ytick_labels`, `xorder/yorder`, `equal_axis`.
  
  + **Legends**: apply to legends and can be set by a `legend_spec()` object: 
  `legends`,`legend.position`, `legend.title.position`,
  `legend_nrow`/`legend_ncol`, `legend.title.hjust`, `caption_hjust`
  
  + **Facets**: apply to facets:
  `facet`, `facet_scales`, `facet_nrow`/`facet_ncol`
  
  + **Theme**: apply a ggplot2 `theme` object.

`pmxhelpr` sets a shared house theme based on `ggplot2::theme_bw()`, with the 
all minor gridlines and major vertical gridlines removed.


## **Per-series map** Roles by plot family

Unique **roles** in each plot family style spec are as follows that can be 
passed to **per-series maps** are summarized below.

  + `style_dvtime()`: Roles include `obs_point`, `obs_line`, `cent_point`, 
`cent_line`, `cent_errorbar`, `ref_line`, `loq_line`
  + `style_gof()`: Roles include the elements in `style_dvtime`, plus the 
  overlay `colors` keyed by label (`OBS`, `DV`, `PRED`, `IPRED`)
  + `style_dvconc()`: Roles include `obs_point`, `ref_line`, `loess`, `linear` 
  + `style_doseprop()`: Roles include `obs_point`, `linear`
  + `style_vpc()`: Roles include `obs_point`, `obs_median_line`, `obs_pi_line`, 
    `sim_pi_line`, `sim_median_line`, `loq_line`, `sim_pi_ci`, `sim_pi_area`,
    `sim_median_ci` 

Available **roles** in **per-series maps** can be viewed from the style present
object using the `$` operator. 

```{r view-defaults-colors}
style_dvtime()$alphas
```

## Builder arguments versus style fields

A few aesthetic decisions belong to the plot builder rather than to the style,
because the builder needs them to decide *what to draw*, not just how to draw 
it. Where the two overlap, the builder argument wins:

  + **`log_y`**:  `plot_dvtime()`, `plot_dvconc()`, and `plot_gof()` set the 
  style `logy` from the plotting function `log_y` argument. This is because 
  `log_y` argument in the plotting function also controls the central tendency 
  summary measure (e.g., arithmetic versus geometric mean). 
  + **`errorbar_width`**: `plot_dvtime()` / `plot_gof()` set the errorbar cap 
  width as a function of the time range (default = 2.5% the max(NTIME)).
  + **`plot_doseprop()` axes and facets**: this plotting family applies its own
  log-log `scale_*_log10()` and per-metric `facet_wrap()`, so the style spec's
  `logx`/`logy`/`xbreaks`/`facet` fields are overwritten.
  + **Visibility**: `plot_gof_shown()` / `plot_vpc_shown()` control which 
  layers are *built* and the style spec controls how the built layers *look*. 
  Setting `style_vpc(alphas = c(obs_point = 0))` draws an invisible layer and 
  includes it in the legend whereas `plot_vpc_shown(obs_point = FALSE)` omits 
  it entirely. `shown` is the preferred method when a layer is not wanted.


## Examples

### Inline override

A style present can be passed and modified inline within the plotting function
call.

```{r ex-inline, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(
  data = data_pk, dv_var = ODV, col_var = `Dose and Food`,
  cent = "mean_sdl", log_y = TRUE, loq_method = 2,
  style = style_dvtime(alphas = c(obs_point = 0))
) +
  scale_x_continuous(breaks = seq(0, 168, 24))+
  labs(y = "Concentration (ng/mL)", x = "Time (hours)")
```

### Reusable style object

Often, one may be generating a series of plots with common styling. In this 
case, it is advantageous to build a style object once and reuse it across plots.
In addition to elements mapped to `ggplot2` aesthetic layers, the style spec 
object can be used to specify axis elements, labels, legends, and themes. 

```{r ex-reusable, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
new_dvtime_style <- style_dvtime(
  alphas     = c(obs_point = 0),
  xbreaks = seq(0, 168, 24),
  ybreaks = c(0.1, 1, 10, 100, 1000),
  ylims   = c(0.1, 2000),
  xlabel = "Time (hours)",
  ylabel = "Concentration (ng/mL)",
  title = "Single Ascending Dose"
)

plot_dvtime(data_pk, dv_var = ODV, col_var = `Dose and Food`,
            cent = "mean_sdl", log_y = TRUE, loq_method = 2,
            style = new_dvtime_style) 
```

### Error bar cap width

Error bar cap width is a builder argument, not part of the style, since it 
depends on the the values of the nominal time variable:

```{r ex-ebwidth, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(data_pk, dv_var = ODV, col_var = `Dose and Food`,
            cent = "mean_sdl", log_y = TRUE, loq_method = 2,
            style = new_dvtime_style, 
            errorbar_width = 10)
```


## Labeling with symbols using `byquote()`

The label keys `title`, `xlabel`, and `ylabel` accept `bquote()`, which is the 
practical way to symbols and subscripts into an label:

```{r ex-bquote, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(
  data = data_pd, dv_var = CFB, idv_var = CONC, ref = 0, loess = TRUE,
  style = style_dvconc(
    xlabel = bquote("Concentration" ~ (ng %.% mL^-1)),
    ylabel = bquote(Delta ~ "Response (%)")
  )
)
```



# Styling legends with `legend_spec()`


## Overview

The `legends` **map** accepts a `legend_spec()` object or list
of legend spec objects. A legend spec describes one legend, keyed either by the 
data column displayed (`col`) or by the plot **map** channel (`channel`). 

```{r ex-legend, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(
  data_pk, dv_var = ODV, col_var = `Food Status`, 
  dose_var = "DOSE", dosenorm = T,
  cent = "median", log_y = TRUE,
  style = style_dvtime(
    ylabel  = "Dose-normalized Concentration (ng/mL)",
    xlabel  = "Time (hours)",
    xbreaks = seq(0, 168, 24),
    legends = legend_spec(
      col = `Food Status`,
      title  = "Meal Status",
      labels = c(Fasted = "Fasted (overnight)", Fed = "Fed (high-fat)")
    ),
    legend.position = "bottom",
    legend_nrow     = 1
  )
)
```
Other `legend_spec()` arguments include:

  + `hide = TRUE` suppresses a legend while leaving its mapping in place
  + `order` sets which legend is drawn first when a plot has multiple
  + `nrow`/`ncol` override the `legend_nrow`/`legend_ncol` for the legend
  specified
  + `combine_with` folds a variable into an existing legend, when each key of 
  that legend identifies exactly one value of the folded column.


## Label overrides in legends and facets

A legend spec's `labels` also relabel matching facet strips, if the facetting
is specified in the style spec, rather than added as a new layer with 
`facet_wrap()`. 

```{r ex-legend-facet, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(
  data_pk, dv_var = ODV, col_var = `Food Status`, 
  dose_var = "DOSE", dosenorm = T,
  cent = "median", log_y = TRUE,
  style = style_dvtime(
    ylabel  = "Dose-normalized Concentration (ng/mL/mg)",
    xlabel  = "Time (hours)",
    xbreaks = seq(0, 168, 24),
    legends = legend_spec(
      col = `Food Status`,
      title  = "Meal Status",
      labels = c(Fasted = "Fasted (overnight)", Fed = "Fed (high-fat)")
    ),
    legend.position = "bottom",
    legend_nrow     = 1,
    facet = "Food Status",
    facet_nrow = 1,
    facet_ncol = 2,
    facet_scales = "fixed"
  )
) 
```

The original `labels` for the variable are retained in facet strips with
`facet_wrap()`. 

```{r ex-legend-facetwrap, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(
  data_pk, dv_var = ODV, col_var = `Food Status`, 
  dose_var = "DOSE", dosenorm = T,
  cent = "median", log_y = TRUE,
  style = style_dvtime(
    ylabel  = "Dose-normalized Concentration (ng/mL/mg)",
    xlabel  = "Time (hours)",
    xbreaks = seq(0, 168, 24),
    legends = legend_spec(
      col = `Food Status`,
      title  = "Meal Status",
      labels = c(Fasted = "Fasted (overnight)", Fed = "Fed (high-fat)")
    ),
    legend.position = "bottom",
    legend_nrow     = 1
  )
)  + 
  facet_wrap(~`Food Status`)
```


## Remove a legend

To drop a legend entirely, one can key the legend spec to the `channel`.

```{r ex-legend-hide, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(
  data_pk, dv_var = ODV, col_var = `Dose and Food`, cent = "none", id_var = ID,
  log_y = TRUE, 
  style = style_dvtime(
    ylabel  = "Concentration (ng/mL)",
    xlabel  = "Time (hours)",
    xbreaks = seq(0, 168, 24),
    legends = legend_spec(channel = "color", hide = TRUE)
  ) 
) + 
  facet_wrap(~`Dose and Food`)
```



# **Per-series maps** in `pmxhelpr` style presents


## Palette functions

A per-series map can be a palette `function(n)` rather than a named vector. 
`ggstylekit` calls the palette function with the unique number of groups in 
each plot, so one house style colors every plot without naming its groups.

```{r ex-palette, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
viridis_pal <- function(n) grDevices::hcl.colors(n, "Viridis")

plot_dvtime(
  data_pk, dv_var = ODV, col_var = `Dose and Food`, cent = "mean", log_y = TRUE,
  style = style_dvtime(colors = viridis_pal,
                       ylabel = "Concentration (ng/mL)", xlabel = "Time (hours)",
                       xbreaks = seq(0, 168, 24))
) + 
  facet_wrap(~PART)
```

Any `function(n)` returning `n` values works. A palette has no names to merge 
entry-wise, so it replaces the preset default map rather than merging onto it. 


## Trend line customization (`plot_dvconc()`)

Trend line colors are controlled using the **per-series map** colors with the 
`loess`/`linear` **roles**. The 95% CI ribbon is controlled by the `ggstylekit` 
default `line_fill`.

```{r ex-trend, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvconc(
  data = data_pd, dv_var = CFB, idv_var = CONC, ref = 0,
  loess = TRUE, se_loess = TRUE,
  style = style_dvconc(colors = c(loess = "darkred"), line_fill = "darkred")
) +
  labs(x = "Concentration (ng/mL)", y = "Response (% CFB)")
```

## GOF overlay colors (`plot_gof()`)

The DV (OBS), PRED, and PRED central tendency lines are the **role** keys for
the **per-series map** `colors`.

```{r ex-gof, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_gof(
  data = data_pkfit, dv_var = ODV, cent = "mean", log_y = TRUE,
  style = style_gof(colors = c(OBS = "grey60", DV = "black",
                               PRED = "#3388cc", IPRED = "firebrick"), 
                    xlabel = "Time (hours)", ylabel = "Concentration (ng/mL)", 
                    xbreaks = seq(0, 168, 24)))
```

## VPC plot color scheme (`plot_vpc_cont()`)

The **per-series maps** `colors` controls colors for points and lines while 
confidence/prediction interval ribbon fills are controlled by `fill`. 

```{r ex-vpc, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
sim <- df_mrgsim_replicate(data = data_pk, model = model, replicates = 100,
                           dv_var = ODV, carry_out = c("LLOQ", "FOOD"))

pcvpc_style <- style_vpc(
  colors = c(obs_point = "#000000", obs_median_line = "#000000",
             obs_pi_line = "#000000"),
  fill   = c(sim_median_ci = "#3388cc", sim_pi_ci = "#3388cc"),
  ylabel = "PRED-corrected Conc. (ng/mL)",
  xlabel = "Time (hours)",
  xbreaks = seq(0, 168, 24),
  xlim = c(0, 168)
)

plot_vpc_cont(sim, loq = 1, pcvpc = TRUE,
              style = pcvpc_style) + 
  scale_y_log10(guide = "axis_logticks")
```
## VPC plot color scheme (`plot_vpc_cont()`)

The VPC plotting functions do not generate a legend. `plot_vpc_legend()` builds 
a standalone legend plot intended to be combined with the VPC plot into a 
single figure. 

Pass the **same** style object along with the same `shown`, `ci`, and `pi` as
the plotting function to generate the corresponding legend. 

```{r ex-vpc-legend, fig.height=2.5, fig.width=4, message=FALSE, warning=FALSE}
plot_vpc_legend(style = pcvpc_style, lloq = 1)
```

# Updating the plot theme

The `theme` field takes any ggplot2 theme object and replaces the `pmxhelpr` 
house theme wholesale. Build on the house theme by adding to it rather than starting over:

```{r ex-theme, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_dvtime(
  data_pk, dv_var = ODV, col_var = `Dose and Food`,
            cent = "mean_sdl_upper",
  style = style_dvtime(
    ylabel = "Concentration (ng/mL)", xlabel = "Time (hours)",
    xbreaks = seq(0, 168, 24),
    theme  = theme_minimal(base_size = 8) + 
      theme(panel.grid.minor = element_blank(),
                   panel.grid.major = element_blank(),
                   legend.position  = "bottom")
  )
)
```

# Post-hoc styling: making aesthetic changes with `restyle_plot()`

`ggstylekit` not only handles plot styling *a priori*, it also handles plot
styling *a posteriori* after a finished plot object has been generated.

`restyle_plot()` merges style fields onto a finished plot. Restyling works 
analogously to the up front styling with style specs. 

For example, if one wanted to remove points and the legend from an existing 
plot, this can be accomplished as follows with `restyle_plot`.

```{r restyle, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
p <- plot_dvtime(data_pk, dv_var = ODV, cent = "median", log_y = TRUE, 
                 loq_method = 2,
                 style = style_dvtime(
                   xlabel = "Time (hours)", 
                   ylabel = "Dose-normalized Conc. (ng/mL/mg)",
                   xbreaks = seq(0, 168, 24)
                 ))
p

restyle_plot(p, 
             alphas = c(obs_point = 0), 
             legends = legend_spec(channel = "linetype", hide = TRUE))
```

# Post-hoc styling: surfacing new variables with `reveal()`

One of the most powerful features of `ggstylekit` is the `reveal()` function.

`reveal()` maps a variable present in the underlying plot dataset that is not
currently mapped to a graphical channel (or a facet). 

Use `as =` to inheret that channel's default values.

```{r reveal, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
reveal(p, `Food Status`, as = "colors")
```

Alternatively, the values can be supplied manually. 

```{r reveal-values, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
reveal(p, `Food Status`,
       colors = c(Fasted = "#3388cc", Fed = "firebrick"),
       labels = c(Fasted = "Fasted (overnight)", Fed = "Fed (high-fat)"))
```

Facets are also a channel that can be mapped like any other with `reveal`.

```{r reveal-facet, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
p2 <- reveal(p, `Food Status`, as = "facet", 
       colors = c(Fasted = "#3388cc", Fed = "firebrick"),
       labels = c(Fasted = "Fasted (overnight)", Fed = "Fed (high-fat)"))

p2
```


## Targeting specific layers with `on`

By default, `reveal()` maps the variable onto every layer that can display the
channel. The `on` argument narrows this to an **entity** — `"point"`, `"line"`,
`"bar"`, `"area"`, or `"box"` — or to a named **series**.

Here the central tendency line and its error bars are mapped to food status,
while the observed points are left unmapped.

```{r reveal-on-line, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
p_sdl <- plot_dvtime(data_pk, dv_var = ODV, cent = "mean_sdl", log_y = TRUE,
                     style = style_dvtime(xlabel = "Time (hours)",
                                          ylabel = "Concentration (ng/mL)",
                                          xbreaks = seq(0, 168, 24)))

reveal(p_sdl, `Food Status`, as = "colors", on = "line")
```

Error bars belong to the `"line"` entity, so they are summarized and colored
with the central tendency line they annotate.

Targeting `"point"` instead maps the observed and central tendency points,
leaving the central tendency line and error bars unmapped.

```{r reveal-on-point, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
reveal(p_sdl, `Food Status`, as = "colors", on = "point")
```

Values are supplied with `on` exactly as they are without it.

```{r reveal-on-values, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
reveal(p_sdl, `Food Status`, on = "line",
       colors = c(Fasted = "#3388cc", Fed = "firebrick"),
       labels = c(Fasted = "Fasted (overnight)", Fed = "Fed (high-fat)"))
```

`on` is also how a channel that is already in use is revealed onto. A plot built
with `col_var` already maps color, and revealing a second variable onto color
without `on` returns the plot unchanged, with a message naming the entities and
series available to target. Two variables sharing one channel are drawn in a
single legend, so revealing onto a free channel — `as = "shapes"`,
`as = "linetypes"`, `as = "facet"` — is usually the clearer choice.



# Combining plots with `combine_styled_plots`

`combine_styled_plots()` arranges several styled plots into one figure with a
**single collected legend**. 

Build the panels with a shared style so the collected legend is meaningful.

`ggstylekit::set_style()` returns an updated copy of a finished style, 
which is the tidy way to vary one field per plot panel

```{r combine-panels, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
panel_style <- style_dvtime(
  alphas = c(obs_point = 0.3),
  xlabel = "Time (hours)",
  ylabel = "Concentration (ng/mL)"
)

p_mean <- plot_dvtime(data_pk, dv_var = ODV, col_var = `Food Status`,
                      cent = "mean", log_y = TRUE,
                      style = set_style(panel_style, title = "Geometric mean"))

p_median <- plot_dvtime(data_pk, dv_var = ODV, col_var = `Food Status`,
                        cent = "median", log_y = TRUE,
                        style = set_style(panel_style, title = "Median"))
```

Then combine them. Unnamed arguments are the plots. Named arguments pass through
to `patchwork::wrap_plots()`, so `axes = "collect"` also de-duplicates the 
shared axis labels.

```{r combine, fig.height=4, fig.width=9, message=FALSE, warning=FALSE}
p_combo <- combine_styled_plots(p_mean, p_median, ncol = 2, axes = "collect")
p_combo
```

The combined figure is still a styled object, so `restyle_plot()` and `reveal()`
continue to work. Here the collected legend is re-titled and moved to the
bottom of the figure:

```{r combine-restyle, fig.height=5, fig.width=9, message=FALSE, warning=FALSE}
restyle_plot(p_combo, 
             legends = legend_spec(`Food Status`, title = "Meal Status",
                                   labels = c(Fasted = "Fasted (overnight)", 
                                              Fed = "Fed (high-fat)")),
             colors = c(Fasted = "#3388cc", Fed = "firebrick"),
             legend.position = "bottom")
```

`legend.position` moves the collected legend on the combined figure, as
`restyle_plot()` reaches the assembled figure and not only the individual
panels. Note that `labels` is not a style field: legend entries are relabeled
through the `labels` argument of `legend_spec()`, as above.

A combined plot can also be modified with the `&` operator in `patchwork` (or
`patchwork::plot_annotation(theme = )`). Make those changes **after**
`restyle_plot()`, as a plot modified with `+`/`&` from `patchwork` can no longer
be restyled.

Panels do not have to come from the same plot family; however, legends need 
to be **identical** and drawn from the same layer sets. Here concentration vs
time and response vs concentration panels colored by an indicator variable for
female sex (SEXF). 

The legend is suppressed in one plot in order to have only a single legend 
collected into the paneled combined plot, as even though the legends are 
visually identical, they are built from different layers.

```{r combine-families, fig.height=4, fig.width=9, message=FALSE, warning=FALSE}
p_pk <- plot_dvtime(data_pk, dv_var = ODV, col_var = SEXF,
                    dose_var = DOSE, dosenorm = TRUE,
                    cent = "mean", log_y = TRUE,
                    style = style_dvtime(xlabel = "Time (hours)",
                                         ylabel = "Dose-normalized Conc. (ng/mL/mg)", 
                                         xbreaks = seq(0, 168, 24),
                                         legends = legend_spec(channel = "color", 
                                                               hide = TRUE)))

p_pd <- plot_dvconc(data_pd, dv_var = CFB, idv_var = CONC, col_var = SEXF,
                    col_trend = TRUE, ref = 0, loess = TRUE,
                    style = style_dvconc(xlabel = "Concentration (ng/mL)",
                                         ylabel = "Response (% CFB)"))

combine_styled_plots(p_pk, p_pd, ncol = 2)
```

Two things to keep in mind when generated combined plots with
`combine_styled_plots`:

  + `combine_styled_plots()` **flattens** the input plot panels and discards 
  any patchwork annotations or layouts. Add titles and tags to the combined 
  plot returned with `patchwork::plot_annotation()`.
  + `guides = "collect"` is the default. Pass `guides` explicitly to override 
  it.

```{r combine-annotate, fig.height=4, fig.width=9, message=FALSE, warning=FALSE}
combine_styled_plots(p_pk, p_pd, ncol = 2) +
  plot_annotation(title = "Single ascending dose PK/PD", tag_levels = "A")
```
