---
title: "Visual Predictive Check Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visual Predictive Check Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r usethis, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

This vignette will demonstrate `pmxhelpr` functions for simulation-based visual predictive check (VPC) and prediction-corrected visual predictive check (pcVPC) evaluation of models fit to concentration-time data.  

```{r setup, echo=TRUE, warning=FALSE}
options(scipen = 999, rmarkdown.html_vignette.check_title = FALSE)
library(pmxhelpr)
library(dplyr, warn.conflicts =  FALSE)
library(forcats, warn.conflicts =  FALSE)
library(ggplot2, warn.conflicts =  FALSE)
library(mrgsolve, warn.conflicts =  FALSE)
library(withr, warn.conflicts =  FALSE)
library(patchwork, warn.conflicts = FALSE)
```

The continuous-range VPC plots in this vignette share a common axes / labels layer (time on x with breaks every 24 hours, concentration on a log10 y-axis). Defining a single list object lets each plot reuse it with `+ vpc_scales_labs`; pcVPC views override the y label with `+ labs(y = ...)` afterwards. The censored-range VPCs use a parallel `vpc_cens_scales_labs` for the BLQ-proportion plots.

```{r vpc-scales-labs}
vpc_scales_labs <- list(
  labs(x = "Time (hours)", y = "Concentration (ng/mL)"),
  scale_x_continuous(breaks = seq(0, 168, 24)),
  scale_y_log10(guide = "axis_logticks")
)

vpc_cens_scales_labs <- list(
  labs(x = "Time (hours)", y = "Proportion BLQ"),
  scale_x_continuous(breaks = seq(0, 168, 24))
)
```



# Data

The example dataset used in this vignette (`data_sad`) is based on a single ascending dose (SAD) study of an orally administered drug product with a parallel group food effect (FE) cohort. This vignette will assume familiarity with the `data_sad` internal dataset and the `plot_dvtime()` function, as described in the [Exploratory Analyses of PK Data](eda-pk-workflow.html) vignette. These elements will not be reviewed in detail in this vignette.

## data_sad

Dataset definitions can be viewed by calling `?data_sad`. A quick summary of variables and types is printed below.

```{r data-glimpse}
glimpse(data_sad)
```

PART specifies the two study cohorts:

  - Single Ascending Dose (SAD)
  - Food Effect (FE).

```{r data-part}
unique(data_sad$PART)
```

This dataset also contains an exact binning variable:

  - Nominal Time (NTIME).

This variable represents the nominal time of sample collection relative to first dose per study protocol whereas Actual Time (TIME) represents the actual time the sample was collected.

```{r data-ntime}
##Unique values of NTIME
unique(data_sad$NTIME)

##Comparison of number of unique values of NTIME and TIME
length(unique(data_sad$NTIME))
length(unique(data_sad$TIME))
```

Let's define some variables that may be useful in plotting and filter down to PK relevant records.

```{r data-factors, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
data_sad_pk <- data_sad %>% 
  filter(CMT %in% c(1,2)) %>% 
  mutate(`Food Status` = ifelse(FOOD == 0, "Fasted", "Fed"), 
         DoseFood = paste(DOSE, "mg", `Food Status`)) %>% 
  mutate(`Food Status` = var_addn(`Food Status`, ID),
         `Dose and Food` = var_addn(DoseFood, ID)) %>% 
  mutate(`Dose and Food` = fct_relevel(`Dose and Food`, "50 mg Fasted (n=6)", after = 1))
```



# PK Model

An example PK model (`pkmodel`) in `mrgmod` format is provided in the internal package library. This can be loaded using the helper function `model_mread_load()`, which wraps `mrgsolve::mread`.

```{r model-pk}
model <- model_mread_load("pkmodel")
```

We can take a look at the model code using `mrgsolve::see`

```{r model-see}
mrgsolve::see(model)
```

We can use this PK model to add population predictions (`"PRED"`) to our dataset with the function `df_mrgsim_addpred()`, which wraps `mrgsolve::mrgsim_df()` with `mrgsolve::zero_re()` to return an input dataset with population predictions appended. This function is called automatically within `df_mrgsim_replicate()` in the VPC Workflow; therefore, it does not need to be separately called for the VPC purpose.

This function takes `data.frame` and  `mrgmod` objects as input and an optional argument to specify the model variable to capture as `"PRED"` with random effects zeroed. The default is `"IPRED"`.

```{r df-mrgsim-addpred}
data_sad_pk_pred <- df_mrgsim_addpred(data_sad_pk, model, output_var = "IPRED")

colnames(data_sad_pk_pred)[!colnames(data_sad_pk_pred) %in% colnames(data_sad_pk)]
```

# VPC Workflow

 `pmxhelpr` includes two functions for calculating summary statistics across simulated replicates of an input dataset and plotting the output for model evaluation: `df_vpcstats` and `plot_vpc_cont()`.
 
The `df_vpcstats` function can be called within `plot_vpc_cont()`, for a one function call calculation and visualization of VPC statistics. `df_vpcstats` returns a `pmx_stats` container with `stats` (summary statistics of simulated and observed data), `obs` (observed data points for overlay), and `config` (run configuration) slots. The returned output can be passed directly into `plot_vpc_cont()` or the full pipeline can be run within the plotting function.

Many VPC tools use automatic binning algorithms (e.g., Jenks natural breaks, k-means, density) to determine bin intervals from the data. While these are useful for datasets without a pre-defined binning variable, they are not optimized to leverage input datasets with a variable representing exact bin times (e.g., nominal or protocol-specified times). Additionally, when a dataset contains an exact binning variable, these algorithms may not reproduce the exact bins. For example, absorption phase timepoints that are close together may be grouped into a single bin, obscuring important pharmacokinetic details.

The presence of exact bins in the data is a common scenario in pharmacometrics, as Clinical Study Protocols specify the times of PK sampling. `plot_vpc_cont()` uses the unique values of the nominal time variable directly as exact bins, ensuring that summary statistics are calculated at visualized at the protocol-specified sampling times.


# Running the simulation with `df_mrgsim_replicate()`

## Overview

`pmxhelpr` includes a function for running multiple replicates of a data set via simulation: `df_mrgsim_replicate()`

 `df_mrgsim_replicate()` is a wrapper function for `mrgsim_df()`, which uses `lapply()` to iterate the simulation over integers from 1 to the value passed to the argument `replicates`.

There are 3 required arguments to `df_mrgsim_replicate()`

  - `data`, a `data.frame` modeling analysis dataset
  - `model`, a `mrgmod` model object
  - `replicates`, numeric number of replicates to perform.
  
The are optional arguments specifying key dataset variables to be input into the simulation or captured in output. These include: 

  - `dv_var` = DV, dependent variable
  - `time_var` = TIME, actual time variable
  - `ntime_var` = NTIME, nominal time variable
  - `pred_var` = PRED, population prediction variable (fixed effects only)
  - `ipred_var` = IPRED, individual prediction variable (fixed + level 1 random effects)
  - `sim_dv_var` = DV, dependent variable captured in the simulated output (fixed + level 1 and 2 random effects)
  
These arguments use non-standard evaluation and can be passed as a bare column name or as a string.

We can pass `data_sad_pk` and `model` and run the simulation for 100 `replicates`. The names of actual and nominal time variables in `data_sad` match the default arguments; however, our dependent variable is named `ODV`, which must be specified in the `dv_var` argument, since it differs from the default (`DV`). 

```{r df_mrgsim_replicate}
simout <- df_mrgsim_replicate(
  data = data_sad_pk,
  model = model,
  replicates = 100,
  dv_var = ODV,
  carry_out = c("DOSE", "FOOD", "BLQ", "LLOQ"),
  recover  = c("PART", "DoseFood"))

glimpse(simout)
max(simout$SIM)
```

The simulation has appended some key variables with standardized names that will allow us to take advantage of default arguments in later pipeline functions:
  - `PRED` (population prediction)
  - `IPRED` (individual prediction)
  - `SIMDV` (simulated dependent variable)
  - `OBSDV` (observed dependent variable).
  
These variable standard names are output regardless of the input name. For example, we can see from the `glimpse()` output that the simulation also outputs `"Y"`, which represents a simulated version of DV. We will still receive the same standardized output variable names if we change argument form `sim_dv_var`, as the variables are renamed to VPC workflow standards on output

```{r df_mrgsim_replicate-y}
simout <- df_mrgsim_replicate(data = data_sad_pk,
                     model = model,
                     replicates = 100,
                     dv_var = ODV,
                     sim_dv_var = Y,
                     carry_out = c("DOSE", "FOOD", "BLQ", "LLOQ"),
                     recover  = c("PART", "DoseFood"))

glimpse(simout)
```

## Changing the Replicate Counter

The maximum value of our replicate count variable (default = `"SIM"`) indicates that the dataset has been replicated 100 times. The variable name output for the count of replicates can be specified using the `irep_name` argument.

This argument uses non-standard evaluation and can be supplied as a bare name or string

```{r df_mrgsim_replicate-irep}
simout <- df_mrgsim_replicate(data = data_sad_pk,
                     model = model,
                     replicates = 100,
                     dv_var = ODV,
                     irep_name = IREP,
                     carry_out = c("DOSE", "FOOD", "BLQ", "LLOQ"),
                     recover  = c("PART", "DoseFood"))

max(simout$IREP)
```


## Carrying Input Columns to the Output

`df_mrgsim_replicate()` does not auto-carry input columns. Use `carry_out` and `recover` (passed through `...` to `mrgsim_df()`) to list the columns you want propagated to the output:

  - `carry_out`: character vector of numeric input column names to carry to the output. Best for numeric columns (`DOSE`, `FOOD`, `BLQ`, `LLOQ`, ...).
  - `recover`: character vector of input column names of any type to restore to the output. Best for character / factor columns (`PART`, `DoseFood`, ...).

The always-carried set (`EVID`, `MDV`, `CMT`, `TIME`, `NTIME`, `OBSDV`, `PRED`) is appended internally so the wrapper's standardized output structure is preserved regardless of what you pass.

```{r df_mrgsim_replicate-capture}
simout <- df_mrgsim_replicate(data = data_sad_pk,
                     model = model,
                     replicates = 100,
                     dv_var = ODV,
                     carry_out = c("DOSE", "FOOD", "BLQ", "LLOQ"),
                     recover  = c("PART", "DoseFood"))

colnames(simout)
```

## Setting the seed

The unique seed can be set using the `seed` argument. The default is `1234567889`.

```{r df_mrgsim_replicate-seed}
simout <- df_mrgsim_replicate(data = data_sad_pk,
                     model = model,
                     replicates = 100,
                     dv_var = ODV,
                     seed = 42,
                     carry_out = c("DOSE", "FOOD", "BLQ", "LLOQ"),
                     recover  = c("PART", "DoseFood"))

glimpse(simout)
```

## Additional arguments passed to `mrgsim_df()`

Any other `mrgsim_df()` argument can be passed via `...`. One useful argument is `obsonly = TRUE`, which removes dose records from the simulation output to reduce file size.

```{r df_mrgsim_replicate-obsonly}
# Default behavior (dose records included) — simout from the prior chunk
nrow(simout)

# With obsonly = TRUE, dose records are removed
simout <- df_mrgsim_replicate(data = data_sad_pk,
                              model = model,
                              replicates = 100,
                              dv_var = ODV,
                              carry_out = c("DOSE", "FOOD", "BLQ", "LLOQ"),
                              recover  = c("PART", "DoseFood"),
                              obsonly = TRUE)

nrow(simout)
```

## Parallel processing

For large `replicates` values or expensive model evaluations, `df_mrgsim_replicate()` accepts `parallel = TRUE` to dispatch per-replicate simulations across worker processes via `future.apply::future_lapply()`. This requires the `future.apply` package and a parallel plan set by the user. `df_mrgsim_replicate()` does not modify the plan itself, so the user retains full control over the worker topology and is responsible for restoring the prior plan after the call.

The pattern is to wrap the parallel simulation between a `future::plan()` setup and a teardown back to `future::sequential`:

```{r df_mrgsim_replicate-parallel, eval=FALSE}
future::plan(future::multisession, workers = 4)

simout <- df_mrgsim_replicate(data = data_sad_pk,
                     model = model,
                     replicates = 1000,
                     dv_var = ODV,
                     carry_out = c("DOSE", "FOOD", "BLQ", "LLOQ"),
                     recover  = c("PART", "DoseFood"),
                     parallel = TRUE)

future::plan(future::sequential)
```

Under `parallel = TRUE`, per-replicate RNG streams are generated from `seed` using L'Ecuyer-CMRG (via the `future.seed = seed` argument to `future_lapply()`), so output is reproducible given the same `seed` and `future::plan()`. The numerical output differs from `parallel = FALSE` because the RNG mechanism differs, but the two are statistically equivalent — the same per-replicate distribution is sampled, only the per-record realizations change. As a consequence, downstream `df_vpcstats()` summary statistics (quantiles, BLQ proportions, simulated CI bands) will not match bit-for-bit between the sequential and parallel paths even when the seed is held constant; both are valid Monte Carlo samples of the same target distribution.

Speedup is best for high `replicates` (the per-worker dispatch overhead amortizes over many replicates) and models with expensive numerical integration. For modest workloads — e.g., the 100-replicate examples used throughout this vignette — the sequential path is typically faster end-to-end because worker-startup cost dominates the simulation cost.


# Calculating the summary statistics with `df_vpcstats()`

`pmxhelpr` contains a function to pre-process and derive summary statistics from the observed and simulated data for plotting in downstream VPC and pcVPC plots to evaluate the model fit for longitudinal, continuous repeated measures data: `df_vpcstats()` 

This function is designed around the concept of actual and nominal time variables, with the latter used to bin data together to calculate summary statistics across simulation replicates. Functionality is included for handling of data missing due to assay sensitivity (BLQ) separately depending on user request of standard or prediction-corrected values.

## Overview of `df_vpcstats()`

The `df_vpcstats` function handles the first two steps in the VPC plot pipeline after simulation:
  
  1. Pre-processing the data to validate arguments, filter to observation records, and handle missing data
  2. Grouping, prediction-correction (if requested), and calculation of summary statistics
  
VPC summary statistics are calculated for the input dataset `data` at the study replicate-level (median and quantiles specified by `pi`), and subsequently summarized across replicates (median and uantiles specified by `ci`) to identify the non-parametric confidence interval. 

Outputs are grouped by the variables passed to `ntime_var` and `strat_var`. 

The remaining arguments control processing for prediction-correction (`pcvpc`) and BLQ handling (`loq`).

`df_vpcstats()` returns a [`pmx_stats`][is_pmx_stats] container with three slots:

  + `stats` — the quantile summary statistics of the observed and simulated data
  + `obs` — first-replicate observation rows used as the scatter overlay in `plot_vpc_cont()`
  + `config` — the run configuration (`n_replicates`, `loq`, `strat_var`) used by downstream plot builders

Let's calculate VPC summary statistics for our simulated dataset. The minimum stratification based on extrinsic participant factors in the study design for a standard vpc is stratifying by Dose and Food.

```{r df-vpcstats-dose-food, warning=FALSE}
vpcstats_ntime_dose_food <- df_vpcstats(
  data = simout,
  strat_var = DoseFood
)

glimpse(vpcstats_ntime_dose_food$stats)
```

A helpful informational message prints to let us know that the `loq` argument was automatically inherited from our input dataset passed to `data` since it contained a variable named `LLOQ`.

`df_vpcstats()` always emits **both** standard-VPC and prediction-corrected (pcVPC) summary statistics in a single call. The standard-flavor columns are unprefixed; the prediction-corrected counterparts carry a `pc_` prefix. Selecting which view to render is a plot-time decision via `plot_vpc_cont(out, pcvpc = TRUE/FALSE)`.

The `stats` element contains the following summary statistics:

  + `obs_n / obs_n_blq / obs_prop_blq`: count of observations (EVID = 0) in each bin, count of BLQ-encoded observations (MDV=1, OBSDV<loq, is.na(OBSDV)), and the BLQ proportion. Single (not duplicated under `pc_*`) — these are row counts, unaffected by prediction-correction.
  + `sim_low_low / sim_low_med / sim_low_hi`: standard-flavor quantiles of the lower `pi` quantile (`pi_low`) across replicates. Suffix `_low/_med/_hi` refers to the lower/upper `ci` bound and the median.
  + `sim_med_low / sim_med_med / sim_med_hi`: standard-flavor quantiles of the simulated median across replicates.
  + `sim_hi_low / sim_hi_med / sim_hi_hi`: standard-flavor quantiles of the upper `pi` quantile (`pi_hi`) across replicates.
  + `obs_low / obs_med / obs_hi`: standard-flavor observed quantiles in each bin.
  + `sim_prop_blq_low / sim_prop_blq_med / sim_prop_blq_hi`: standard-flavor quantiles of the proportion of simulated data < `loq` across replicates. Std-only — LOQ has no meaning on the prediction-corrected scale, so the pc flavor does not emit a `pc_sim_prop_blq_*` set.
  + `pc_obs_low/med/hi`, `pc_sim_low_low/med/hi`, `pc_sim_med_low/med/hi`, `pc_sim_hi_low/med/hi`: prediction-corrected counterparts of the standard observed and simulated quantile groups.
  + `ci`: confidence interval width (run config; constant within result).
  + `pi_low / pi_hi`: lower and upper prediction-interval probabilities (run config; constant within result).
  

## Pre-processing of Data for VPCs

*NOTE* The VPC Workflow functions work best when used in sequence as a tool chain. `df_vpcstats` is designed to work with an input simulated dataset generated by `df_mrgsim_replicate`. No pre-processing of the simulated output is required before passing directly into `df_vpcstats`. Indeed, data processing to handle BLQ imputation and prediction-corrected is built into the `df_vpcstats` pre-processing function pipeline; therefore, the user should pass in a simulated dataset that includes all observations collected and analyzed regardless of assay result (e.g., `MDV=0` or `MDV=1`).

The handling of BLQ imputation for derivation of quantiles within and across replicates is controlled by the `mode` argument. Options include:

  + `"auto"` (default) apply rank when `pcVPC = FALSE` and drop when `pcVPC = TRUE`
  + `"rank"` include BLQ data in quantile calculations
  + `"drop"` drop BLQ data in quantile calculations

The default method recommended by `pmxhelpr` is `"auto"`, which applies the `"rank"` method for standard VPCs and the `"drop"` method for pcVPCs. The other options are included for comparability and reproducibility across other VPC plotting functions and to give the user control on handling. Under the hood, this option controls whether missing values are inputed to -Inf (rank) or NA (drop) prior to calculation of quantile summary statistics, which drop missing values with `na.rm = TRUE`. 

For standard VPCs without prediction-correction, both the observed and simulated data are handled with `"rank"`; however, no BLQ handling is applied to the simulated data and rankings are based on raw simulated values without BLQ censoring to illustrate the underlying "true" profile based on the model.

We can see that specifying `mode = "drop"` results in differential summary statistics from the default `mode = "auto"` in 15 of the 114 rows in the output `stats` data.frame.

```{r df-vpcstats-dose-food-drop, warning=FALSE}
vpcstats_ntime_dose_food_drop <- df_vpcstats(
  data = simout,
  strat_var = DoseFood,
  mode = "drop"
)
nrow(vpcstats_ntime_dose_food_drop$stats)
anti_join(vpcstats_ntime_dose_food$stats, vpcstats_ntime_dose_food_drop$stats) %>% nrow()
```


## Prediction-correction

Prediction-correction is applied internally as part of every `df_vpcstats()` call; therefore, the prediction-corrected columns (`pc_*`) appear alongside the standard columns in the output. 

The prediction-correction pipeline censors both observed and simulated values at the `loq` prior to prediction-correction (when `loq` is known) and then calculates quantile summary statistics on the continuous, quantifiable range. If `loq` is unknown, missing observations (`MDV == 1`) are excluded from quantile summary statistics for both the observed and simulated data.

```{r df-vpcstats-pc-part, warning=FALSE}
vpcstats_ntime_part <- df_vpcstats(
  data = simout,
  strat_var = PART
)
```

## Inspecting the `vpc_stats` object

The container returned by `df_vpcstats()` is class-tagged `c("vpc_stats", "pmx_stats")`, which provides three S3 methods designed for interactive inspection without dumping every column.

`print()` shows a focused summary — object dimensions, run-config values (`n_replicates`, `loq`, `strat_var`), the column groups present in `stats` (counts, simulated-BLQ proportions, standard and prediction-corrected observed/simulated quantiles, run-config metadata), and a short head preview.

```{r vpc-stats-print}
print(vpcstats_ntime_dose_food)
```

`summary()` returns the same content as `print()` but suppresses the head preview, which is convenient for snapshotting the run configuration without the row content.

```{r vpc-stats-summary}
summary(vpcstats_ntime_dose_food)
```

`as.data.frame()` returns the `stats` slot as a plain data.frame. Access `obs` and `config` separately via `vpcstats_ntime_dose_food$obs` and `vpcstats_ntime_dose_food$config`.

```{r vpc-stats-as-df}
glimpse(as.data.frame(vpcstats_ntime_dose_food))
```

Use `is_vpc_stats()` to test class membership programmatically — for example, in user code that should accept either a precomputed `vpc_stats` object or a raw simulated data.frame. Pass `strict = TRUE` to additionally validate structural integrity.

```{r vpc-stats-predicate}
is_vpc_stats(vpcstats_ntime_dose_food)
is_vpc_stats(vpcstats_ntime_dose_food$stats)  # FALSE -- that's the inner data.frame
```


# VPC Plots with `plot_vpc_cont()`

`pmxhelpr` contains a function to produce VPC and pcVPC plots for models fit to longitudinal, continuous repeated measures data: `plot_vpc_cont()`

This function is designed around the concept of actual and nominal time variables, with the latter used to bin data together to calculate and display summary statistics across simulation replicates.

`plot_vpc_cont` has 1 required argument: `data`, the input simulated dataset for the VPC procedure. Standard VPCs require stratification by dose, and other extrinsic study conditions, for interpretability; therefore, we will also specify `strat_var=DoseFood` to also stratify our plots by this variable.

```{r plot-vpc, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_vpc_cont(
  data = simout,
  strat_var = DoseFood
) +
  vpc_scales_labs
```

The default for standard VPCs in `pmxhelpr` is to censor observed quantiles at the `loq` when it is known without censoring simulated values. The `loq` is inherited from the input dataset automatically since the variable `"LLOQ"` is present in `data`.

## Prediction-correction

Under `pcvpc = TRUE`, BLQ censoring is still applied (before prediction-correction, to both observed and simulated data), but the LOQ reference line is not drawn since `loq` has no meaning on the prediction-corrected scale.

```{r plot-vpc-pc, fig.height=4, fig.width=6, warning=FALSE}
plot_vpc_cont(
  data = simout,
  pcvpc = TRUE,
  strat_var = PART
) +
  vpc_scales_labs +
  labs(y = "Pred-corrected Conc. (ng/mL)")
```

## Stratifying plots

The other option to prediction-correction is to stratify by confounding factors. Dose is generally the minimum stratification for a VPC plot. The variable for stratification *MUST* be passed to the argument `strat_var` and not added later only as a `facet_wrap`/`facet_grid` call to the returned object. `plot_vpc_cont()` returns a `pmx_vpc_plot` object that emits a warning when a `facet_wrap()` or `facet_grid()` layer is added with `+`, calling out that the pre-computed VPC statistics will be inconsistent with the new panels.

Let's take a look at our prediction-corrected VPC stratified by Food Status.

```{r plot-vpc-pc-food, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_vpc_cont(
  data = mutate(simout, FoodStatus = ifelse(FOOD == 1, "Fed", "Fasted")),
  pcvpc = TRUE,
  strat_var = FoodStatus
) +
  vpc_scales_labs +
  labs(y = "Pred-corrected Conc. (ng/mL)")
```

*NOTE* Because `plot_vpc_cont` performs BOTH the summary statistic calculation and the plotting, one cannot simply add a `facet_wrap` layer with the stratification after calling `plot_vpc_cont`. This will result in the binned observed quantiles and simulated intervals plotted reflecting the unstratified data across all conditions with only the observed points stratified.

Simulated and observed intervals in the plot below are *INCORRECT* and reflect the full dataset, not the `DoseFood` stratification. Only the observed points are correctly stratified, as they are added to the final plot layer. The warning printed above the plot is emitted by the `+.pmx_vpc_plot` method as the `facet_wrap()` layer is added.

```{r plot-vpc-dosefood-facet, fig.height=4, fig.width=6, message=TRUE, warning=TRUE}
plot_vpc_cont(
  data = simout
) +
  facet_wrap(~DoseFood) +
  vpc_scales_labs
```

## Dropping small sample bins

There may be nominal time bins that contain only a small number of observations that skew the simulated intervals. `plot_vpc_cont()` includes the argument `min_bin_count` (default = 1), which filters out exact bins with fewer quantifiable observations than the minimum set by this argument. Importantly, the observed data points in these small bins are *still plotted*; however, they do not influence the calculation of summary statistics or summary plot elements (shaded intervals, lines). This provides the greatest fidelity to the data visualized without introducing visual artifacts due to small sample timepoints.

When setting `min_bin_count = 10`, summary statistics are not plotted for the final two timepoints containing fewer than 10 quantifiable observations; however, the observations themselves are still plotted.

```{r plot-vpc-min-bin-count, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
plot_vpc_cont(
  data = simout,
  pcvpc = TRUE,
  min_bin_count = 10
) +
  vpc_scales_labs +
  labs(y = "Pred-corrected Conc. (ng/mL)")
```

## BLQ handling

BLQ data are often ignored in population PK modeling (i.e., set as missing; M1 method) in population. This is unlikely to introduce appreciable bias in model parameter estimation, provided the absolute frequency of BLQ data is low (e.g., <10-20%) and the pattern of BLQ censoring is in line with the expectations by dose and time.

However, the handling of BLQ data in VPCs may impact the interpretation of these graphical model diagnostics.

Let's generate some VPC plots to demonstrate the impact of different approaches to BLQ handling to the visual assessment of the adequacy of model fit and the true underlying PK profile. We will continue using the `simout` object generated above.

Let's start by focusing on evaluating the model fit of the 100 mg dose level we have been exploring across Part1-SAD and Part2-FE.

```{r sim-input}
sim100 <- simout %>%
  filter(DOSE == 100)
```

A common approach to plotting VPCs, especially pcVPCs, is to exclude all `MDV=1` observations. Let's plot using `plot_vpc_cont()` filtering out `MDV=1` in the input dataset to exclude observed BLQ timepoints.

```{r plot-vpc-blq-drop, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
plot_vpc_cont(
  data = filter(sim100, MDV == 0),
  pcvpc = TRUE,
  strat_var = PART
) +
  vpc_scales_labs
```

The observed median appears consistent with the exploratory concentration-time profiles generated earlier excluding data below the lower limit of quantification (i.e., plotting `"ODV"`). Notice that the increasing trend in the observed quantiles at later timelines coincides with completely overlapping simulated confidence intervals for all quantiles. This is due to the decreasing sample size with time due to exclusion of timepoints where `MDV=1`.

Let's instead try specifying a new argument `loq` to `plot_vpc_cont()`.

`loq` is the numeric value of the lower limit of quantification in the units of the dependent variable. When specified, observed quantiles are computed using censored quantile estimation, where values below the LLOQ (including missing values) are treated as left-censored. If the quantile of the observed data falls below the LLOQ, it is returned as `NA`.

```{r plot-vpc-blq-obs-cens, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
vpc <- plot_vpc_cont(
  data = sim100,
  strat_var = PART,
  loq = 1
) +
  vpc_scales_labs

vpc
```

Now we see a red horizontal line depicting the LLOQ (1 ng/mL). Notice also the increase in the y-axis range, the change in the shape of the observed quantiles, and the greater resolution in separation between the confidence intervals.

Specifying `loq` is the preferred method for plotting VPC diagnostics with the `pmxhelpr` package. This is not the default method due to the all-too-common case where the assay LLOQ is not known to the analyst.

`loq` may also be passed for prediction-corrected VPCs (`pcvpc = TRUE`). In that case, observed and simulated values below `loq` are censored to `NA_real_` before prediction-correction so that both data streams are treated symmetrically, and no LOQ reference line is drawn (the LOQ has no meaning on the prediction-corrected scale).

The unique vector of `LLOQ` values is used automatically if the variable `"LLOQ"` is present in the input dataset. This allows for the scenario in which there are multiple unique LLOQ values within a single pooled analysis dataset.

```{r plot-vpc-blq-obs-cens2, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
plot_vpc_cont(
  data = sim100,
  strat_var = PART
) +
  vpc_scales_labs
```

## Pooled Data with Multiple LLOQs

A common real-world scenario is a bioanalytical assay update mid-development: an early cohort runs on the original assay (e.g., LLOQ = 1 ng/mL) and a later cohort runs on an updated assay (e.g., LLOQ = 2 ng/mL). When the analysis dataset pools both, the `LLOQ` column carries different values across rows. `pmxhelpr` resolves `loq` per row so each observation is censored at its own threshold; the unique non-NA values are surfaced via `config$loq` for plotting.

The example below mimics this scenario by re-coding `LLOQ = 2` for the Part 2 food-effect cohort while leaving Part 1 at the original `LLOQ = 1`.

```{r sim-multi-loq, message=FALSE, warning=FALSE}
sim100_multi_loq <- sim100
sim100_multi_loq$LLOQ[sim100_multi_loq$PART == "Part 2-FE"] <- 2
distinct(sim100_multi_loq, PART, LLOQ)
```

Calling `df_vpcstats()` on this pooled dataset emits the per-row inheritance message and stores the sorted unique values in `config$loq`. The row-aligned threshold remains accessible on `out_multi$obs$LOQ` for any downstream code that needs the per-observation value.

```{r vpc-stats-multi-loq, message=TRUE, warning=FALSE}
out_multi <- df_vpcstats(sim100_multi_loq, strat_var = PART)
out_multi$config$loq
```

`plot_vpc_cont()` consumes the precomputed stats and draws one dashed reference line per unique LLOQ value.

```{r plot-vpc-multi-loq, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
vpc_multi <- plot_vpc_cont(out_multi) + vpc_scales_labs
vpc_multi
```

Each facet renders only the LLOQ(s) applicable to that strat-level: Part 1-SAD shows the LLOQ = 1 line and Part 2-FE shows LLOQ = 2. The ref-line set is recomputed at plot time from the row-aligned `obs$LOQ` column grouped by `strat_var`, so the visualization stays in sync with whatever per-row inheritance produced. The legend (built from `out_multi$config$loq`) still summarizes the analysis-wide set of LLOQs, since the legend is a figure-level annotation rather than a facet-level one.

`plot_vpc_legend()` accepts the same vector and registers one entry per unique LLOQ value.

```{r vpc-multi-loq-legend, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_multi_legend <- plot_vpc_legend(lloq = out_multi$config$loq)
vpc_multi_legend
```

Composing the plot and legend with `patchwork` yields a single figure where both LLOQ thresholds are simultaneously labeled.

```{r plot-vpc-multi-loq-w-legend, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
vpc_multi + vpc_multi_legend + plot_layout(heights = c(2.5, 1))
```

Passing `out_multi$config$loq` directly to `plot_vpc_legend()` keeps the legend in sync with whatever LLOQs `df_vpcstats()` discovered, with no need to hardcode the value set at the call site.

## Reusing precomputed stats

`plot_vpc_cont()` also accepts the list returned by `df_vpcstats()` directly. This skips the preprocess + compute steps and is useful when you want to inspect the summary statistics and then plot the same data multiple times — for example, flipping between the standard and prediction-corrected views, or rendering with a different `min_bin_count`, `shown`, or `theme` — without paying the summarization cost on every call.

The `df_vpcstats()` return is class-tagged `"vpc_stats"` (see [Inspecting the `vpc_stats` object](#inspecting-the-vpc_stats-object) above for `print()` / `summary()` / `as.data.frame()` / `is_vpc_stats()` usage). Use `out$stats` and `out$obs` to access the underlying frames directly.

```{r plot-vpc-precomputed, fig.height=4, fig.width=6, warning=FALSE}
out <- df_vpcstats(data = simout, strat_var = DoseFood)

# Standard VPC view from the precomputed result
plot_vpc_cont(out, pcvpc = FALSE) + vpc_scales_labs

# Prediction-corrected view from the same precomputed result
plot_vpc_cont(out, pcvpc = TRUE) +
  vpc_scales_labs +
  labs(y = "Pred-corrected Conc. (ng/mL)")
```

When `data` is a precomputed `df_vpcstats()` result, the pipeline arguments (`strat_var`, `loq`, `mode`, `pi`, `ci`, column-name args) **cannot** be honored — those decisions were already made when `df_vpcstats()` ran, and re-passing them on the cached path would silently shadow the original values. Passing any of them aborts with a message pointing the caller back at `df_vpcstats()`. Only plot-only arguments (`min_bin_count`, `show_rep`, `shown`, `theme`, `pcvpc`) are accepted on this path; to change a pipeline setting, re-run `df_vpcstats()` and pass the new result.

`plot_vpc_cont` censors quantiles of the observed data displayed as lines when the corresponding quantile of observations is BLQ. This is done automatically if the variable `LLOQ` is present in the input simulated dataset.

## Adjusting the elements displayed

The default elements shown in the plot are controlled by the `shown` argument in `plot_vpc_cont()`.

The default is as follows:

```{r vpc-plot-shown}
plot_vpc_shown()
```

The components of the list correspond to the following vpc plot elements:

-   Observed points: `obs_point`
-   Observed quantile lines: `obs_pi_line`
-   Observed median line: `obs_median_line`
-   Simulated prediction interval lines: `sim_pi_line`
-   Simulated prediction interval CI: `sim_pi_ci`
-   Simulated prediction interval area: `sim_pi_area`
-   Simulated median line: `sim_median_line`
-   Simulated median CI: `sim_median_ci`

One or more elements to be updated from the defaults above can be passed as a list to the argument `shown`. Any elements not specified in `shown` will inherit the defaults.

For example, the 90% prediction interval (i.e., 5th to 95th percentiles) can be visualized in place of CIs of each of the 5th and 95th percentiles with the observed and simulated medians shown as follows:

```{r plot-vpc-pcvpc-90pi, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
plot_vpc_cont(
  data = simout,
  pcvpc = TRUE,
  min_bin_count = 5,
  shown = plot_vpc_shown(obs_pi_line = FALSE, sim_pi_ci = FALSE, sim_median_ci = FALSE, sim_median_line = TRUE, sim_pi_area = TRUE)
) +
  vpc_scales_labs +
  labs(y = "Pred-corrected Conc. (ng/mL)")
```

Now, let's say we want to remove the observed data points from the plot above to better visualize the observed quantile lines relative to their corresponding simulated confidence intervals. This can be accomplished with `shown`.

```{r plot-vpc-pcvpc-noobs, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
plot_vpc_cont(
  data = simout,
  pcvpc = TRUE,
  min_bin_count = 5,
  shown = plot_vpc_shown(obs_point = FALSE)
) +
  vpc_scales_labs +
  labs(y = "Pred-corrected Conc. (ng/mL)")
```

We could also take this one step further and only look at the median and the simulated confidence interval of the median, to closely interrogate central tendency. This is common for VPC strata which have few observations, leading to inadequate sample size to discriminate between the confidence intervals of the median and the extremes. This is common scenario when evaluating VPC plots stratified by individual study arms in early phase trials.

```{r plot-vpc-pcvpc-noobs-medonly, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
plot_vpc_cont(
  data = simout,
  pcvpc = TRUE,
  min_bin_count = 5,
  shown = plot_vpc_shown(obs_point = FALSE, obs_pi_line = FALSE, sim_pi_ci = FALSE)
) +
  vpc_scales_labs +
  labs(y = "Pred-corrected Conc. (ng/mL)")
```

## Building plots from a `vpc_stats` object directly

For downstream code or custom workflows that produce a `vpc_stats`-shaped object outside `df_vpcstats()` (e.g. external preprocessing, snapshot fixtures), `plot_build_vpc()` is exported as the public renderer. It is the same engine that `plot_vpc_cont()` uses internally and accepts the same plot-only arguments (`min_bin_count`, `show_rep`, `shown`, `theme`, `pcvpc`). `strat_var` and `loq` inherit from the container's `$config` slot when not passed explicitly.

```{r plot-build-vpc-direct, fig.height=4, fig.width=6, warning=FALSE}
plot_build_vpc(out, pcvpc = FALSE) + vpc_scales_labs
```


# Adjusting the Plot Theme with `plot_vpc_theme()`

The default aesthetics for VPC plots are controlled via `plot_vpc_theme()`. See the [Plot Themes and Aesthetics](plot-themes.html) vignette for details on the theme system, element constructors, and examples of customizing VPC aesthetics.

```{r vpc-theme}
plot_vpc_theme()
```

We can define an alternative theme using a blue / grey color schema using this function and the constructor helpers.

```{r vpc-new-theme}
vpc_new_theme <- plot_vpc_theme(
  obs_point = pmx_point(color = "#000000"),
  obs_median_line = pmx_line(color = "#000000"),
  obs_pi_line = pmx_line(color = "#000000"),
  sim_median_ci = pmx_ribbon(fill = "#3388cc"),
  sim_pi_ci = pmx_ribbon(fill = "#3388cc")
)
```

Regenerating our BLQ quantile censored plot stratified by part with the new theme yields the following plot

```{r plot-vpc-blq-obs-cens-newtheme, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
vpc2 <- plot_vpc_cont(
  data = sim100,
  strat_var = PART,
  loq = 1,
  theme = vpc_new_theme
) +
  vpc_scales_labs

vpc2
```


## Modifying Axes and Labels

Because `plot_vpc_cont()` returns a `ggplot2` object, the y-axis can be transformed to log10 scale by adding `scale_y_log10()` as a layer. This approach avoids any data transformation prior to calculating summary statistics, ensuring all quantile calculations are performed on the original scale.

Similarly, the axis labels can be controlled with `labs` and the x-axis breaks can be set to more reasonable values for time units using `scale_x_continuous()`.


# VPC Plots with `plot_vpc_cens()`

`plot_vpc_cens()` is the companion diagnostic to `plot_vpc_cont()`. Where `plot_vpc_cont()` evaluates whether the model reproduces the dynamic range of the data above the LOQ, `plot_vpc_cens()` evaluates whether it reproduces the proportion of BLQ observations over time. It plots `obs_prop_blq` (the per-bin observed BLQ proportion) and the non-parametric confidence interval of `sim_prop_blq` across replicates. 

Both quantities are already computed by `df_vpcstats()` whenever a LOQ source (scalar `loq` or `LLOQ` column) is available, so `plot_vpc_cens()` is a thin plotting wrapper over the same VPC summary pipeline used by `plot_vpc_cont()`.

`plot_vpc_cens()` is standard-VPC only, as LOQ has no meaning on the prediction-corrected scale, and the `pcvpc` argument is omitted. A LOQ source is required (scalar `loq` or an `LLOQ` column inherited from `data`).

Like `plot_vpc_cont()`, `plot_vpc_cens()` also accepts a pre-computed list returned by `df_vpcstats()` directly. This workflow adds efficiencies for cases where a user wants to evaluate the continuous and censored ranges with the same binning.

```{r plot-vpc-cens, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
plot_vpc_cens(
  data = sim100,
  strat_var = PART,
  loq = 1
) +
  vpc_cens_scales_labs
```

The ribbon is the non-parametric CI of the simulated BLQ proportion across replicates (controlled by `ci`), and the solid red line and points show the observed proportion per bin. A well-fitting model places the observed line inside the simulated CI band. The simulated median line is hidden by default. It can be enable it via `shown = plot_vpc_shown(sim_median_line = TRUE)`.

`plot_vpc_cens()` reuses `plot_vpc_shown()` and `plot_vpc_theme()` — only the keys `obs_point`, `obs_median_line`, `sim_median_line`, and `sim_median_ci` are read with other keys ignored.

## Dropping small sample bins

`plot_vpc_cens()` accepts the same `min_bin_count` argument as `plot_vpc_cont()`, but the count statistic it filters against differs by design. `plot_vpc_cont()` counts **quantifiable** (non-missing, non-BLQ) observations per bin — values that contribute to the continuous range quantiles. `plot_vpc_cens()` counts **total** observations per bin, including BLQ records, because the BLQ frequency is the diagnostic.

The practical consequence: for a bin with 12 observations and 8 BLQ values, `plot_vpc_cont(min_bin_count = 10)` drops the bin (only 4 quantifiable observations meet the threshold) while `plot_vpc_cens(min_bin_count = 10)` keeps it (12 total observations meet the threshold). Set `min_bin_count` per-function rather than reusing a single threshold across both diagnostics — the same numeric value will not in general produce the same bin-retention behavior.

A second contrast worth noting: because `obs_n` is determined by the protocol-specified sampling schedule (every subject in a stratum contributes one observation per nominal time bin), `obs_n` is effectively constant across bins within a stratum. As a result, `min_bin_count` on a cens VPC behaves more like a stratum-level filter than the bin-level filter it is on a cont VPC. The example below uses the full `simout` (across all SAD dose levels) so that Part 1-SAD has 30 observations per bin (5 dose cohorts × 6 subjects) while Part 2-FE has 6. Setting `min_bin_count = 10` retains every Part 1-SAD bin and drops the entire Part 2-FE panel.

```{r plot-vpc-cens-min-bin, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
cens_vpc <- plot_vpc_cens(
  data = simout,
  strat_var = PART,
  loq = 1,
  min_bin_count = 10
) +
  vpc_cens_scales_labs
cens_vpc
```


## Pairing pcVPC and cens VPC for the same strata

`plot_vpc_cont()` and `plot_vpc_cens()` are complementary halves of a single diagnostic: the pcVPC interrogates whether the model captures the dynamic range above LOQ, while the cens VPC interrogates whether it captures the censoring frequency. Composed together on a shared time axis they make one figure that summarizes both halves of the diagnostic for the same stratification.

Build the pcVPC and cens VPC on `sim100` using `strat_var = PART`. The bottom panel will carry the shared x-axis label, so the top panel sets `x = NULL`.

```{r plot-vpc-pc-cens-patchwork, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
patch_stats <- df_vpcstats(data = simout, strat_var = PART)

pcvpc_patch <- plot_vpc_cont(
  data = patch_stats,
  pcvpc = TRUE
) +
  vpc_scales_labs +
  labs(x = NULL, y = "Pred-corrected Conc. (ng/mL)")

cens_vpc_patch <- plot_vpc_cens(
  data = patch_stats
) +
  vpc_cens_scales_labs
```

Stack the two diagnostics vertically with `patchwork`, prediction-corrected continuous range on top and censored range below. Both panels are built from the same `sim100` data and the same `PART` stratification, so the time bins align across the figure.

```{r plot-vpc-pc-cens-stack, fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
pcvpc_patch / cens_vpc_patch + plot_layout(heights = c(2, 1))
```


# VPC Plot Legends with `plot_vpc_legend()`

This section will review the functionality for generating legends for VPC plots with `plot_vpc_legend()`

`plot_vpc_legend()` is a helper plotting function that creates a legend for VPC plots. These legends can then be merged with the VPC plot into a single plot object using the `patchwork` package.

## Defaults for Continuous Range VPCs in `plot_vpc_legend()`

To obtain a legend for a quantifiable range VPC plot using default aesthetics, run `plot_vpc_legend()`. The default type argument is `type="cont"` for continuous range VPC. 

Pass `lloq` to include an LLOQ entry that mirrors the reference line drawn by `plot_vpc_cont()`. 

```{r vpc-legend, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_legend <- plot_vpc_legend(lloq = 1)
vpc_legend
```

The legend can then be combined with the ggplot object returned from `plot_vpc_cont()` into a single plot object with the `patchwork` package. We pair the default-theme legend with `vpc` (the LOQ-censored, PART-stratified plot from the [BLQ handling](#blq-handling) section, which also uses the default theme).

```{r plot-vpc-pcvpc-w-legend, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_wleg <- vpc + vpc_legend + plot_layout(heights = c(2.5, 1))
vpc_wleg
```

## Updating the Continuous Range Legend Elements with `plot_vpc_shown()`

The legend can be updated to remove or add elements based on `shown` argument common to `plot_vpc_cont()` and `plot_vpc_legend()`. The defaults list can be viewed and updated globally to be passed to all plots in a VPC workflow using `plot_vpc_shown()`. Assigning the modified element set to a `shown_elements` object lets the same configuration be reused by both the plot and the legend.

```{r vpc-legend-medonly}
shown_elements <- plot_vpc_shown(obs_pi_line = FALSE, sim_pi_ci = FALSE)
plot_vpc_legend(shown = shown_elements)
```

## Updating the Continuous Range Legend Theme with `plot_vpc_legend()`

The legend can also be updated with the updated VPC plot theme elements. This is easiest to do by setting a new theme object, which we did previously when we created `vpc_new_theme` with `plot_vpc_theme()`. We can use this same theme list object to generate the VPC plot with `plot_vpc_cont()` and the VPC legend with `plot_vpc_legend()`, again passing `lloq = 1` to keep the legend in sync with the LLOQ ref line on `vpc2`.

```{r vpc-legend2, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_new_legend <- plot_vpc_legend(lloq = 1, theme = vpc_new_theme)
vpc_new_legend
```

The legend can then be combined with the ggplot object returned from `plot_vpc_cont()` into a single plot object with the `patchwork` package. We pair `vpc_new_legend` (which uses `vpc_new_theme`) with `vpc2` (the same VPC plot rendered with `vpc_new_theme` earlier).

```{r plot-vpc-pcvpc-w-legend2, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
vpc_wleg2 <- vpc2 + vpc_new_legend + plot_layout(heights = c(2.5, 1))
vpc_wleg2
```


## Updating the censored range legend elements and theme with `plot_vpc_shown()` and `plot_vpc_legend(type = "cens")`

By default `sim_median_line` is hidden (`plot_vpc_shown()` sets it to `FALSE`) and `sim_median_ci` inherits the same red fill as `obs_median_line` from `plot_vpc_theme()`. This is fine when only the ribbon is shown, but enabling the simulated median line on top of the default theme places a black dashed line over a red ribbon next to a red observed line. With these aesthetics, the red observed line could easily be mistaken for a simulation element given color homology with the simulated interval. 

A practical convention is to (a) enable `sim_median_line` so the simulated central tendency is visible alongside its CI, and (b) recolor the simulated layers together so the line and ribbon read as one element, with the observed layers in a contrasting color.

`plot_vpc_cens()` reads four `shown` keys (`obs_point`, `obs_median_line`, `sim_median_line`, `sim_median_ci`) and the four corresponding `theme` keys. Build matching `cens_shown` and `cens_theme` objects so the change is explicit and reusable:

```{r vpc-cens-shown-theme}
cens_shown <- plot_vpc_shown(sim_median_line = TRUE)

cens_theme <- plot_vpc_theme(
  obs_point       = pmx_point(color = "#000000"),
  obs_median_line = pmx_line(color = "#000000"),
  sim_median_line = pmx_line(color = "#3388cc", linetype = "solid"),
  sim_median_ci   = pmx_ribbon(fill = "#3388cc")
)
```

Re-render the cens VPC with both objects applied. The simulated median sits inside its 90% CI ribbon (both blue), while the observed proportion is drawn in black for visual separation.

```{r plot-vpc-cens-themed, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
cens_vpc_themed <- plot_vpc_cens(
  data = sim100,
  strat_var = PART,
  loq = 1,
  shown = cens_shown,
  theme = cens_theme
) +
  vpc_cens_scales_labs

cens_vpc_themed
```

`plot_vpc_legend()` accepts the same `shown` and `theme` objects, so passing them alongside `type = "cens"` keeps the legend in sync with whatever combination of visibility toggles and aesthetic overrides was used for the panel. The `Sim Prop BLQ` entry now appears (because `sim_median_line` is on) and both simulated entries adopt the unified blue.

```{r vpc-cens-legend-themed, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
cens_legend_themed <- plot_vpc_legend(
  type = "cens",
  lloq = 1,
  shown = cens_shown,
  theme = cens_theme
)

cens_vpc_themed + cens_legend_themed + plot_layout(heights = c(2.5, 1))
```


## Patchworked legend with `plot_vpc_legend(type = "cens")`

`plot_vpc_legend()` accepts `type = "cens"` argument to produce a legend with the default labels for a censored data VPC (`"Obs Prop BLQ"`, `"Sim Prop BLQ"`, `"Sim 90% CI Prop BLQ"`). Prediction-interval related entries are suppressed regardless of `shown`. 

Pass `lloq` to mirror the LOQ source the cens VPC was built against.

```{r vpc-cens-legend, fig.height=4, fig.width=8, message=TRUE, warning=FALSE}
cens_legend <- plot_vpc_legend(type = "cens", lloq = 1)
cens_legend
```

The cens VPC and its legend can be combined into a single figure with the `patchwork` package, using the same heights ratio established for the cont legend pattern.

```{r plot-vpc-cens-w-legend, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
cens_vpc + cens_legend + plot_layout(heights = c(2.5, 1))
```

## Panel observed and censored range VPCs together with legends using `patchwork`

We can add legends to our stacked continous dara range pcVPC on top and censored data range VPC on bottom. Both panels are built from the same `sim100` data and the same `PART` stratification, so the time bins align across the figure.

```{r plot-vpc-pc-cens-stack-w-legends, fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
pcvpc_patch / vpc_legend / cens_vpc_patch / cens_legend + 
  plot_layout(heights = c(2, 0.5, 2, 0.5))
```

# See also

  + [PK and PK/PD EDA workflow](eda-pk-pkpd-workflow.html) — exploratory analysis of continuous longitudinal concentration-time data, response-time, and response-concentration data.
